### 2023-04-12 paper/language.tex
syntax; pipeline diagram

\section{Language Overview}

\subsection{Syntax}

\subsubsection{Shared Syntax}

The following syntactic forms are shared between the Pika surface language and PikaCore.

\begin{align*}
  &i ::= \cdots \mid -2 \mid -1 \mid 0 \mid 1 \mid 2 \mid \cdots
  \\
  &b ::= \True{} \mid \False
  \\
  &e ::= x \mid i \mid b \mid e\; e \mid \lambda x.\; e
  \\
  &\tau ::= \Int \mid \Bool \mid \tau \ra \tau
\end{align*}

\subsection{Pika Syntax}
\begin{align*}
  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \liftExpr_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{A}).\; e
  \\
  &\tau ::= \cdots \mid A \mid \alpha \mid (\alpha \isA \layout{A}) \Ra \tau
\end{align*}

\subsection{PikaCore Syntax}
\begin{align*}
  &e ::= \cdots \mid \ssl{\overline{x}}{\overline{P}} \mid \withIn{x := e : \tau} e
  \\
  &\tau ::= \cdots \mid \SSL(\overline{\textnormal{ssl-param}})
  \\
  &\textnormal{ssl-param} ::= \langle x, \tau \rangle
\end{align*}

### 2023-04-14 c/listConvert.c
handwritten C code file

@@ -1,0 +1,130 @@
+#include <stdlib.h>
+#include <stdio.h>
+
+typedef struct Sll {
+  int x_0;
+  struct Sll* x_1;
+} Sll;
+
+typedef struct Dll {
+  int x_0;
+  struct Dll* x_1;
+  struct Dll* x_2;
+} Dll;
+
+void convertList1(Dll* arg, Sll** r) {
+  if (arg == 0) {
+      // pattern match {}
+
+      // ssl {}
+    *r = 0;
+  } else {
+      // pattern match {x :-> h, (x+1) :-> w, (x+2) :-> z}
+    int h = arg->x_0;
+    Dll* w = arg->x_1;
+    Dll* z = arg->x_2;
+
+      // allocations for result
+    *r = malloc(sizeof(Sll));
+
+      // with {nxt} := convertList1 {w x} in ...
+    Sll* nxt = 0;
+    convertList1(arg, &nxt);
+
+      // ssl {y :-> h, (y+1) :-> nxt}
+    (*r)->x_0 = h;
+    (*r)->x_1 = nxt;
+  }
+}
+
+void convertList2(Sll* arg, Dll** r, Dll** z) {
+  if (arg == 0) {
+      // pattern match {}
+
+      // ssl {}
+    *r = 0;
+  } else {
+      // pattern match {y :-> h, (y+1) :-> nxt}
+    int h = arg->x_0;
+    Sll* nxt = arg->x_1;
+
+      // allocations for result
+    *r = malloc(sizeof(Dll));
+
+      // with {w x} := convertList2 {nxt} in ...
+    Dll* w = 0;
+    convertList2(nxt, &w, r);
+
+      // ssl {x :-> h, (x+1) :-> w, (x+2) :-> z}
+    (*r)->x_0 = h;
+    (*r)->x_1 = w;
+    (*r)->x_2 = *z;
+  }
+}
+
+Sll* generateSll(int start, int end) {
+  Sll* n = malloc(sizeof(Sll));
+  Sll* original = n;
+
+  for (int i = start; i < end; ++i) {
+    n->x_0 = i;
+
+    if (i != end-1) {
+      n->x_1 = malloc(sizeof(Sll));
+      n = n->x_1;
+    }
+  }
+
+  n->x_1 = 0;
+  return original;
+}
+
+void printSll(Sll* n) {
+  printf("[");
+
+  Sll* curr = n;
+
+  while (curr != 0) {
[…]

### 2023-04-14 paper/language.tex
A bit more description; some renaming
points-to heaplet in grammar
locations in grammar
inst; \Pika and \PikaCore
C code for example; simplify denotation function
Work on PikaCore syntax and give some explanation

@@
-\section{Language Overview}
+\section{Syntax}
 
-\subsection{Syntax}
+\subsection{Shared Syntax}
 
-\subsubsection{Shared Syntax}
-
-The following syntactic forms are shared between the Pika surface language and PikaCore.
+The following syntactic forms are shared between the \Pika{} surface language and \PikaCore.
 
@@
   \\
-  &e ::= x \mid i \mid b \mid e\; e \mid \lambda x.\; e
+  &e ::= x \mid i \mid b \mid \lambda x.\; e
   \\
@@
 
-\subsection{Pika Syntax}
+\subsection{\Pika{} Syntax}
 \begin{align*}
-  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \liftExpr_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{A}).\; e
+  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \liftExpr_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{X}).\; e
   \\
-  &\tau ::= \cdots \mid A \mid \alpha \mid (\alpha \isA \layout{A}) \Ra \tau
+  &\tau ::= \cdots \mid X \mid \alpha \mid e\; e \mid (\alpha \isA \layout{X}) \Ra \tau
 \end{align*}
 
-\subsection{PikaCore Syntax}
+\subsection{\PikaCore{} Syntax}
 \begin{align*}
-  &e ::= \cdots \mid \ssl{\overline{x}}{\overline{P}} \mid \withIn{x := e : \tau} e
+  &e ::= \cdots \mid e\; \overline{a} \mid \ssl{\overline{P}} \mid \withIn{x := e} e
   \\
-  &\tau ::= \cdots \mid \SSL(\overline{\textnormal{ssl-param}})
+  &a ::= \{ \overline{x} \} \mid \{ i \} \mid \{ b \}
   \\
-  &\textnormal{ssl-param} ::= \langle x, \tau \rangle
+  &\tau ::= \cdots \mid \SSL(n)
+  \\
+  &P ::= \ell \mapsto e
+  \\
+  &\ell ::= x \mid (x + n)
+  \\
+  &n ::= 0 \mid 1 \mid 2 \mid \cdots
 \end{align*}
+
+The general form of a one-parameter \PikaCore{} function definition is
+\begin{align*}
+  &f : \SSL(m) \ra \SSL(n)\\
+  &f\; \{ \overline{\ell \mapsto y} \} \ra \{ \overline{r} \} := e\\
+  &f\; \{ \overline{\ell' \mapsto y} \} \ra \{ \overline{r} \} := e'\\
+  &\cdots
+\end{align*}
+
+\noindent
+where $\overline{r}$ are the names of the output parameters. These output parameters
+must be the same in each branch and it is also required that the number of output parameters matches the type signature: $|\overline{r}| = n$.
+
+Output parameters are given in \PikaCore{} as part of the \verb|with-in| construct:
+
+\begin{align*}
+  \withIn{\{ a, b \} := f\; \{ x \}} e
+\end{align*}
+
+This applies the function $f$ to the input parameter $x$ and gives $a$ and $b$ as output parameters. These two
+output parameters of the application are brought into scope for the \PikaCore{} expression $e$.

### 2023-04-15 paper/language.tex
Translation into PikaCore

@@
 \begin{align*}
-  &e ::= \cdots \mid e\; \overline{a} \mid \ssl{\overline{P}} \mid \withIn{x := e} e
+  &e ::= \cdots \mid e\; \overline{a} \mid \sslmath{\overline{x}}{\overline{h}} \mid \withIn{\{ \overline{x} \} := e} e
   \\
@@
   \\
-  &P ::= \ell \mapsto e
+  &h ::= \ell \mapsto e
   \\
@@
 
-The general form of a one-parameter \PikaCore{} function definition is
-\begin{align*}
-  &f : \SSL(m) \ra \SSL(n)\\
-  &f\; \{ \overline{\ell \mapsto y} \} \ra \{ \overline{r} \} := e\\
-  &f\; \{ \overline{\ell' \mapsto y} \} \ra \{ \overline{r} \} := e'\\
-  &\cdots
-\end{align*}
+The construct $\sslmath{\overline{x}}{\overline{h}}$ is an SSL assertion with heaplets $\overline{h}$ and
+bound variables $\overline{x}$. Note that it can have free variables as well.
 
-\noindent
-where $\overline{r}$ are the names of the output parameters. These output parameters
-must be the same in each branch and it is also required that the number of output parameters matches the type signature: $|\overline{r}| = n$.
-
-Output parameters are given in \PikaCore{} as part of the \verb|with-in| construct:
-
-\begin{align*}
-  \withIn{\{ a, b \} := f\; \{ x \}} e
-\end{align*}
-
-This applies the function $f$ to the input parameter $x$ and gives $a$ and $b$ as output parameters. These two
-output parameters of the application are brought into scope for the \PikaCore{} expression $e$.
+% The general form of a one-parameter \PikaCore{} function definition is
+% \begin{align*}
+%   &f : \SSL(m) \ra \SSL(n)\\
+%   &f\; \{ \overline{\ell \mapsto y} \} \outs \{ \overline{r} \} := e\\
+%   &f\; \{ \overline{\ell' \mapsto y} \} \outs \{ \overline{r} \} := e'\\
+%   &\cdots
+% \end{align*}
+%
+% \noindent
+% where $\overline{r}$ are the names of the output parameters. These output parameters
+% must be the same in each branch and it is also required that the number of output parameters matches the type signature: $|\overline{r}| = n$.
+%
+% Output parameters are given in \PikaCore{} as part of the \verb|with-in| construct:
+%
+% \begin{align*}
+%   \withIn{\{ a, b \} := f\; \{ x \}} e
+% \end{align*}
+%
+% This applies the function $f$ to the input parameter $x$ and gives $a$ and $b$ as output parameters. These two
+% output parameters of the application are brought into scope for the \PikaCore{} expression $e$.
+%

### 2023-04-16 paper/elaboration.tex
starting section on elaboration

\section{Elaboration}

The body of a layout polymorphic \Pika{} function will often make use of the 
layout parameters that it is given. For example, consider the function \texttt{singleton}.

\begin{align*}
  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
  &\texttt{singleton}\; i := \textrm{Cons}\; i\; \textrm{Nil}
\end{align*}

\noindent
We know just by looking at the type that we want to give back a \textrm{List} using
whatever layout is given as $a$. However, without elaboration, this is not enough. We must
explicitly lower the constructors to use the layout $a$ like this:

\begin{align*}
  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
  &\texttt{singleton}\; [a]\; i := \apply_a(\textrm{Cons}\; i\; (\apply_a(\textrm{Nil})))
\end{align*}

\noindent
Elaboration will turn the first definition of \texttt{singleton} into the second definition.

### 2023-04-16 paper/elaboration.tex
improve font

@@
 \begin{align*}
-  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
+  &\texttt{singleton} : (a \isA \layout{\textrm{List}}) \Ra \Int \ra a\\
   &\texttt{singleton}\; i := \textrm{Cons}\; i\; \textrm{Nil}
@@
 \begin{align*}
-  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
+  &\texttt{singleton} : (a \isA \layout{\textrm{List}}) \Ra \Int \ra a\\
   &\texttt{singleton}\; [a]\; i := \apply_a(\textrm{Cons}\; i\; (\apply_a(\textrm{Nil})))

### 2023-04-16 paper/elaboration.tex
brief description of PikaCore

--- 2023-04-16 paper/elaboration.tex
@@
 
-\begin{align*}
-  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
+\begin{flalign*}
+  &\texttt{singleton} : (a \isA \layout{\textrm{List}}) \Ra \Int \ra a\\
   &\texttt{singleton}\; i := \textrm{Cons}\; i\; \textrm{Nil}
-\end{align*}
+\end{flalign*}
 
@@
 
-\begin{align*}
-  &\texttt{singleton} : (a \isA \layout{List}) \Ra \Int \ra a\\
+\begin{flalign*}
+  &\texttt{singleton} : (a \isA \layout{\textrm{List}}) \Ra \Int \ra a\\
   &\texttt{singleton}\; [a]\; i := \apply_a(\textrm{Cons}\; i\; (\apply_a(\textrm{Nil})))
-\end{align*}
+\end{flalign*}
 
 \noindent
-Elaboration will turn the first definition of \texttt{singleton} into the second definition.
+Elaboration will turn the first definition of \texttt{singleton} into the second definition, using
+a modified type inference algorithm. \Pika{} definitions must be fully elaborated before being
+translated into \PikaCore.

### 2023-04-16 paper/language.tex
brief description of PikaCore
more description of PikaCore

@@
 \section{Syntax}
+
+\Pika{} code is translated into \PikaCore{} before being translated into the target language. \PikaCore{}
+is a layout monomorphic version of \Pika{}. While it retains most of the functional syntax of \Pika,
+all data structures are manipulated as collections of SSL heaplets. This includes both construction
+of data structure values and pattern matching. Also, the argument of every function application in
+\PikaCore{} is a literal or a variable. This is because it is entirely given in a kind of ANF, using
+the \verb|with-in| construct to introduce an ANF binding.
 

### 2023-04-16 paper/language.tex
remove 'lift' constructor
Some description of the layout value introduction syntax and how parameters and variable freshness works for it

@@
 the \verb|with-in| construct to introduce an ANF binding.
+
+The syntactic form that introduces an SSL assertion in \PikaCore{} can be seen as a kind of binding
+form, the values of which can be ``applied''. For example, $\sslmath{x, y}{x \mapsto y, (x+1) \mapsto z}$
+can be seen as an SSL assertion that is parametrized over two locations: $x$ and $y$. Also, the name
+$z$ occurs free. When this layout value is applied to two locations, $z$ must remain a fresh variable in
+the use-site.
 
@@
 \begin{align*}
-  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \liftExpr_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{X}).\; e
+  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{X}).\; e
   \\

### 2023-04-16 paper/language.tex
explanation of with-in construct and how it relates to layout values and function calls in PikaCore

@@
 the use-site.
+
+There is only one situation in \PikaCore{} in which layouts can be applied. This is when a function
+is called. A function, generally, takes at least one layout value as input and produces a layout value as
+output. The \verb|with-in| construct allows both of the input and output layouts to be applied with
+appropriate variables, representing the input and output variables that the function will use when
+it's called.
+
+For example, given a function
+
+\begin{align*}
+  &f\; \{ x \mapsto a \} := \sslmath{y, z}{(y+1) \mapsto a, z \mapsto w}
+\end{align*}
+
+\noindent
+we can apply it like this
+
+\begin{align*}
+  \withIn{\{p, q\} := f\; \{s\}} e
+\end{align*}
+
+\noindent
+If $s$ is nonzero (meaning $s$ points to something and this branch of $f$ is actually used), this will result in an SSL assertion
+for the input $\{s \mapsto a\}$ and an SSL assertion for the output $\{p \mapsto a, q \mapsto w\}$ where $w$ is fresh. These
+SSL assertions are combined with the SSL assertion given by the expression $e$.
 

### 2023-04-16 paper/language.tex
Preserve type information in PikaCore for use in later stages

@@
 The syntactic form that introduces an SSL assertion in \PikaCore{} can be seen as a kind of binding
-form, the values of which can be ``applied''. For example, $\sslmath{x, y}{x \mapsto y, (x+1) \mapsto z}$
+form, the values of which can be ``applied''. For example, $\sslmath{x, y}{T}{x \mapsto y, (x+1) \mapsto z}$
 can be seen as an SSL assertion that is parametrized over two locations: $x$ and $y$. Also, the name
@@
 \begin{align*}
-  &f\; \{ x \mapsto a \} := \sslmath{y, z}{(y+1) \mapsto a, z \mapsto w}
+  &f\; \{ x \mapsto a \} := \sslmath{y, z}{T}{(y+1) \mapsto a, z \mapsto w}
 \end{align*}
@@
   \\
-  &\tau ::= \Int \mid \Bool \mid \tau \ra \tau
+  &\tau ::= \Int \mid \Bool \mid \alpha \mid \tau \ra \tau
 \end{align*}
@@
 \begin{align*}
-  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \Lambda (\alpha \isA \layout{X}).\; e
+  &e ::= \cdots \mid \apply_{\alpha}(e) \mid \Lambda (\alpha \isA \Layout{X}).\; e
   \\
-  &\tau ::= \cdots \mid X \mid \alpha \mid e\; e \mid (\alpha \isA \layout{X}) \Ra \tau
+  &\tau ::= \cdots \mid X \mid e\; e \mid (\alpha \isA \Layout{X}) \Ra \tau
 \end{align*}
@@
 \begin{align*}
-  &e ::= \cdots \mid e\; \overline{a} \mid \sslmath{\overline{x}}{\overline{h}} \mid \withIn{\{ \overline{x} \} := e} e
+  &e ::= \cdots \mid e\; \overline{a} \mid \sslmath{\overline{x}}{T}{\overline{h}} \mid \withIn{\{ \overline{x} \} := e} e
   \\
@@
   \\
-  &\tau ::= \cdots \mid \SSL(n)
+  &\tau ::= \cdots \mid A
   \\
   &h ::= \ell \mapsto e
+  \\
+  &T ::= \overline{\ell : \tau}
+  % \\
+  % &d ::= \recspec{\overline{n : \tau}}
   \\
@@
 
-The construct $\sslmath{\overline{x}}{\overline{h}}$ is an SSL assertion with heaplets $\overline{h}$ and
+The construct $\sslmath{\overline{x}}{T}{\overline{h}}$ is an SSL assertion with heaplets $\overline{h}$ and
 bound variables $\overline{x}$. Note that it can have free variables as well.
+
+$\mu\alpha.\; \SSL(\overline{x : \tau})$ is an equi-recursive type, with $\mu$ being the recursive binder,
+introducing the type variable $\alpha$. When
+the variable bound by $\mu$ does not appear in any of the $\tau$, we leave out the binder and write $\SSL(\overline{x : \tau})$
+as a shorthand.
+
+% $d$ is a record specification, given by index. Consider the following SSL assertion.
+% \[
+%   \sslmath{x,y}{x \mapsto 1, (x+2) \mapsto \True, y \mapsto 22, (y+1) \mapsto 0}
+% \]
+% This could have the type $\SSL(x : \recspec{0 : \Int, 2 : \Bool}, y : \recspec{0 : \Int, 1 : \Int})$.
+% It could have other types as well, however. For example, $y$ could also fit the
+% layout of a singly-linked list. We might not know at a type-level how long a list is. So, the
+% assertion also has the type
+% $\SSL(x : \recspec{0 : \Int, 2 : \Bool}, y : \recspec{0 : \Layout})$. We use $\Layout$ to indicate
+% that we only know
+
+% $d$ is a record specification, given by index. Consider the following SSL assertion.
+% \[
+%   \sslmath{x,y}{x \mapsto 1, (x+2) \mapsto \True, y \mapsto 22, (y+1) \mapsto 0}
+% \]
+% This could have the type $\SSL(x : \recspec{0 : \Int, 2 : \Bool}, y : \recspec{0 : \Int, 1 : \Int})$.
+% It could have other types as well, however. For example, $y$ could also fit the
+% layout of a singly-linked list. If we have such a singly-linked list layout
+% named \verb|Sll|, then the assertion would also have the type
+% $\SSL(x : \recspec{0 : \Int, 2 : \Bool}, y : \recspec{0 : \texttt{Sll}})$. Regarding $y$, this
+% type says that, at offset $0$, $y$ satisfies the \texttt{Sll} layout.
 

### 2023-05-15 src/PikaC/Ppr.hs
PikaCore implementation

@@ -1,30 +1,52 @@
 {-# LANGUAGE FlexibleInstances #-}
 
 module PikaC.Ppr
+  (module Text.PrettyPrint.HughesPJ
+  ,Ppr
+  ,ppr
+  ,ppr'
+  ,pprP
+  ,IsNested
+  ,isNested
+  )
   where
 
-data Brackets = NoBrackets | Brackets
+import Text.PrettyPrint.HughesPJ hiding ((<>))
 
 class Ppr a where
-  pprS :: a -> Brackets -> String
+  ppr :: a -> Doc
 
-ppr :: Ppr a => a -> String
-ppr = flip pprS NoBrackets
+ppr' :: Ppr a => a -> String
+ppr' = render . ppr
 
-pprGrouped :: Ppr a => a -> String
-pprGrouped x = pprS x Brackets
+pprP :: (IsNested a, Ppr a) => a -> Doc
+pprP x = maybeParens (isNested x) (ppr x)
 
-instance Ppr String where pprS = const
-instance Ppr Int where pprS = (show .) . const
+instance Ppr String where ppr = text
+instance Ppr Int where ppr = text . show
+instance Ppr Bool where ppr = text . show
 
-grouped :: String -> Brackets -> String
-grouped = groupedWith ("(", ")")
+class IsNested a where
+  isNested :: a -> Bool
 
-groupedWith :: (String, String) -> String -> Brackets -> String
-groupedWith (start, end) x NoBrackets = x
-groupedWith (start, end) x Brackets = start <> x <> end
-
-standalone :: String -> Brackets -> String
-standalone = const
-
-
+--   pprS :: a -> Brackets -> String
+-- data Brackets = NoBrackets | Brackets
+--
+-- ppr :: Ppr a => a -> String
+-- ppr = flip pprS NoBrackets
+--
+-- pprGrouped :: Ppr a => a -> String
+-- pprGrouped x = pprS x Brackets
+--
+--
+-- grouped :: String -> Brackets -> String
+-- grouped = groupedWith ("(", ")")
+--
+-- groupedWith :: (String, String) -> String -> Brackets -> String
+-- groupedWith (start, end) x NoBrackets = x
+-- groupedWith (start, end) x Brackets = start <> x <> end
+--
+-- standalone :: String -> Brackets -> String
+-- standalone = const
+--
+--

### 2023-05-15 src/PikaC/Syntax/Heaplet.hs

@@ -1,0 +1,65 @@
+{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE FlexibleInstances #-}
+
+module PikaC.Syntax.Heaplet
+  where
+
+import Data.Semigroup
+import Data.List
+import Data.Function
+import PikaC.Ppr
+
+-- data Heaplet a
+--   deriving (Show, Functor)
+
+data PointsTo a e = Loc a :-> e
+  deriving (Show, Functor)
+
+data Loc a = a :+ Int
+  deriving (Show, Functor)
+
+instance Ppr a => Ppr (Loc a) where
+  ppr (x :+ 0) = ppr x
+  ppr (x :+ i) = parens $ hsep [ppr x, text "+", ppr i]
+
+instance (Ppr a, Ppr e) => Ppr (PointsTo a e) where
+  ppr (lhs :-> rhs) = hsep [ppr lhs, text ":->", ppr rhs]
+
+instance (Ppr a, Ppr e) => Ppr [PointsTo a e] where
+  ppr xs = braces $ sep $ punctuate (text "**") (map ppr xs)
+
+pointsToLhs :: PointsTo a e -> Loc a
+pointsToLhs (lhs :-> _) = lhs
+
+pointsToRhs :: PointsTo a e -> e
+pointsToRhs (_ :-> rhs) = rhs
+
+locBase :: Loc a -> a
+locBase (x :+ _) = x
+
+locIx :: Loc a -> Int
+locIx (_ :+ i) = i
+
+pointsToNames :: Ord a => [PointsTo a e] -> [a]
+pointsToNames = nub . map (locBase . pointsToLhs)
+
+data Allocation a = Alloc a Int
+  deriving (Show)
+
+findAllocations :: forall a b. Eq a => [a] -> [PointsTo a b] -> [Allocation a]
+findAllocations names xs = map toAlloc locMaximums
+  where
+      -- Grouped by base name. Only include locations that have base names
+      -- included in the 'names' argument
+    locGroups :: [[Loc a]]
+    locGroups = groupBy ((==) `on` locBase) . filter ((`elem` names) . locBase) $ map pointsToLhs xs
+
+    locMaximums :: [Loc a]
+    locMaximums = map (foldr1 go) locGroups
+
+    go :: Loc a -> Loc a -> Loc a
+    go (name :+ i) (_ :+ j) = name :+ max i j
+
+    toAlloc :: Loc a -> Allocation a
+    toAlloc (x :+ i) = Alloc x i
+

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -1,0 +1,81 @@
+module PikaC.Syntax.PikaCore.Expr
+  where
+
+import PikaC.Syntax.Heaplet
+import PikaC.Ppr
+import Data.List
+
+data Base a
+  = V a
+  | LayoutV (LayoutArg a)    -- {x ...}
+  | IntLit Int
+  | BoolLit Bool
+
+  | Add (Base a) (Base a)
+  | Sub (Base a) (Base a)
+  | Equal (Base a) (Base a)
+  | Not (Base a)
+  | And (Base a) (Base a)
+  deriving (Show, Functor)
+
+newtype LayoutArg a = LayoutArg [Base a]
+  deriving (Show, Functor)
+
+data SimpleExpr a
+  = BaseExpr (Base a)
+  | WithIn                -- with
+      (LayoutArg a)       --   {x ...}
+      (Expr a)            --     := e
+      (SimpleExpr a)      -- in e
+
+  | SslAssertion            -- layout
+      (LayoutArg a)         --     {x ...}
+      (ExprAssertion a)     --   { (x+1) :-> e ** ... }
+  deriving (Show)
+
+data Expr a
+  = SimpleExpr (SimpleExpr a)
+  | App (Expr a) (LayoutArg a)
+  deriving (Show)
+
+type PointsToExpr a = PointsTo a (Expr a)
+type ExprAssertion a = [PointsToExpr a]
+
+instance Ppr a => Ppr (LayoutArg a) where
+  ppr (LayoutArg xs) = text "{" <+> hsep (punctuate (text ",") (map ppr xs)) <+> text "}"
+
+instance Ppr a => Ppr (Base a) where
+  ppr (V x) = ppr x
+  ppr (LayoutV x) = ppr x
+  ppr (IntLit i) = ppr i
+  ppr (BoolLit b) = ppr b
+  ppr (Add x y) = sep [pprP x, text "+", pprP y]
+  ppr (Sub x y) = sep [pprP x, text "-", pprP y]
+  ppr (Equal x y) = sep [pprP x, text "==", pprP y]
+  ppr (Not x) = sep [text "!", pprP x]
+  ppr (And x y) = sep [pprP x, text "&&", pprP y]
+
+instance Ppr a => Ppr (SimpleExpr a) where
+  ppr (BaseExpr e) = ppr e
+  ppr (WithIn vars bnd body) =
+    sep [text "with", ppr vars, text ":=", ppr bnd, text "in", ppr body]
+
[…]

### 2023-05-15 src/PikaC/Syntax/PikaCore/FnDef.hs

@@ -1,0 +1,67 @@
+{-# LANGUAGE ScopedTypeVariables #-}
+
+module PikaC.Syntax.PikaCore.FnDef
+  where
+
+import PikaC.Syntax.PikaCore.Expr
+import PikaC.Syntax.Heaplet
+import PikaC.Ppr
+
+data FnDef a =
+  FnDef
+  { fnDefName :: String
+  , fnDefBranches :: [FnDefBranch a]
+  }
+
+data FnDefBranch a =
+  FnDefBranch
+  { fnDefOutputParams :: LayoutArg a
+  , fnDefBranchInputAssertions :: [ExprAssertion a]
+  , fnDefBranchBody :: Expr a
+  }
+
+fnDefInputNames :: Ord a => FnDef a -> [a]
+fnDefInputNames = concatMap fnDefBranchInputNames . fnDefBranches
+
+fnDefBranchInputNames :: Ord a => FnDefBranch a -> [a]
+fnDefBranchInputNames =
+  concatMap pointsToNames . fnDefBranchInputAssertions
+
+computeBranchCondition :: Ord a => FnDef a -> FnDefBranch a -> Base a
+computeBranchCondition def branch = go allInputNames
+  where
+    allInputNames = fnDefInputNames def
+    branchNames = fnDefBranchInputNames branch
+
+    checkName name =
+      if name `elem` branchNames
+        then Equal (V name) (IntLit 0)
+        else Not (Equal (V name) (IntLit 0))
+
+    go [] = BoolLit True
+    go [name] = checkName name
+    go (name:rest) =
+      and' (checkName name) (go rest)
+
+instance forall a. Ppr a => Ppr (FnDef a) where
+  ppr def = go (fnDefBranches def)
+    where
+      go :: [FnDefBranch a] -> Doc
+      go [] = mempty
+      go (branch : rest) = (ppr (fnDefName def) <+> ppr branch) $$ go rest
+
+instance Ppr a => Ppr (FnDefBranch a) where
+  ppr branch =
+    sep
+      [ hsep $ punctuate (text " ") (map ppr (fnDefBranchInputAssertions branch))
+      , text "==>"
+      , ppr (fnDefOutputParams branch)
+      , text ":="
+      , ppr (fnDefBranchBody branch)
+      , text ";"
+      ]
+
+and' :: Base a -> Base a -> Base a
+and' x (BoolLit True) = x
+and' x y = And x y
+

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
C backend

@@ -1,0 +1,5 @@
+module PikaC.Backend.C.CodeGen
+  where
+
+
+

### 2023-05-15 src/PikaC/Backend/C/Syntax.hs

@@ -1,0 +1,80 @@
+module PikaC.Backend.C.Syntax
+  where
+
+import PikaC.Syntax.Heaplet
+import PikaC.Ppr
+
+data CExpr a
+  = V a
+  | BoolLit Bool
+  | IntLit Int
+  | Add (CExpr a) (CExpr a)
+  | Sub (CExpr a) (CExpr a)
+  | Equal (CExpr a) (CExpr a)
+  | And (CExpr a) (CExpr a)
+  | Not (CExpr a)
+  | Deref (Loc a)
+
+data Command a
+  = Assign (Loc a) (CExpr a)
+  | IfThenElse (CExpr a) [Command a] [Command a]
+  | Call String [CExpr a]
+  | IntoMalloc a Int
+  | Free a
+
+data CFunction a =
+  CFunction
+    { cfunctionName :: String
+    , cfunctionParams :: [a]
+    , cfunctionBody :: [Command a]
+    }
+
+instance Ppr a => Ppr (Command a) where
+  ppr (Assign loc e) =
+    hsep [text "*", ppr loc, text "=", ppr e] <> text ";"
+
+  ppr (IfThenElse c t f) =
+    sep [hsep [text "if", text "(", ppr c, text ")", text "{"]
+        ,nest 1 (hsep (map ppr t))
+        ,hsep [text "}", text "else", text "{"]
+        ,nest 1 (hsep (map ppr f))
+        ,text "}"
+        ]
+
+  ppr (Call f args) =
+    hsep [text f, text "(", hsep (punctuate (text ",") (map ppr args)), text ")", text ";"]
+
+  ppr (IntoMalloc target size) =
+    hsep [ppr target, text "=", text "malloc(", ppr size, text "*sizeof(loc));"]
+
+  ppr (Free x) =
+    hsep [text "free(", ppr x, text ");"]
+
+instance Ppr a => Ppr (CFunction a) where
+  ppr fn =
+    vcat [text "void", text (cfunctionName fn), text "(", hsep (punctuate (text ",") (map ((text "loc" <+>) . ppr) (cfunctionParams fn))), text ")", text "{"
+         ,vcat (map ppr (cfunctionBody fn))
+         ,text "}"
+         ]
+
+instance Ppr a => Ppr (CExpr a) where
+  ppr (V x) = ppr x
+  ppr (IntLit i) = ppr i
+  ppr (BoolLit b) = ppr b
[…]

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -58,7 +58,7 @@ instance Ppr a => Ppr (Base a) where
 instance Ppr a => Ppr (SimpleExpr a) where
   ppr (BaseExpr e) = ppr e
   ppr (WithIn vars bnd body) =
-    sep [text "with", ppr vars, text ":=", ppr bnd, text "in", ppr body]
+    sep [text "with", hsep [ppr vars, text ":=", ppr bnd], text "in", ppr body]
 
   ppr (SslAssertion vars heaplets) =
     sep [text "layout", ppr vars, ppr heaplets]

### 2023-05-15 src/PikaC/Syntax/PikaCore/FnDef.hs

@@ -57,7 +57,7 @@ instance Ppr a => Ppr (FnDefBranch a) where
       , text "==>"
       , ppr (fnDefOutputParams branch)
       , text ":="
-      , ppr (fnDefBranchBody branch)
+      , nest 1 $ ppr (fnDefBranchBody branch)
       , text ";"
       ]
 

### 2023-05-15 src/PikaC/Syntax/PikaCore/Layout.hs

@@ -1,0 +1,16 @@
+module PikaC.Syntax.PikaCore.Layout
+  where
+
+import PikaC.Syntax.Heaplet
+
+-- data Layout a =
+--   Layout
+--     { layoutName :: String
+--     , layoutBranches :: [LayoutBranch a]
+--     }
+--
+-- data LayoutBranch a =
+--   LayoutBranch
+--     { layoutPattern :: Pattern a
+--     }
+

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
renaming

@@ -1,5 +1,32 @@
 module PikaC.Backend.C.CodeGen
   where
 
+import qualified PikaC.Syntax.PikaCore.Expr as PikaCore
+import qualified PikaC.Backend.C.Syntax as C
+import PikaC.Backend.C.Syntax (CExpr, Command, CFunction)
+import PikaC.Syntax.Heaplet
 
+import GHC.Stack
+
+type Outputs = LayoutArg
+
+codeGenBase :: HasCallStack =>
+  PikaCore.Base a ->
+  CExpr a
+codeGenBase (PikaCore.V x) = C.V x
+codeGenBase (PikaCore.LayoutV xs) = error "codeGenBase: LayoutV"
+codeGenBase (PikaCore.IntLit i) = C.IntLit i
+codeGenBase (PikaCore.BoolLit b) = C.BoolLit b
+codeGenBase (PikaCore.Add x y) = C.Add (codeGenBase x) (codeGenBase y)
+codeGenBase (PikaCore.Sub x y) = C.Sub (codeGenBase x) (codeGenBase y)
+codeGenBase (PikaCore.Equal x y) = C.Equal (codeGenBase x) (codeGenBase y)
+codeGenBase (PikaCore.Not x) = C.Not (codeGenBase x)
+codeGenBase (PikaCore.And x y) = C.And (codeGenBase x) (codeGenBase y)
+
+codeGenSimple :: HasCallStack => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
+codeGenSimple outputs (PikaCore.WithIn vars bnd body) =
+  codeGenExpr vars bnd ++ codeGenSimple outputs body
+
+codeGenExpr :: HasCallStack => Outputs a -> PikaCore.Expr a -> [Command a]
+codeGenExpr = undefined
 

### 2023-05-15 src/PikaC/Syntax/Heaplet.hs

@@ -9,29 +9,63 @@ import Data.List
 import Data.Function
 import PikaC.Ppr
 
+import Data.Maybe
+import Control.Monad.Identity
+
 -- data Heaplet a
 --   deriving (Show, Functor)
 
-data PointsTo a e = Loc a :-> e
+newtype LayoutArg a = LayoutArg [a]
   deriving (Show, Functor)
+
+class LayoutRename f where
+  renameLayoutArg ::
+    Ord a => LayoutArg a -> LayoutArg a -> f a -> f a
+
+instance LayoutRename LayoutArg where
+  renameLayoutArg old new (LayoutArg xs) =
+      LayoutArg $ map (renameLayoutArg' old new) xs
+    where
+      go = renameLayoutArg' old new
+
+instance LayoutRename Identity where
+  renameLayoutArg old new (Identity x) =
+    Identity $ renameLayoutArg' old new x
+
+renameLayoutArg' :: Ord a =>LayoutArg a -> LayoutArg a -> a -> a
+renameLayoutArg' (LayoutArg old) (LayoutArg new) x =
+      fromMaybe x $ lookup x assocs
+    where
+      assocs = zip old new
+
+data PointsTo f a = Loc a :-> f a
+  deriving (Show)
 
 data Loc a = a :+ Int
   deriving (Show, Functor)
+
+instance LayoutRename f => LayoutRename (PointsTo f) where
+  renameLayoutArg old new (lhs :-> rhs) =
+    renameLayoutArg old new lhs :-> renameLayoutArg old new rhs
+
+instance LayoutRename Loc where
+  renameLayoutArg old new (x :+ i) =
+    renameLayoutArg' old new x :+ i
 
 instance Ppr a => Ppr (Loc a) where
   ppr (x :+ 0) = ppr x
   ppr (x :+ i) = parens $ hsep [ppr x, text "+", ppr i]
 
-instance (Ppr a, Ppr e) => Ppr (PointsTo a e) where
+instance (Ppr a, Ppr (f a)) => Ppr (PointsTo f a) where
   ppr (lhs :-> rhs) = hsep [ppr lhs, text ":->", ppr rhs]
 
-instance (Ppr a, Ppr e) => Ppr [PointsTo a e] where
+instance (Ppr a, Ppr (f a)) => Ppr [PointsTo f a] where
   ppr xs = braces $ sep $ punctuate (text "**") (map ppr xs)
 
[…]

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -6,9 +6,9 @@ import PikaC.Ppr
 import Data.List
 
 data Base a
-  = V a
-  | LayoutV (LayoutArg a)    -- {x ...}
-  | IntLit Int
+  -- = V a
+  = LayoutV (LayoutArg a)    -- {x ...}
+  | IntLit Int -- TODO: Add output locations?
   | BoolLit Bool
 
   | Add (Base a) (Base a)
@@ -16,9 +16,6 @@ data Base a
   | Equal (Base a) (Base a)
   | Not (Base a)
   | And (Base a) (Base a)
-  deriving (Show, Functor)
-
-newtype LayoutArg a = LayoutArg [Base a]
   deriving (Show, Functor)
 
 data SimpleExpr a
@@ -38,14 +35,55 @@ data Expr a
   | App (Expr a) (LayoutArg a)
   deriving (Show)
 
-type PointsToExpr a = PointsTo a (Expr a)
+type PointsToExpr a = PointsTo Expr a
 type ExprAssertion a = [PointsToExpr a]
+
+instance LayoutRename Base where
+  -- renameLayoutArg old new (V x) = V x
+  renameLayoutArg old new (LayoutV xs) =
+    LayoutV $ renameLayoutArg old new xs
+  renameLayoutArg old new (IntLit i) = IntLit i
+  renameLayoutArg old new (BoolLit b) = BoolLit b
+  renameLayoutArg old new (Add x y) =
+    Add (renameLayoutArg old new x)
+        (renameLayoutArg old new y)
+  renameLayoutArg old new (Sub x y) =
+    Sub (renameLayoutArg old new x)
+        (renameLayoutArg old new y)
+  renameLayoutArg old new (Equal x y) =
+    Equal (renameLayoutArg old new x)
+          (renameLayoutArg old new y)
+  renameLayoutArg old new (Not x) =
+    Not (renameLayoutArg old new x)
+  renameLayoutArg old new (And x y) =
+    And (renameLayoutArg old new x)
+        (renameLayoutArg old new y)
+
+instance LayoutRename SimpleExpr where
+  renameLayoutArg old new (BaseExpr e) =
+    BaseExpr $ renameLayoutArg old new e
+
+  renameLayoutArg old0@(LayoutArg old) new0@(LayoutArg new) (WithIn vars0@(LayoutArg vars) bnd body) =
+    WithIn
+      vars0
+      (renameLayoutArg old0 new0 bnd)
[…]

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
work on topological sorting of points-to dependencies

@@ -2,9 +2,14 @@ module PikaC.Backend.C.CodeGen
   where
 
 import qualified PikaC.Syntax.PikaCore.Expr as PikaCore
+import PikaC.Syntax.PikaCore.Expr (PointsToExpr)
 import qualified PikaC.Backend.C.Syntax as C
 import PikaC.Backend.C.Syntax (CExpr, Command, CFunction)
 import PikaC.Syntax.Heaplet
+
+import PikaC.Preorder.Preorder
+
+import Data.Foldable
 
 import GHC.Stack
 
@@ -13,7 +18,7 @@ type Outputs = LayoutArg
 codeGenBase :: HasCallStack =>
   PikaCore.Base a ->
   CExpr a
-codeGenBase (PikaCore.V x) = C.V x
+-- codeGenBase (PikaCore.V x) = C.V x
 codeGenBase (PikaCore.LayoutV xs) = error "codeGenBase: LayoutV"
 codeGenBase (PikaCore.IntLit i) = C.IntLit i
 codeGenBase (PikaCore.BoolLit b) = C.BoolLit b
@@ -23,10 +28,34 @@ codeGenBase (PikaCore.Equal x y) = C.Equal (codeGenBase x) (codeGenBase y)
 codeGenBase (PikaCore.Not x) = C.Not (codeGenBase x)
 codeGenBase (PikaCore.And x y) = C.And (codeGenBase x) (codeGenBase y)
 
-codeGenSimple :: HasCallStack => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
+codeGenSimple :: (HasCallStack, Ord a) => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
 codeGenSimple outputs (PikaCore.WithIn vars bnd body) =
   codeGenExpr vars bnd ++ codeGenSimple outputs body
 
-codeGenExpr :: HasCallStack => Outputs a -> PikaCore.Expr a -> [Command a]
-codeGenExpr = undefined
+codeGenSimple outputs (PikaCore.SslAssertion params asn0) =
+  let asn = fmap (renameLayoutArg params outputs) asn0
+      sorted = topologicalSortPointsTo asn
+  in
+  map codeGenPointsTo sorted
+
+topologicalSortPointsTo :: [PointsToExpr a] -> [PointsToExpr a]
+topologicalSortPointsTo = map to . topologicalSort . map from
+  where
+    to (x, y) = x :-> y
[…]

### 2023-05-15 src/PikaC/Preorder/Preorder.hs

@@ -1,0 +1,371 @@
+{-# LANGUAGE DeriveFoldable #-}
+{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE DeriveGeneric #-}
+{-# LANGUAGE OverloadedStrings #-}
+
+module PikaC.Preorder.Preorder
+  (Le (..)
+  ,Preorder
+  ,preorder
+  ,CompareResult (..)
+  ,compareP
+  ,opposite
+  ,toOrdering
+  ,preorderElements
+  ,pMax
+
+  ,topologicalSort
+
+  ,Contains (..)
+  ,getContainsLabel
+  ,getContainsChildren
+  ,immediatelyLe
+  ,immediatelyGe
+  ,leastElems
+  ,greatestElems
+  ,validContains
+  ,validContainsList
+
+  ,preorderToContains
+  ,containsToPreorder
+  ,getSymmetricElements
+  )
+  where
+
+import Data.List
+import Data.Foldable
+
+import Hedgehog
+import qualified Hedgehog.Gen as Gen
+import qualified Hedgehog.Range as Range
+import Hedgehog.Gen (element)
+
+import qualified Data.Set as Set
+
+import Control.Applicative
+
+import PikaC.Utils
+
+import GHC.Generics
+
+import Data.Ord
+import Data.Maybe
+
+topologicalSort :: Ord a => [(a, a)] -> [(a, a)]
+topologicalSort orig =
+  filter (`elem` orig) .
+  map go .
+  getPreorder .
+  containsToPreorder .
+  preorderToContains .
+  preorder $
+  map (uncurry (:<=)) orig
+  where
+    go (x :<= y) = (x, y)
+
+data Le a = a :<= a
+  deriving (Show, Eq, Ord, Foldable, Generic)
+
+data Preorder a = Preorder { rawPreorderElements :: [a], getPreorder :: [Le a] }
+  deriving (Show, Eq)
+
+opposite :: Preorder a -> Preorder a
+opposite (Preorder elems p) = Preorder elems $ map flipR p
+  where
+    flipR (x :<= y) = y :<= x
+
+-- | Input: Generators for the preorder (the preorder is the reflexive transitive closure of these)
+--   Output: The transitive closure of the generators
+preorder :: Ord a => [Le a] -> Preorder a
+preorder xs = Preorder elems . fastNub $ removeRefls $ preorder' xs
+  where
[…]

### 2023-05-15 src/PikaC/Syntax/Heaplet.hs

@@ -16,7 +16,7 @@ import Control.Monad.Identity
 --   deriving (Show, Functor)
 
 newtype LayoutArg a = LayoutArg [a]
-  deriving (Show, Functor)
+  deriving (Show, Functor, Semigroup, Monoid, Foldable)
 
 class LayoutRename f where
   renameLayoutArg ::

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -32,10 +32,11 @@ data SimpleExpr a
 
 data Expr a
   = SimpleExpr (SimpleExpr a)
-  | App (Expr a) (LayoutArg a)
+  -- | App (Expr a) (LayoutArg a)
+  | App String [LayoutArg a] -- | Fully saturated function application
   deriving (Show)
 
-type PointsToExpr a = PointsTo Expr a
+type PointsToExpr a = PointsTo Base a
 type ExprAssertion a = [PointsToExpr a]
 
 instance LayoutRename Base where
@@ -77,7 +78,7 @@ instance LayoutRename Expr where
     SimpleExpr $ renameLayoutArg old new e
 
   renameLayoutArg old new (App f x) =
-    App (renameLayoutArg old new f) (renameLayoutArg old new x)
+    App f (map (renameLayoutArg old new) x)
 
 instance Ppr a => Ppr (LayoutArg a) where
   ppr (LayoutArg xs) = text "{" <+> hsep (punctuate (text ",") (map ppr xs)) <+> text "}"
@@ -103,7 +104,16 @@ instance Ppr a => Ppr (SimpleExpr a) where
 
 instance Ppr a => Ppr (Expr a) where
   ppr (SimpleExpr e) = ppr e
-  ppr (App f x) = ppr f <+> ppr x
+  ppr (App f x) = ppr f <+> hsep (map ppr x)
+
+instance IsNested (Expr a) where
+  isNested (SimpleExpr e) = isNested e
+  isNested (App {}) = True
+
+instance IsNested (SimpleExpr a) where
+  isNested (BaseExpr e) = isNested e
+  isNested (WithIn {}) = True
+  isNested (SslAssertion {}) = True
 
 instance IsNested (Base a) where
   -- isNested (V _) = False

### 2023-05-15 src/PikaC/Utils.hs

@@ -1,0 +1,44 @@
+module PikaC.Utils
+  where
+
+import qualified Data.Set as Set
+
+import Data.Set (Set)
+
+import Data.Bifunctor
+import Data.List
+
+fastNub :: Ord a => [a] -> [a]
+fastNub = Set.toList . Set.fromList
+
+removeOne :: [a] -> [(a, [a])]
+removeOne = removeOneSuchThat (const True)
+
+removeOneSuchThat :: (a -> Bool) -> [a] -> [(a, [a])]
+removeOneSuchThat p [] = []
+removeOneSuchThat p (x:xs)
+  | p x = (x, xs) : map (second (x:)) (removeOneSuchThat p xs)
+  | otherwise = map (second (x:)) (removeOneSuchThat p xs)
+
+remove :: (a -> Maybe b) -> [a] -> Maybe (b, [a])
+remove f [] = Nothing
+remove f (x:xs) =
+  case f x of
+    Just y -> Just (y, xs)
+    Nothing -> fmap (second (x:)) (remove f xs)
+
+lookupRemove :: Eq a => a -> [(a, b)] -> Maybe (b, [(a, b)])
+lookupRemove x = remove $ \(lhs, rhs) ->
+  if lhs == x
+    then Just rhs
+    else Nothing
+
+countOccurrences :: Ord a => [a] -> Set (a, Int)
+countOccurrences = Set.fromList . go []
+  where
+    go acc [] = acc
+    go acc (x:xs) =
+      case lookupRemove x acc of
+        Nothing -> go ((x, 1) : acc) xs
+        Just (n, acc') -> go ((x, n+1) : acc') xs
+

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
topological sort data dependencies

@@ -38,11 +38,14 @@ codeGenSimple outputs (PikaCore.SslAssertion params asn0) =
   in
   map codeGenPointsTo sorted
 
-topologicalSortPointsTo :: [PointsToExpr a] -> [PointsToExpr a]
-topologicalSortPointsTo = map to . topologicalSort . map from
+topologicalSortPointsTo :: Ord a => [PointsToExpr a] -> [PointsToExpr a]
+topologicalSortPointsTo = map to . topologicalSortBy isLe . map from
   where
     to (x, y) = x :-> y
     from (x :-> y) = (x, y)
+
+    isLe (lhs1, rhs1) (lhs2, rhs2) =
+      all (`notElem` lhs2) (toList rhs1)
 
 codeGenPointsTo :: PointsToExpr a -> Command a
 codeGenPointsTo (lhs :-> rhs) =

### 2023-05-15 src/PikaC/Preorder/Preorder.hs

@@ -14,7 +14,9 @@ module PikaC.Preorder.Preorder
   ,preorderElements
   ,pMax
 
+  ,topologicalSortPairs
   ,topologicalSort
+  ,topologicalSortBy
 
   ,Contains (..)
   ,getContainsLabel
@@ -51,8 +53,8 @@ import GHC.Generics
 import Data.Ord
 import Data.Maybe
 
-topologicalSort :: Ord a => [(a, a)] -> [(a, a)]
-topologicalSort orig =
+topologicalSortPairs :: Ord a => [(a, a)] -> [(a, a)]
+topologicalSortPairs orig =
   filter (`elem` orig) .
   map go .
   getPreorder .
@@ -62,6 +64,17 @@ topologicalSort orig =
   map (uncurry (:<=)) orig
   where
     go (x :<= y) = (x, y)
+
+topologicalSort :: Ord a => [(a, a)] -> [a]
+topologicalSort = fastNub . concatMap to . topologicalSortPairs
+  where
+    to (x, y) = [x, y]
+
+topologicalSortBy :: Ord a => (a -> a -> Bool) -> [a] -> [a]
+topologicalSortBy isLe xs =
+  topologicalSort (concat (go <$> xs <*> xs))
+  where
+    go x y = [(x, y) | isLe x y]
 
 data Le a = a :<= a
   deriving (Show, Eq, Ord, Foldable, Generic)

### 2023-05-15 src/PikaC/Syntax/Heaplet.hs

@@ -16,7 +16,7 @@ import Control.Monad.Identity
 --   deriving (Show, Functor)
 
 newtype LayoutArg a = LayoutArg [a]
-  deriving (Show, Functor, Semigroup, Monoid, Foldable)
+  deriving (Show, Functor, Semigroup, Monoid, Foldable, Eq, Ord)
 
 class LayoutRename f where
   renameLayoutArg ::
@@ -39,10 +39,10 @@ renameLayoutArg' (LayoutArg old) (LayoutArg new) x =
       assocs = zip old new
 
 data PointsTo f a = Loc a :-> f a
-  deriving (Show)
+  deriving (Show, Foldable)
 
 data Loc a = a :+ Int
-  deriving (Show, Functor)
+  deriving (Show, Functor, Eq, Ord, Foldable)
 
 instance LayoutRename f => LayoutRename (PointsTo f) where
   renameLayoutArg old new (lhs :-> rhs) =

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -1,9 +1,12 @@
+{-# LANGUAGE DeriveFoldable #-}
+
 module PikaC.Syntax.PikaCore.Expr
   where
 
 import PikaC.Syntax.Heaplet
 import PikaC.Ppr
 import Data.List
+import Data.Foldable
 
 data Base a
   -- = V a
@@ -16,7 +19,7 @@ data Base a
   | Equal (Base a) (Base a)
   | Not (Base a)
   | And (Base a) (Base a)
-  deriving (Show, Functor)
+  deriving (Show, Functor, Eq, Ord, Foldable)
 
 data SimpleExpr a
   = BaseExpr (Base a)

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
Code gen for LayoutV

@@ -18,7 +18,7 @@ type Outputs = LayoutArg
 codeGenBase :: HasCallStack =>
   PikaCore.Base a ->
   CExpr a
--- codeGenBase (PikaCore.V x) = C.V x
+codeGenBase (PikaCore.V x) = C.V x
 codeGenBase (PikaCore.LayoutV xs) = error "codeGenBase: LayoutV"
 codeGenBase (PikaCore.IntLit i) = C.IntLit i
 codeGenBase (PikaCore.BoolLit b) = C.BoolLit b
@@ -29,12 +29,20 @@ codeGenBase (PikaCore.Not x) = C.Not (codeGenBase x)
 codeGenBase (PikaCore.And x y) = C.And (codeGenBase x) (codeGenBase y)
 
 codeGenSimple :: (HasCallStack, Ord a) => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
+codeGenSimple outputs (PikaCore.BaseExpr (PikaCore.LayoutV xs)) =
+  codeGenAsn $ connectLayoutArgs outputs xs
+
 codeGenSimple outputs (PikaCore.WithIn vars bnd body) =
   codeGenExpr vars bnd ++ codeGenSimple outputs body
 
 codeGenSimple outputs (PikaCore.SslAssertion params asn0) =
   let asn = fmap (renameLayoutArg params outputs) asn0
-      sorted = topologicalSortPointsTo asn
+  in
+  codeGenAsn asn
+
+codeGenAsn :: Ord a => PikaCore.ExprAssertion a -> [Command a]
+codeGenAsn asn =
+  let sorted = topologicalSortPointsTo asn
   in
   map codeGenPointsTo sorted
 
@@ -58,7 +66,6 @@ codeGenExpr outputs (PikaCore.App f xs) =
     -- TODO: Translate PikaCore function names into C function names
   [C.Call f (map C.V . toList $ (fold xs <> outputs))]
 
--- assignToOutputs :: Outputs a -> [PikaCore.Expr a] -> [Command a]
--- assignToOutputs (LayoutArg []) [] = []
--- assignToOutputs (LayoutArg (outputVar:vars)) (e:es) = undefined
+connectLayoutArgs :: LayoutArg a -> LayoutArg a -> PikaCore.ExprAssertion a
+connectLayoutArgs (LayoutArg xs) (LayoutArg ys) = zipWith (:->) (map (:+ 0) xs) (map PikaCore.V ys)
 

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -9,8 +9,8 @@ import Data.List
 import Data.Foldable
 
 data Base a
-  -- = V a
-  = LayoutV (LayoutArg a)    -- {x ...}
+  = V a
+  | LayoutV (LayoutArg a)    -- {x ...}
   | IntLit Int -- TODO: Add output locations?
   | BoolLit Bool
 
@@ -43,7 +43,7 @@ type PointsToExpr a = PointsTo Base a
 type ExprAssertion a = [PointsToExpr a]
 
 instance LayoutRename Base where
-  -- renameLayoutArg old new (V x) = V x
+  renameLayoutArg old new (V x) = V x
   renameLayoutArg old new (LayoutV xs) =
     LayoutV $ renameLayoutArg old new xs
   renameLayoutArg old new (IntLit i) = IntLit i
@@ -87,7 +87,7 @@ instance Ppr a => Ppr (LayoutArg a) where
   ppr (LayoutArg xs) = text "{" <+> hsep (punctuate (text ",") (map ppr xs)) <+> text "}"
 
 instance Ppr a => Ppr (Base a) where
-  -- ppr (V x) = ppr x
+  ppr (V x) = ppr x
   ppr (LayoutV x) = ppr x
   ppr (IntLit i) = ppr i
   ppr (BoolLit b) = ppr b
@@ -119,7 +119,7 @@ instance IsNested (SimpleExpr a) where
   isNested (SslAssertion {}) = True
 
 instance IsNested (Base a) where
-  -- isNested (V _) = False
+  isNested (V _) = False
   isNested (LayoutV _) = False
   isNested (IntLit _) = False
   isNested (BoolLit _) = False

### 2023-05-15 src/PikaC/Backend/C/CodeGen.hs
Fix a topological sorting bug

@@ -28,7 +28,7 @@ codeGenBase (PikaCore.Equal x y) = C.Equal (codeGenBase x) (codeGenBase y)
 codeGenBase (PikaCore.Not x) = C.Not (codeGenBase x)
 codeGenBase (PikaCore.And x y) = C.And (codeGenBase x) (codeGenBase y)
 
-codeGenSimple :: (HasCallStack, Ord a) => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
+codeGenSimple :: (HasCallStack, Ord a, Show a) => Outputs a -> PikaCore.SimpleExpr a -> [Command a]
 codeGenSimple outputs (PikaCore.BaseExpr (PikaCore.LayoutV xs)) =
   codeGenAsn $ connectLayoutArgs outputs xs
 
@@ -40,26 +40,27 @@ codeGenSimple outputs (PikaCore.SslAssertion params asn0) =
   in
   codeGenAsn asn
 
-codeGenAsn :: Ord a => PikaCore.ExprAssertion a -> [Command a]
+codeGenAsn :: (Show a, Ord a) => PikaCore.ExprAssertion a -> [Command a]
 codeGenAsn asn =
   let sorted = topologicalSortPointsTo asn
   in
   map codeGenPointsTo sorted
 
-topologicalSortPointsTo :: Ord a => [PointsToExpr a] -> [PointsToExpr a]
-topologicalSortPointsTo = map to . topologicalSortBy isLe . map from
+topologicalSortPointsTo :: (Show a, Ord a) => [PointsToExpr a] -> [PointsToExpr a]
+topologicalSortPointsTo xs =
+    map to . topologicalSortBy isLe . map from $ xs
   where
     to (x, y) = x :-> y
     from (x :-> y) = (x, y)
 
     isLe (lhs1, rhs1) (lhs2, rhs2) =
-      all (`notElem` lhs2) (toList rhs1)
+      all (`notElem` (getLocs lhs2)) (getLocs rhs1)
 
 codeGenPointsTo :: PointsToExpr a -> Command a
 codeGenPointsTo (lhs :-> rhs) =
   C.Assign lhs (codeGenBase rhs)
 
-codeGenExpr :: (HasCallStack, Ord a) => Outputs a -> PikaCore.Expr a -> [Command a]
[…]

### 2023-05-15 src/PikaC/Preorder/Preorder.hs

@@ -17,6 +17,7 @@ module PikaC.Preorder.Preorder
   ,topologicalSortPairs
   ,topologicalSort
   ,topologicalSortBy
+  ,preorderBy
 
   ,Contains (..)
   ,getContainsLabel
@@ -59,22 +60,38 @@ topologicalSortPairs orig =
   map go .
   getPreorder .
   containsToPreorder .
-  preorderToContains .
-  preorder $
-  map (uncurry (:<=)) orig
+  preorderToContains $
+  thePreorder
   where
+    thePreorder = preorder $ map (uncurry (:<=)) orig
     go (x :<= y) = (x, y)
 
 topologicalSort :: Ord a => [(a, a)] -> [a]
-topologicalSort = fastNub . concatMap to . topologicalSortPairs
+topologicalSort orig = fastNub . combine (preorderElements thePreorder) . concatMap to . topologicalSortPairs $ orig
   where
+    thePreorder = preorder $ map (uncurry (:<=)) orig
     to (x, y) = [x, y]
+    combine [] ys = ys
+    combine (x:xs) ys =
+      if x `notElem` ys
+        then x : combine xs ys
+        else combine xs ys
+
 
 topologicalSortBy :: Ord a => (a -> a -> Bool) -> [a] -> [a]
-topologicalSortBy isLe xs =
-  topologicalSort (concat (go <$> xs <*> xs))
+topologicalSortBy isLe =
+  topologicalSort . map to . leListBy isLe
   where
-    go x y = [(x, y) | isLe x y]
+    to (x :<= y) = (x, y)
+
+preorderBy :: Ord a => (a -> a -> Bool) -> [a] -> Preorder a
+preorderBy isLe =
+  preorder . leListBy isLe
+
+leListBy :: Eq a => (a -> a -> Bool) -> [a] -> [Le a]
+leListBy isLe xs = concat (go <$> xs <*> xs)
+  where
+    go x y = [x :<= y | isLe x y && x /= y]
 
 data Le a = a :<= a
   deriving (Show, Eq, Ord, Foldable, Generic)

### 2023-05-15 src/PikaC/Syntax/Heaplet.hs

@@ -44,6 +44,18 @@ data PointsTo f a = Loc a :-> f a
 data Loc a = a :+ Int
   deriving (Show, Functor, Eq, Ord, Foldable)
 
+class HasLocs f where
+  getLocs :: f a -> [Loc a]
+
+instance HasLocs Loc where
+  getLocs x = [x]
+
+instance HasLocs LayoutArg where
+  getLocs (LayoutArg xs) = map (:+ 0) xs
+
+instance HasLocs f => HasLocs (PointsTo f) where
+  getLocs (lhs :-> rhs) = lhs : getLocs rhs
+
 instance LayoutRename f => LayoutRename (PointsTo f) where
   renameLayoutArg old new (lhs :-> rhs) =
     renameLayoutArg old new lhs :-> renameLayoutArg old new rhs

### 2023-05-15 src/PikaC/Syntax/PikaCore/Expr.hs

@@ -42,6 +42,19 @@ data Expr a
 type PointsToExpr a = PointsTo Base a
 type ExprAssertion a = [PointsToExpr a]
 
+instance HasLocs Base where
+  getLocs (V x) = [x :+ 0]
+  getLocs (LayoutV x) = getLocs x
+  getLocs (IntLit i) = []
+  getLocs (BoolLit b) = []
+  getLocs (Add x y) = getLocs x ++ getLocs y
+  getLocs (Sub x y) = getLocs x ++ getLocs y
+  getLocs (Equal x y) = getLocs x ++ getLocs y
+  getLocs (Not x) = getLocs x
+  getLocs (And x y) = getLocs x ++ getLocs y
+
+  
+
 instance LayoutRename Base where
   renameLayoutArg old new (V x) = V x
   renameLayoutArg old new (LayoutV xs) =

### 2023-08-03 outline/outline.tex
outline

\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{syntax}
\usepackage{listings}

\lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}

\input{defs.tex}

\begin{document}

\title{Reflections on Pika}
\maketitle

\section{Introduction}

There is a tool called \SuSLik{} that will take separation logic specification and generate
a program in a C-like language. While working with this tool, it became clear that there is a strong
connection between the form of separation logic used by \SuSLik{} and functional programming languages such as Haskell and Standard ML.

In particular, we made the following observation. The SuSLik specification language has \textit{inductive predicates}. These are
allow the programmer to define the heap layout of various inductively defined data structures, such as singly-linked lists and binary trees.
However, we noticed that inductive predicates could also encode function definitions!

Using this fact, functions written in a functional programming language can be translated into inductive predicates. The translation
of a function definition into an inductive predicate is very similar to translation into ANF. However, this also gives us easy
interoperability with the rest of the SuSLik specification language. For example, not only can we define data structures as inductive
predicates, we can also dictate the heap layout of our data structures.

Finally, this enables us to implement a program synthesis feature at the level of the functional programming language.

\section{Examples}

\section{Type System}

\section{Intermediate Representation}

\section{Translation to SSL}

\section{Translation to C}

\section{\texttt{synth} keyword}

\section{Formal Semantics}

% \input{introduction.tex}
% \input{language.tex}
% \input{types.tex}
% \input{elaboration.tex}
% \input{semantics.tex}
% \input{codegen.tex}
% \input{related-work.tex}

\bibliographystyle{splncs04}
\bibliography{outline}

\end{document}

### 2023-08-03 outline/outline.tex
outline title
A bit more on the intro

@@
 
-\title{Reflections on Pika}
+\title{Reflections On Synthesis With Separation Logic}
 \maketitle
@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 interoperability with the rest of the SuSLik specification language. For example, not only can we define data structures as inductive
-predicates, we can also dictate the heap layout of our data structures.
+predicates, we can also dictate the heap layout of our data structures (similar to \Dargent\cite{Dargent}). These ideas are explored
+with examples in Section~\ref{sec:examples}. The representation of data types as inductive predicates are explored in more formal detail
+in Section~\ref{sec:types}.
 
-Finally, this enables us to implement a program synthesis feature at the level of the functional programming language.
+We develop an ANF-like intermediate language. This also enables a straightforward translation into C (Section~\ref{sec:IR}).
+In Section~\ref{sec:to-ssl} and Section~\ref{sec:to-c} we describe the two translations from this intermediate representation.
+
+Finally, this enables us to implement a program synthesis feature at the level of the functional programming language (Section~\ref{sec:synth}).
+The form this takes is similar to synthesis tools based on refinement types (such as \Synquid), but based on separation logic.
 
 \section{Examples}
+\label{sec:examples}
 
 \section{Type System}
+\label{sec:types}
 
 \section{Intermediate Representation}
+\label{sec:IR}
 
 \section{Translation to SSL}
+\label{sec:to-ssl}
 
 \section{Translation to C}
+\label{sec:to-c}
 
 \section{\texttt{synth} keyword}
+\label{sec:synth}
 
 \section{Formal Semantics}
+\label{sec:semantics}
+
+\section{Conclusion}
+\label{sec:conclusion}
 
@@
 \bibliographystyle{splncs04}
-\bibliography{outline}
+\bibliography{refs}
 

### 2023-08-21 new-paper/paper.tex
New paper

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{graphicx}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{syntax}
\usepackage{listings}
\lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}

\input{defs.tex}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Compilation With Program Synthesis}

% %%
% %% The "author" command and its associated commands are used to define
% %% the authors and their affiliations.
% %% Of note is the shared affiliation of the first two authors, and the
% %% "authornote" and "authornotemark" commands
% %% used to denote shared contribution to the research.
% \author{Ben Trovato}
% \authornote{Both authors contributed equally to this research.}
% \email{trovato@corporation.com}
% \orcid{1234-5678-9012}
% \author{G.K.M. Tobin}
% \authornotemark[1]
% \email{webmaster@marysville-ohio.com}
% \affiliation{%
%   \institution{Institute for Clarity in Documentation}
%   \streetaddress{P.O. Box 1212}
%   \city{Dublin}
%   \state{Ohio}
%   \country{USA}
%   \postcode{43017-6221}
% }
%
% \author{Lars Th{\o}rv{\"a}ld}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{larst@affiliation.org}
%
% \author{Valerie B\'eranger}
% \affiliation{%
%   \institution{Inria Paris-Rocquencourt}
%   \city{Rocquencourt}
%   \country{France}
% }
%
% \author{Aparna Patel}
% \affiliation{%
%  \institution{Rajiv Gandhi University}
%  \streetaddress{Rono-Hills}
%  \city{Doimukh}
%  \state{Arunachal Pradesh}
%  \country{India}}
%
% \author{Huifen Chan}
% \affiliation{%
%   \institution{Tsinghua University}
%   \streetaddress{30 Shuangqing Rd}
%   \city{Haidian Qu}
%   \state{Beijing Shi}
%   \country{China}}
%
% \author{Charles Palmer}
% \affiliation{%
%   \institution{Palmer Research Laboratories}
%   \streetaddress{8600 Datapoint Drive}
%   \city{San Antonio}
%   \state{Texas}
%   \country{USA}
%   \postcode{78229}}
% \email{cpalmer@prl.com}
%
% \author{John Smith}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{jsmith@affiliation.org}
%
% \author{Julius P. Kumquat}
% \affiliation{%
%   \institution{The Kumquat Consortium}
%   \city{New York}
%   \country{USA}}
% \email{jpkumquat@consortium.net}
%
% %%
% %% By default, the full list of authors will be used in the page
% %% headers. Often, this list is too long, and will overlap
% %% other information printed in the page headers. This command allows
% %% the author to define a more concise list
% %% of authors' names for this purpose.
% \renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Synthetic Separation Logic (SSL) is a specification language used by the \SuSLik{} program synthesis tool.
  \SuSLik{} will take a Hoare-style specification expressed in SSL and synthesize a program that satisfies this
  specification in a C-like language. However, these specifications are very low-level and SSL does not allow for much abstraction. In this paper,
  we explore a connection between SSL specifications and functional programming. We give a first-order functional
  programming language, \Pika, which has both a standard semantics and a translation into SSL that respects the standard semantics.

  Along the way, we encounter limitations in the \SuSLik{} tool. In confronting these limitations, we develop a new feature
  in \Pika{} that allows the programmer to express synthesis tasks at a higher-level.

  % A clear and well-documented \LaTeX\ document is presented as an
  % article formatted for publication by ACM in a conference proceedings
  % or journal publication. Based on the ``acmart'' document class, this
  % article presents and explains many of the common variations, as well
  % as many of the formatting elements an author may use in the
  % preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>00000000.0000000.0000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
\end{CCSXML}

% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

% %%
% %% Keywords. The author(s) should pick words that accurately describe
% %% the work being presented. Separate the keywords with commas.
% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
%   Your, Paper}

% \received{20 February 2007}
% \received[revised]{12 March 2009}
% \received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

Synthetic separation logic (SSL) is the specification language for the \SuSLik{} program synthesizer. \SuSLik{} will take a precondition and postcondition pair together
with a function prototype and use these to synthesize a function that satisfies that specification in a C-like language called \SuSLang.

We have developed a first-order functional programming language, \Pika, which will compile to SSL specifications. \SuSLik{} can then take these SSL specifications and generate \SuSLang. Finally, the resulting \SuSLang{} code can be converted to C and then run. In addition to this translational semantics, we also provide a standard abstract machine semantics for \Pika{} and use this to show correctness of the translation into SSL.

\paragraph{Contributions} In this paper, we make the following contributions:
\begin{itemize}
  \item We explore the connection between SSL and functional programming using a pair of formal semantics for \Pika: A standard abstract machine semantics and a
    semantics that translates \Pika{} into SSL specifications. The second semantics is shown to be sound relative to the first semantics. (Section~\ref{sec:semantics})

  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic datatypes and their representation in memory.
    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic datatype. (Section~\ref{sec:layouts})

  \item We address the limitations of this approach. In particular, we develop an extension of \Pika{} that allows \SuSLik's synthesis capabilities to be exposed more naturally inside the \Pika{} language. (Section~\ref{sec:synth})

  \item A comparison of examples that work on the first version of \Pika{} is compared with examples that are enabled by the new feature in the second version of \Pika. This new version of \Pika{} is not backwards compatible with the first version, so we also examine which examples in the old version no longer work in the new version. (Section~\ref{sec:examples})
\end{itemize}

\section{Overview}
\label{sec:overview}

\subsection{Syntax}
\label{sec:syntax}

\subsection{Layouts}
\label{sec:layouts}

\section{Formal Semantics}
\label{sec:semantics}

\section{Synthesis Limitations And Solutions}
\label{sec:limitations}

There is a significant limitation to our approach so far: \SuSLik will often fail to synthesize code from valid specifications. One instance of this problem is the following: In a certain class of functions, \SuSLik{} requires a different number of pointer indirections than it will with other functions. We have not found a way to perfectly characterize such functions and, since this will change the way the functions will need to be called, this represents a major impediment to our approach. It is worth noting that this is also not the \textit{only} situation in which \SuSLik{} fails to synthesize a program.

From this, we draw the conclusion that \SuSLik{} is not well-suited to this task. In fact, this task is very different from the original goals of \SuSLik, which could be stated: \textit{Given a potentially ambiguous specification, generate any program that satisfies this specification.}

However, in our case, we already know which specific program we want to generate. This is fully specified by the abstract machine semantics for \Pika.

With this in mind, we take a new approach. What if we compile \Pika{} directly to C, but we add a \textit{new} feature that allows us to express synthesis problems at the level of \Pika{} code? This takes the form of the \verb|synth| keyword, which we explain in detail in Section~\ref{sec:synth}. In this case, we can get the advantages of a traditional compiler while also exploiting the synthesis capabilities of \SuSLik. Furthermore, due to \Pika's design, we already have easy interoperability with \SuSLik. Also, recall that \SuSLik's target language, \SuSLang, is already very similar to C. This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.

\section{\texttt{synth} Keyword}
\label{sec:synth}

\section{Examples}
\label{sec:examples}

\subsection{Version 1}

\subsection{Version 2}

\section{Future Work}
\label{sec:future-work}

\section{Conclusion}
\label{sec:conclusion}

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2023-08-21 new-paper/paper.tex
synth syntax

@@
 From this, we draw the conclusion that \SuSLik{} is not well-suited to this task.
-In fact, this task is very different from the original goals of \SuSLik, which could be stated: \textit{Given a potentially ambiguous specification, generate any program that satisfies this specification.}
-
-However, in our case, we already know which specific program we want to generate. This is fully specified by the abstract machine semantics for \Pika.
+In fact, this task is very different from the original goals of \SuSLik, which could be stated: \textit{Given a potentially ambiguous specification, generate any program that satisfies this specification.} However, in our case, we already know which specific program we want to generate.
+This is fully specified by the abstract machine semantics for \Pika.
 
@@
 Furthermore, due to \Pika's design, we already have easy interoperability with \SuSLik.
-Also, recall that \SuSLik's target language, \SuSLang, is already very similar to C.
+Also, recall that \SuSLik's target language, \SuSLang, can be straightforwardly translated directly to C.
 This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.
 
-\section{\texttt{synth} Keyword}
+\section{\synth{} Keyword: Synthesis Inside \Pika}
 \label{sec:synth}
+
+Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature.
+This special kind of type signature allows us to directly access \SuSLik's pure propositions.
+These type signatures are called \synth-signatures and the syntax for them is
+\[
+  \synth\; f : \sigma
+\]
+\noindent
+where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is as follows.
+\[
+\begin{aligned}
+  \sigma &::= p \synthsep \tau\\
+  p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
+    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p
+\end{aligned}
+\]
 

### 2023-08-21 new-paper/paper.tex
synth example

@@
 This special kind of type signature allows us to directly access \SuSLik's pure propositions.
-These type signatures are called \synth-signatures and the syntax for them is
+These type signatures are called \synth-signatures and the syntax for invoking the \Pika{} synthesis on a \synth-signature is:
 \[
@@
 \noindent
-where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is as follows.
+where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is given in Figure~\ref{fig:synth-sig-syntax}.
+
+\begin{figure}
 \[
 \begin{aligned}
-  \sigma &::= p \synthsep \tau\\
-  p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
-    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p
+  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
+  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
+    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
+  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
+  \textit{Ghost variables}\;\; G &::= @x
 \end{aligned}
 \]
+  \caption{\synth-signature syntax}
+  \label{fig:synth-sig-syntax}
+\end{figure}
+
+\noindent
+We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
+
+This extension to layouts is best explain by way of example. First, consider the original definition of the singly-linked list type:
+
+\begin{lstlisting}
+Sll : layout(List)
+Sll Nil := emp;
+Sll (Cons head tail) :=
+  x :-> head ** (x+1) :-> nxt ** Sll tail [nxt];
+\end{lstlisting}
+
+\noindent
+This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification.
+A common pattern in \SuSLik{} specifications is to use a \verb|set| type ghost variable to keep track of the elements in a data structure.
+Then, we can require that the elements are preserved or that new elements be added when using this in specifications.
+The feature we add to layouts allows us to use this feature directly:
+
+\begin{lstlisting}
+Sll : @(s : set) layout(List);
+Sll Nil := s == {} ;; emp;
+Sll (Cons head tail) :=
+  s == {head} ++ s1
+    ;;
+  x :-> head ** (x+1) :-> nxt ** Sll tail @s1 [nxt];
+\end{lstlisting}
+
+\noindent
+Putting this all together, we can use this new kind of parameter together with the \synth{} keyword to specify a list append function:
+
+\begin{lstlisting}
+synth append :
+  (s == (s1 ++ s2))
+    ;;
+  Sll @s1 -> Sll @s2 -> Sll @s
+\end{lstlisting}
 

### 2023-08-21 new-paper/paper.tex
clarify with formatting
typo
a small fix

@@
 \begin{lstlisting}
-Sll : layout(List)
+Sll : layout[x](List)
 Sll Nil := emp;
 Sll (Cons head tail) :=
-  x :-> head ** (x+1) :-> nxt ** Sll tail [nxt];
+  x :-> head ** (x+1) :-> nxt
+    ** Sll tail [nxt];
 \end{lstlisting}
@@
 \begin{lstlisting}
-Sll : @(s : set) layout(List);
+Sll : @(s : set) layout[x](List);
 Sll Nil := s == {} ;; emp;
@@
     ;;
-  x :-> head ** (x+1) :-> nxt ** Sll tail @s1 [nxt];
+  x :-> head ** (x+1) :-> nxt
+    ** Sll tail @s1 [nxt];
 \end{lstlisting}
@@
     ;;
-  Sll @s1 -> Sll @s2 -> Sll @s
+  Sll @s1 -> Sll @s2 -> Sll @s;
 \end{lstlisting}

### 2023-08-21 new-paper/paper.tex
remove unnecessary parentheses
summarize section

@@
 synth append :
-  (s == (s1 ++ s2))
+  s == s1 ++ s2
     ;;
@@
 \end{lstlisting}
+
+\noindent
+Using this feature, we can use \SuSLik{} specifically for \textit{synthesis} tasks inside \Pika. That is, we are able to directly use \SuSLik{} for its intended purpose.
 

### 2023-08-21 new-paper/conclusion.tex
Expand future work with some existing material; separate out into multiple .tex files

\section{Conclusion}
\label{sec:conclusion}

We have seen a technique for generating an SSL specification from a functional program, such that this translation is sound with respect to the standard semantics. In order to do this, we built the notion of a \textit{layout} of an algebraic datatype on top of separation logic. Then we observed that \SuSLik, the tool we use to synthesize programs from SSL, has notable limitations in its ability to synthesize the SSL specifications we generate. This motivated the creation of the \synth{} keyword. This new keyword allows us to have direct access to the synthesis features of \SuSLik{} from inside the \Pika{} programming language.

### 2023-08-21 new-paper/examples.tex
Expand future work with some existing material; separate out into multiple .tex files

\section{Examples}
\label{sec:examples}

\subsection{\Pika{} 1}

\subsection{\Pika{} 2}

Version 2 of \Pika{} is a rewrite of version 1 and it is missing a couple of features of version 1. However, it has the \synth{} feature. It also has the groundwork for layout polymorphism to be implemented. The second aspect, in particular, is a substantial change from version 1 and motivated the rewrite.

### 2023-08-21 new-paper/introduction.tex
Expand future work with some existing material; separate out into multiple .tex files

--- 2023-08-21 new-paper/paper.tex
@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall]{acmart}
-\usepackage{graphicx}
-\usepackage{proof}
-\usepackage{stmaryrd}
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{xcolor}
-\usepackage{tcolorbox}
-\usepackage{tikz}
-\usepackage{tikz-cd}
-\usepackage{syntax}
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}
-
-\input{defs.tex}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%%
-%% These commands are for a JOURNAL article.
-\acmJournal{JACM}
-\acmVolume{37}
-\acmNumber{4}
-\acmArticle{111}
-\acmMonth{8}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{Compilation With Program Synthesis}
-
-% %%
-% %% The "author" command and its associated commands are used to define
-% %% the authors and their affiliations.
-% %% Of note is the shared affiliation of the first two authors, and the
-% %% "authornote" and "authornotemark" commands
-% %% used to denote shared contribution to the research.
-% \author{Ben Trovato}
-% \authornote{Both authors contributed equally to this research.}
-% \email{trovato@corporation.com}
-% \orcid{1234-5678-9012}
-% \author{G.K.M. Tobin}
-% \authornotemark[1]
-% \email{webmaster@marysville-ohio.com}
-% \affiliation{%
-%   \institution{Institute for Clarity in Documentation}
-%   \streetaddress{P.O. Box 1212}
-%   \city{Dublin}
-%   \state{Ohio}
-%   \country{USA}
-%   \postcode{43017-6221}
-% }
-%
-% \author{Lars Th{\o}rv{\"a}ld}
-% \affiliation{%
-%   \institution{The Th{\o}rv{\"a}ld Group}
-%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-%   \city{Hekla}
-%   \country{Iceland}}
-% \email{larst@affiliation.org}
-%
-% \author{Valerie B\'eranger}
-% \affiliation{%
-%   \institution{Inria Paris-Rocquencourt}
-%   \city{Rocquencourt}
-%   \country{France}
-% }
-%
-% \author{Aparna Patel}
-% \affiliation{%
-%  \institution{Rajiv Gandhi University}
-%  \streetaddress{Rono-Hills}
-%  \city{Doimukh}
-%  \state{Arunachal Pradesh}
-%  \country{India}}
-%
-% \author{Huifen Chan}
-% \affiliation{%
-%   \institution{Tsinghua University}
-%   \streetaddress{30 Shuangqing Rd}
-%   \city{Haidian Qu}
-%   \state{Beijing Shi}
-%   \country{China}}
-%
-% \author{Charles Palmer}
-% \affiliation{%
-%   \institution{Palmer Research Laboratories}
-%   \streetaddress{8600 Datapoint Drive}
-%   \city{San Antonio}
-%   \state{Texas}
-%   \country{USA}
-%   \postcode{78229}}
-% \email{cpalmer@prl.com}
-%
-% \author{John Smith}
-% \affiliation{%
-%   \institution{The Th{\o}rv{\"a}ld Group}
-%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-%   \city{Hekla}
-%   \country{Iceland}}
-% \email{jsmith@affiliation.org}
-%
-% \author{Julius P. Kumquat}
-% \affiliation{%
-%   \institution{The Kumquat Consortium}
-%   \city{New York}
-%   \country{USA}}
-% \email{jpkumquat@consortium.net}
-%
-% %%
-% %% By default, the full list of authors will be used in the page
-% %% headers. Often, this list is too long, and will overlap
-% %% other information printed in the page headers. This command allows
-% %% the author to define a more concise list
-% %% of authors' names for this purpose.
-% \renewcommand{\shortauthors}{Trovato et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  Synthetic Separation Logic (SSL) is a specification language used by the \SuSLik{} program synthesis tool.
-  \SuSLik{} will take a Hoare-style specification expressed in SSL and synthesize a program that satisfies this
-  specification in a C-like language. However, these specifications are very low-level and SSL does not allow for much abstraction. In this paper,
-  we explore a connection between SSL specifications and functional programming. We give a first-order functional
-  programming language, \Pika, which has both a standard semantics and a translation into SSL that respects the standard semantics.
-
-  Along the way, we encounter limitations in the \SuSLik{} tool. In confronting these limitations, we develop a new feature
-  in \Pika{} that allows the programmer to express synthesis tasks at a higher-level.
-
-  % A clear and well-documented \LaTeX\ document is presented as an
-  % article formatted for publication by ACM in a conference proceedings
-  % or journal publication. Based on the ``acmart'' document class, this
-  % article presents and explains many of the common variations, as well
-  % as many of the formatting elements an author may use in the
-  % preparation of the documentation of their work.
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-% <ccs2012>
-%  <concept>
-%   <concept_id>00000000.0000000.0000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>500</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>300</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-% </ccs2012>
-\end{CCSXML}
-
-% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-
-% %%
-% %% Keywords. The author(s) should pick words that accurately describe
-% %% the work being presented. Separate the keywords with commas.
-% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-%   Your, Paper}
-
-% \received{20 February 2007}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
 \section{Introduction}
@@
 \end{itemize}
-
-\section{Overview}
-\label{sec:overview}
-
-\subsection{Syntax}
-\label{sec:syntax}
-
-\subsection{Layouts}
-\label{sec:layouts}
-
-\section{Formal Semantics}
-\label{sec:semantics}
-
-\section{Synthesis Limitations And Solutions}
-\label{sec:limitations}
-
-There is a significant limitation to our approach so far: \SuSLik will often fail to synthesize code from valid specifications.
-One instance of this problem is the following: In a certain class of functions, \SuSLik{} requires a different number of pointer indirections than it will with other functions.
-We have not found a way to perfectly characterize such functions and, since this will change the way the functions will need to be called, this represents a major impediment to our approach.
-It is worth noting that this is also not the \textit{only} situation in which \SuSLik{} fails to synthesize a program.
-
-From this, we draw the conclusion that \SuSLik{} is not well-suited to this task.
-In fact, this task is very different from the original goals of \SuSLik, which could be stated: \textit{Given a potentially ambiguous specification, generate any program that satisfies this specification.} However, in our case, we already know which specific program we want to generate.
-This is fully specified by the abstract machine semantics for \Pika.
-
-With this in mind, we take a new approach.
-What if we compile \Pika{} directly to C, but we add a \textit{new} feature that allows us to express synthesis problems at the level of \Pika{} code?
-This takes the form of the \verb|synth| keyword, which we explain in detail in Section~\ref{sec:synth}.
-In this case, we can get the advantages of a traditional compiler while also exploiting the synthesis capabilities of \SuSLik.
-Furthermore, due to \Pika's design, we already have easy interoperability with \SuSLik.
-Also, recall that \SuSLik's target language, \SuSLang, can be straightforwardly translated directly to C.
-This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.
-
-\section{\synth{} Keyword: Synthesis Inside \Pika}
-\label{sec:synth}
-
-Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature.
-This special kind of type signature allows us to directly access \SuSLik's pure propositions.
-These type signatures are called \synth-signatures and the syntax for invoking the \Pika{} synthesis on a \synth-signature is:
-\[
-  \synth\; f : \sigma
-\]
-\noindent
-where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is given in Figure~\ref{fig:synth-sig-syntax}.
-
-\begin{figure}
-\[
-\begin{aligned}
-  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
-  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
-    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
-  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
-  \textit{Ghost variables}\;\; G &::= @x
-\end{aligned}
-\]
-  \caption{\synth-signature syntax}
-  \label{fig:synth-sig-syntax}
-\end{figure}
-
-\noindent
-We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
-
-This extension to layouts is best explain by way of example. First, consider the original definition of the singly-linked list type:
-
-\begin{lstlisting}
-Sll : layout(List)
-Sll Nil := emp;
-Sll (Cons head tail) :=
-  x :-> head ** (x+1) :-> nxt ** Sll tail [nxt];
-\end{lstlisting}
-
-\noindent
-This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification.
-A common pattern in \SuSLik{} specifications is to use a \verb|set| type ghost variable to keep track of the elements in a data structure.
-Then, we can require that the elements are preserved or that new elements be added when using this in specifications.
-The feature we add to layouts allows us to use this feature directly:
-
-\begin{lstlisting}
-Sll : @(s : set) layout(List);
-Sll Nil := s == {} ;; emp;
-Sll (Cons head tail) :=
-  s == {head} ++ s1
-    ;;
-  x :-> head ** (x+1) :-> nxt ** Sll tail @s1 [nxt];
-\end{lstlisting}
-
-\noindent
-Putting this all together, we can use this new kind of parameter together with the \synth{} keyword to specify a list append function:
-
-\begin{lstlisting}
-synth append :
-  (s == (s1 ++ s2))
-    ;;
-  Sll @s1 -> Sll @s2 -> Sll @s
-\end{lstlisting}
-
-\section{Examples}
-\label{sec:examples}
-
-\subsection{Version 1}
-
-\subsection{Version 2}
-
-\section{Future Work}
-\label{sec:future-work}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-acmsmall.tex'.

### 2023-08-21 new-paper/limitations.tex
Expand future work with some existing material; separate out into multiple .tex files

--- 2023-08-21 new-paper/paper.tex
@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall]{acmart}
-\usepackage{graphicx}
-\usepackage{proof}
-\usepackage{stmaryrd}
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{xcolor}
-\usepackage{tcolorbox}
-\usepackage{tikz}
-\usepackage{tikz-cd}
-\usepackage{syntax}
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}
-
-\input{defs.tex}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%%
-%% These commands are for a JOURNAL article.
-\acmJournal{JACM}
-\acmVolume{37}
-\acmNumber{4}
-\acmArticle{111}
-\acmMonth{8}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{Compilation With Program Synthesis}
-
-% %%
-% %% The "author" command and its associated commands are used to define
-% %% the authors and their affiliations.
-% %% Of note is the shared affiliation of the first two authors, and the
-% %% "authornote" and "authornotemark" commands
-% %% used to denote shared contribution to the research.
-% \author{Ben Trovato}
-% \authornote{Both authors contributed equally to this research.}
-% \email{trovato@corporation.com}
-% \orcid{1234-5678-9012}
-% \author{G.K.M. Tobin}
-% \authornotemark[1]
-% \email{webmaster@marysville-ohio.com}
-% \affiliation{%
-%   \institution{Institute for Clarity in Documentation}
-%   \streetaddress{P.O. Box 1212}
-%   \city{Dublin}
-%   \state{Ohio}
-%   \country{USA}
-%   \postcode{43017-6221}
-% }
-%
-% \author{Lars Th{\o}rv{\"a}ld}
-% \affiliation{%
-%   \institution{The Th{\o}rv{\"a}ld Group}
-%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-%   \city{Hekla}
-%   \country{Iceland}}
-% \email{larst@affiliation.org}
-%
-% \author{Valerie B\'eranger}
-% \affiliation{%
-%   \institution{Inria Paris-Rocquencourt}
-%   \city{Rocquencourt}
-%   \country{France}
-% }
-%
-% \author{Aparna Patel}
-% \affiliation{%
-%  \institution{Rajiv Gandhi University}
-%  \streetaddress{Rono-Hills}
-%  \city{Doimukh}
-%  \state{Arunachal Pradesh}
-%  \country{India}}
-%
-% \author{Huifen Chan}
-% \affiliation{%
-%   \institution{Tsinghua University}
-%   \streetaddress{30 Shuangqing Rd}
-%   \city{Haidian Qu}
-%   \state{Beijing Shi}
-%   \country{China}}
-%
-% \author{Charles Palmer}
-% \affiliation{%
-%   \institution{Palmer Research Laboratories}
-%   \streetaddress{8600 Datapoint Drive}
-%   \city{San Antonio}
-%   \state{Texas}
-%   \country{USA}
-%   \postcode{78229}}
-% \email{cpalmer@prl.com}
-%
-% \author{John Smith}
-% \affiliation{%
-%   \institution{The Th{\o}rv{\"a}ld Group}
-%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-%   \city{Hekla}
-%   \country{Iceland}}
-% \email{jsmith@affiliation.org}
-%
-% \author{Julius P. Kumquat}
-% \affiliation{%
-%   \institution{The Kumquat Consortium}
-%   \city{New York}
-%   \country{USA}}
-% \email{jpkumquat@consortium.net}
-%
-% %%
-% %% By default, the full list of authors will be used in the page
-% %% headers. Often, this list is too long, and will overlap
-% %% other information printed in the page headers. This command allows
-% %% the author to define a more concise list
-% %% of authors' names for this purpose.
-% \renewcommand{\shortauthors}{Trovato et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  Synthetic Separation Logic (SSL) is a specification language used by the \SuSLik{} program synthesis tool.
-  \SuSLik{} will take a Hoare-style specification expressed in SSL and synthesize a program that satisfies this
-  specification in a C-like language. However, these specifications are very low-level and SSL does not allow for much abstraction. In this paper,
-  we explore a connection between SSL specifications and functional programming. We give a first-order functional
-  programming language, \Pika, which has both a standard semantics and a translation into SSL that respects the standard semantics.
-
-  Along the way, we encounter limitations in the \SuSLik{} tool. In confronting these limitations, we develop a new feature
-  in \Pika{} that allows the programmer to express synthesis tasks at a higher-level.
-
-  % A clear and well-documented \LaTeX\ document is presented as an
-  % article formatted for publication by ACM in a conference proceedings
-  % or journal publication. Based on the ``acmart'' document class, this
-  % article presents and explains many of the common variations, as well
-  % as many of the formatting elements an author may use in the
-  % preparation of the documentation of their work.
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-% <ccs2012>
-%  <concept>
-%   <concept_id>00000000.0000000.0000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>500</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>300</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-%  <concept>
-%   <concept_id>00000000.00000000.00000000</concept_id>
-%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%   <concept_significance>100</concept_significance>
-%  </concept>
-% </ccs2012>
-\end{CCSXML}
-
-% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-
-% %%
-% %% Keywords. The author(s) should pick words that accurately describe
-% %% the work being presented. Separate the keywords with commas.
-% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-%   Your, Paper}
-
-% \received{20 February 2007}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-\section{Introduction}
-
-Synthetic separation logic (SSL) is the specification language for the \SuSLik{} program synthesizer. \SuSLik{} will take a precondition and postcondition pair together
-with a function prototype and use these to synthesize a function that satisfies that specification in a C-like language called \SuSLang.
-
-We have developed a first-order functional programming language, \Pika, which will compile to SSL specifications.
-\SuSLik{} can then take these SSL specifications and generate \SuSLang.
-Finally, the resulting \SuSLang{} code can be converted to C and then run.
-In addition to this translational semantics, we also provide a standard abstract machine semantics for \Pika{} and use this to show correctness of the translation into SSL.
-
-\paragraph{Contributions} In this paper, we make the following contributions:
-\begin{itemize}
-  \item We explore the connection between SSL and functional programming using a pair of formal semantics for \Pika: A standard abstract machine semantics and a
-    semantics that translates \Pika{} into SSL specifications. The second semantics is shown to be sound relative to the first semantics. (Section~\ref{sec:semantics})
-
-  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic datatypes and their representation in memory.
-    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic datatype. (Section~\ref{sec:layouts})
-
-  \item We address the limitations of this approach.
-In particular, we develop an extension of \Pika{} that allows \SuSLik's synthesis capabilities to be exposed more naturally inside the \Pika{} language.
-(Section~\ref{sec:synth})
-
-  \item A comparison of examples that work on the first version of \Pika{} is compared with examples that are enabled by the new feature in the second version of \Pika.
-This new version of \Pika{} is not backwards compatible with the first version, so we also examine which examples in the old version no longer work in the new version.
-(Section~\ref{sec:examples})
-\end{itemize}
-
-\section{Overview}
-\label{sec:overview}
-
-\subsection{Syntax}
-\label{sec:syntax}
-
-\subsection{Layouts}
-\label{sec:layouts}
-
-\section{Formal Semantics}
-\label{sec:semantics}
-
 \section{Synthesis Limitations And Solutions}
@@
 This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.
-
-\section{\synth{} Keyword: Synthesis Inside \Pika}
-\label{sec:synth}
-
-Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature.
-This special kind of type signature allows us to directly access \SuSLik's pure propositions.
-These type signatures are called \synth-signatures and the syntax for invoking the \Pika{} synthesis on a \synth-signature is:
-\[
-  \synth\; f : \sigma
-\]
-\noindent
-where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is given in Figure~\ref{fig:synth-sig-syntax}.
-
-\begin{figure}
-\[
-\begin{aligned}
-  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
-  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
-    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
-  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
-  \textit{Ghost variables}\;\; G &::= @x
-\end{aligned}
-\]
-  \caption{\synth-signature syntax}
-  \label{fig:synth-sig-syntax}
-\end{figure}
-
-\noindent
-We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
-
-This extension to layouts is best explain by way of example. First, consider the original definition of the singly-linked list type:
-
-\begin{lstlisting}
-Sll : layout(List)
-Sll Nil := emp;
-Sll (Cons head tail) :=
-  x :-> head ** (x+1) :-> nxt ** Sll tail [nxt];
-\end{lstlisting}
-
-\noindent
-This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification.
-A common pattern in \SuSLik{} specifications is to use a \verb|set| type ghost variable to keep track of the elements in a data structure.
-Then, we can require that the elements are preserved or that new elements be added when using this in specifications.
-The feature we add to layouts allows us to use this feature directly:
-
-\begin{lstlisting}
-Sll : @(s : set) layout(List);
-Sll Nil := s == {} ;; emp;
-Sll (Cons head tail) :=
-  s == {head} ++ s1
-    ;;
-  x :-> head ** (x+1) :-> nxt ** Sll tail @s1 [nxt];
-\end{lstlisting}
-
-\noindent
-Putting this all together, we can use this new kind of parameter together with the \synth{} keyword to specify a list append function:
-
-\begin{lstlisting}
-synth append :
-  (s == (s1 ++ s2))
-    ;;
-  Sll @s1 -> Sll @s2 -> Sll @s
-\end{lstlisting}
-
-\section{Examples}
-\label{sec:examples}
-
-\subsection{Version 1}
-
-\subsection{Version 2}
-
-\section{Future Work}
-\label{sec:future-work}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-acmsmall.tex'.

### 2023-08-21 new-paper/paper.tex
Expand future work with some existing material; separate out into multiple .tex files

--- 2023-08-21 new-paper/paper.tex
@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
 \documentclass[acmsmall]{acmart}
@@
 
-\section{Introduction}
-
-Synthetic separation logic (SSL) is the specification language for the \SuSLik{} program synthesizer. \SuSLik{} will take a precondition and postcondition pair together
-with a function prototype and use these to synthesize a function that satisfies that specification in a C-like language called \SuSLang.
-
-We have developed a first-order functional programming language, \Pika, which will compile to SSL specifications.
-\SuSLik{} can then take these SSL specifications and generate \SuSLang.
-Finally, the resulting \SuSLang{} code can be converted to C and then run.
-In addition to this translational semantics, we also provide a standard abstract machine semantics for \Pika{} and use this to show correctness of the translation into SSL.
-
-\paragraph{Contributions} In this paper, we make the following contributions:
-\begin{itemize}
-  \item We explore the connection between SSL and functional programming using a pair of formal semantics for \Pika: A standard abstract machine semantics and a
-    semantics that translates \Pika{} into SSL specifications. The second semantics is shown to be sound relative to the first semantics. (Section~\ref{sec:semantics})
-
-  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic datatypes and their representation in memory.
-    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic datatype. (Section~\ref{sec:layouts})
-
-  \item We address the limitations of this approach.
-In particular, we develop an extension of \Pika{} that allows \SuSLik's synthesis capabilities to be exposed more naturally inside the \Pika{} language.
-(Section~\ref{sec:synth})
-
-  \item A comparison of examples that work on the first version of \Pika{} is compared with examples that are enabled by the new feature in the second version of \Pika.
-This new version of \Pika{} is not backwards compatible with the first version, so we also examine which examples in the old version no longer work in the new version.
-(Section~\ref{sec:examples})
-\end{itemize}
-
-\section{Overview}
-\label{sec:overview}
-
-\subsection{Syntax}
-\label{sec:syntax}
-
-\subsection{Layouts}
-\label{sec:layouts}
-
-\section{Formal Semantics}
-\label{sec:semantics}
-
-\section{Synthesis Limitations And Solutions}
-\label{sec:limitations}
-
-There is a significant limitation to our approach so far: \SuSLik will often fail to synthesize code from valid specifications.
-One instance of this problem is the following: In a certain class of functions, \SuSLik{} requires a different number of pointer indirections than it will with other functions.
-We have not found a way to perfectly characterize such functions and, since this will change the way the functions will need to be called, this represents a major impediment to our approach.
-It is worth noting that this is also not the \textit{only} situation in which \SuSLik{} fails to synthesize a program.
-
-From this, we draw the conclusion that \SuSLik{} is not well-suited to this task.
-In fact, this task is very different from the original goals of \SuSLik, which could be stated: \textit{Given a potentially ambiguous specification, generate any program that satisfies this specification.} However, in our case, we already know which specific program we want to generate.
-This is fully specified by the abstract machine semantics for \Pika.
-
-With this in mind, we take a new approach.
-What if we compile \Pika{} directly to C, but we add a \textit{new} feature that allows us to express synthesis problems at the level of \Pika{} code?
-This takes the form of the \verb|synth| keyword, which we explain in detail in Section~\ref{sec:synth}.
-In this case, we can get the advantages of a traditional compiler while also exploiting the synthesis capabilities of \SuSLik.
-Furthermore, due to \Pika's design, we already have easy interoperability with \SuSLik.
-Also, recall that \SuSLik's target language, \SuSLang, can be straightforwardly translated directly to C.
-This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.
-
-\section{\synth{} Keyword: Synthesis Inside \Pika}
-\label{sec:synth}
-
-Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature.
-This special kind of type signature allows us to directly access \SuSLik's pure propositions.
-These type signatures are called \synth-signatures and the syntax for invoking the \Pika{} synthesis on a \synth-signature is:
-\[
-  \synth\; f : \sigma
-\]
-\noindent
-where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is given in Figure~\ref{fig:synth-sig-syntax}.
-
-\begin{figure}
-\[
-\begin{aligned}
-  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
-  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
-    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
-  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
-  \textit{Ghost variables}\;\; G &::= @x
-\end{aligned}
-\]
-  \caption{\synth-signature syntax}
-  \label{fig:synth-sig-syntax}
-\end{figure}
-
-\noindent
-We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
-
-This extension to layouts is best explain by way of example. First, consider the original definition of the singly-linked list type:
-
-\begin{lstlisting}
-Sll : layout(List)
-Sll Nil := emp;
-Sll (Cons head tail) :=
-  x :-> head ** (x+1) :-> nxt ** Sll tail [nxt];
-\end{lstlisting}
-
-\noindent
-This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification.
-A common pattern in \SuSLik{} specifications is to use a \verb|set| type ghost variable to keep track of the elements in a data structure.
-Then, we can require that the elements are preserved or that new elements be added when using this in specifications.
-The feature we add to layouts allows us to use this feature directly:
-
-\begin{lstlisting}
-Sll : @(s : set) layout(List);
-Sll Nil := s == {} ;; emp;
-Sll (Cons head tail) :=
-  s == {head} ++ s1
-    ;;
-  x :-> head ** (x+1) :-> nxt ** Sll tail @s1 [nxt];
-\end{lstlisting}
-
-\noindent
-Putting this all together, we can use this new kind of parameter together with the \synth{} keyword to specify a list append function:
-
-\begin{lstlisting}
-synth append :
-  (s == (s1 ++ s2))
-    ;;
-  Sll @s1 -> Sll @s2 -> Sll @s
-\end{lstlisting}
-
-\section{Examples}
-\label{sec:examples}
-
-\subsection{Version 1}
-
-\subsection{Version 2}
-
-\section{Future Work}
-\label{sec:future-work}
-
-\section{Conclusion}
-\label{sec:conclusion}
+\input{introduction}
+\input{overview}
+\input{syntax}
+\input{layouts}
+\input{semantics}
+\input{limitations}
+\input{synth}
+\input{examples}
+\input{future-work}
+\input{conclusion}
 
@@
 %%
-%% End of file `sample-acmsmall.tex'.

### 2023-08-21 new-paper/semantics.tex
Expand future work with some existing material; separate out into multiple .tex files

--- 2023-08-03 outline/outline.tex
@@
-\documentclass[runningheads]{llncs}
-
-\usepackage{graphicx}
-\usepackage{proof}
-\usepackage{stmaryrd}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{xcolor}
-\usepackage{tcolorbox}
-\usepackage{tikz}
-\usepackage{tikz-cd}
-\usepackage{syntax}
-\usepackage{listings}
-
-\lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}
-
-\input{defs.tex}
-
-\begin{document}
-
-\title{Reflections on Pika}
-\maketitle
-
-\section{Introduction}
-
-There is a tool called \SuSLik{} that will take separation logic specification and generate
-a program in a C-like language. While working with this tool, it became clear that there is a strong
-connection between the form of separation logic used by \SuSLik{} and functional programming languages such as Haskell and Standard ML.
-
-In particular, we made the following observation. The SuSLik specification language has \textit{inductive predicates}. These are
-allow the programmer to define the heap layout of various inductively defined data structures, such as singly-linked lists and binary trees.
-However, we noticed that inductive predicates could also encode function definitions!
-
-Using this fact, functions written in a functional programming language can be translated into inductive predicates. The translation
-of a function definition into an inductive predicate is very similar to translation into ANF. However, this also gives us easy
-interoperability with the rest of the SuSLik specification language. For example, not only can we define data structures as inductive
-predicates, we can also dictate the heap layout of our data structures.
-
-Finally, this enables us to implement a program synthesis feature at the level of the functional programming language.
-
-\section{Examples}
-
-\section{Type System}
-
-\section{Intermediate Representation}
-
-\section{Translation to SSL}
-
-\section{Translation to C}
-
-\section{\texttt{synth} keyword}
-
 \section{Formal Semantics}
-
-% \input{introduction.tex}
-% \input{language.tex}
-% \input{types.tex}
-% \input{elaboration.tex}
-% \input{semantics.tex}
-% \input{codegen.tex}
-% \input{related-work.tex}
-
-\bibliographystyle{splncs04}
-\bibliography{outline}
-
-\end{document}
+\label{sec:semantics}

### 2023-08-21 new-paper/synth.tex
Expand future work with some existing material; separate out into multiple .tex files

\section{\synth{} Keyword: Synthesis Inside \Pika}
\label{sec:synth}

Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature. This special kind of type signature allows us to directly access \SuSLik's pure propositions. These type signatures are called \synth-signatures and the syntax for invoking the \Pika{} synthesis on a \synth-signature is:
\[
  \synth\; f : \sigma
\]
\noindent
where $\sigma$ is a \synth-signature. The syntax of a \synth-signature is given in Figure~\ref{fig:synth-sig-syntax}.

\begin{figure}
\[
\begin{aligned}
  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
  \textit{Ghost variables}\;\; G &::= @x
\end{aligned}
\]
  \caption{\synth-signature syntax}
  \label{fig:synth-sig-syntax}
\end{figure}

\noindent
We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.

This extension to layouts is best explain by way of example. First, consider the original definition of the singly-linked list type:

\begin{lstlisting}
Sll : layout[x](List)
Sll Nil := emp;
Sll (Cons head tail) :=
  x :-> head ** (x+1) :-> nxt
    ** Sll tail [nxt];
\end{lstlisting}

\noindent
This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification. A common pattern in \SuSLik{} specifications is to use a \verb|set| type ghost variable to keep track of the elements in a data structure. Then, we can require that the elements are preserved or that new elements be added when using this in specifications. The feature we add to layouts allows us to use this feature directly:

\begin{lstlisting}
Sll : @(s : set) layout[x](List);
Sll Nil := s == {} ;; emp;
Sll (Cons head tail) :=
  s == {head} ++ s1
    ;;
  x :-> head ** (x+1) :-> nxt
    ** Sll tail @s1 [nxt];
\end{lstlisting}

\noindent
Putting this all together, we can use this new kind of parameter together with the \synth{} keyword to specify a list append function:

\begin{lstlisting}
synth append :
  s == s1 ++ s2
    ;;
  Sll @s1 -> Sll @s2 -> Sll @s;
\end{lstlisting}

\noindent
Using this feature, we can use \SuSLik{} specifically for \textit{synthesis} tasks inside \Pika. That is, we are able to directly use \SuSLik{} for its intended purpose.

### 2023-08-22 new-paper/overview.tex
Bring more things over

\section{Overview}
\label{sec:overview}

\subsection{Background}
The \Pika{} language is translated into
\SuSLik~\cite{polikarpova:2019:suslik}, which is a program synthesis
tool that uses Synthetic Separation Logic (SSL)-a variant of
Hoare-style~\cite{hoare:1969:axiomatic} Separation Logic
(SL)~\cite{ohearn:2001:seplogic}.

% SSL is a variant of separation
% logic that works particularly well for program
% synthesis.~\cite{polikarpova:2019:suslik}

A synthesis task specification in \SuSLik{} is given as a function
signature together with a pair of pre- and post-conditions, which are
both SL assertions~\cite{polikarpova:2019:suslik}.
%
The synthesiser generates code that satisfies the given specification,
along with the SL \emph{proof} of its correctnees by performing the
search in a space of proofs that can be derived by using the rules of
the underlying logic~\cite{WatanabeGPPS21}.
%
A distinguishing feature of \textit{Synthetic} Separation Logic is the
format of its assertions. An SSL assertion consists of two parts: a
\textit{pure part} and a \textit{spatial part}. The pure part is a
Boolean expression constraining the variables of the specification
using a few basic relations, like equality and the less-than relation
for integers. The spatial part is a \textit{symbolic heap}. This
consists of a list of heaplets. The list separator for the list of
heaplets is the $\sep$ symbol. Each heaplet takes on one of the
following forms~\cite{polikarpova:2019:suslik}:

\begin{itemize}
  \item \texttt{emp}: This represents the empty heap. It is also the left and right identity for $\sep$.
  \item $\ell \mapsto a$: This asserts that memory location $\ell$ points to the value $a$.
  \item $[\ell, \iota]$: At memory location $\ell$, there is a block of size $\iota$.
  \item $p(\overline{\phi})$: This is an application of the
    \textbf{\textit{inductive predicate}} $p$ to the arguments
    $\overline{\phi}$. An inductive predicate has a collection of
    branches, guarded by Boolean expressions (conditions) on the
    parameters. The body of each branch is an SSL assertion. The
    assertion associated with the first branch condition that is
    satisfied is used in place of the application. Note that inductive
    predicates can be (and often are) recursively defined.
\end{itemize}

\noindent
The general form of an SSL assertion is
$(p; h_1 \sep h_2 \sep \cdots \sep h_n)$, where $p$ is the pure part
and $h_1, h_2, \cdots, h_n$ are the heaplets which are the conjuncts
that make up the separating conjunction of the spatial part. A syntax
definition for SSL is given in \autoref{fig:SSL-syntax}, which is
adapted from \textit{Cyclic Program Synthesis} by Itzhaky
\etal~\cite{itzhaky:2021:cyclic-synth}.

\begin{figure}[t]
\setlength{\abovecaptionskip}{5pt}
%\setlength{\belowcaptionskip}{-15pt}
{\small{
\[
  \begin{array}[t]{ l l }
    \text{Variable} &x, y\;\; \text{Alpha-numeric identifiers $\in$ \textnormal{Var}}\\
    \text{Size, offset} &n, \iota\;\; \text{Non-negative integers}\\
    \text{Expression} &e ::= 0\; \mid \texttt{true} \mid x \mid e = e \mid e \land e \mid \neg e \mid d\\
    \text{$\mathcal{T}$-expr.} &d ::= n \mid x \mid d + d \mid n \cdot d \mid \{\} \mid {d} \mid \cdots\\
    \text{Command} &c ::= \begin{aligned}[t] &\texttt{let x = *(x + }\iota\texttt{)} \mid \texttt{*(x + }\iota\texttt{) = e} \mid
      \\
      &\texttt{let x = malloc(n)} \mid \texttt{free(x)} \mid \texttt{error}\newline
      \mid \texttt{f(}\overline{e_i}\texttt{)}
      \end{aligned}\\
    \text{Program} &\Pi ::= \overline{f(\overline{x_i})\; \{\; c\; \}\; ;}\; c\\
    \text{Logical variable} &\nu, \omega\\
    \text{Cardinality variable} &\alpha\\
    \text{$\mathcal{T}$-term} &\kappa ::= \nu \mid e \mid \cdots\\
    \text{Pure logic term} &\phi, \psi, \chi ::= \kappa \mid \phi = \phi \mid \phi \land \phi \mid \neg \phi\\
    \text{Symbolic heap} &P, Q, R ::= \texttt{emp} \mid \mbox{$\langle e, \iota \rangle \mapsto e \mid [e, \iota]$} \mid p^{\alpha}(\overline{\phi_i})
      \mid \mbox{$P * Q$}\\
    \text{Heap predicate} &\mathcal{D} ::= p^{\alpha}(\overline{x_i}) : \overline{e_j \Ra \exists \overline{y}.\{\chi_j;R_j\}}\\
    \text{Assertion} &\mathcal{P},\mathcal{Q} ::= \{\phi; P\}\\
    \text{Environment} &\Gamma ::= \forall\overline{x_i}.\exists\overline{y_j}.\\
    \text{Context} &\Sigma ::= \overline{\mathcal{D}}\\
    \text{Synthesis goal} &\mathcal{G} ::= P \leadsto Q
  \end{array}
\]
}}
\caption{Syntax of Synthetic Separation Logic}
  \label{fig:SSL-syntax}
\end{figure}

\begin{figure}
  \ruleSAdd
  \caption{{\sc S-Add} rule}
  \label{fig:ruleSAdd}
\end{figure}

As the specification language of \SuSLik, SSL serves as the
compilation target for the \Pika{} language. From there, executable
programs are generated through \SuSLik's program synthesis. Consider a
program that takes an integer $x$ and a result in location $r$ and
stores $x+1$ at location $r$. This can be written as the \SuSLik{}
specification:

\begin{lstlisting}
void add1Proc(int x, loc r)
  { r :-> 0 }
  { y == x+1 ; r :-> y }
{ ?? }
\end{lstlisting}

\noindent
In contrast, this example can be written as follows in \Pika:

\begin{lstlisting}
add1Proc : Int -> Int;
add1Proc x := x + 1;
\end{lstlisting}

\noindent
In constrast with \SuSLik{} specs, the one below features no direct
manipulation with pointers.
%
This program is translated using the {\sc S-Add} rule shown in
\autoref{fig:ruleSAdd}.

\subsection{The \Pika{} Language}
\label{sec:language}

While SSL provides a specification language that allows tools like
\SuSLik{} to synthesise code, it is only able to express specifications
as pointers. This is useful for some applications, such as embedded
systems, but it does not provide any high-level abstractions. As a
result, every part of a specification is tailored to a specific memory
representation of each data structure involved.
%
To addresss this shortcoming, we introduce a language with algebraic
data types that gets translated into SSL specifications. Additionally,
we introduce a language construct that allows the programmer to
specify a memory representation of an algebraic data type. This is
called a \textit{layout}. This distinction between algebraic data
types and layouts allows the separation of concerns between the low
level representation of a data structure and code that manipulates it
at a high level.

Syntactically, \Pika{} resembles a functional programming language
of the Miranda~\cite{turner:1986:miranda} and
Haskell~\cite{hudak:2007:haskell} lineage. It supports algebraic data
types, pattern matching at top-level function definitions (though not
inside expressions) and Boolean guards. The primary difference arises
due to the existence of layouts and the fact that the language is
compiled to an SSL specification rather than executable code.

Functions in \Pika{} are only defined by their operations on algebraic
data types. Thus, all function definitions are ``layout polymorphic''
over the particular choices of layouts for their arguments and result.
Giving a layout-polymorphic function, a particular choice of layouts
is called ``instantiation''. Specifying the layout of a non-function
value is called ``lowering.''

There is a subtlety here: The layout of a subexpression inside of a
function definition might not be determined by the layouts that the
function definition is instantiated at. In this case, we must
explicitly specify a layout. This is done using the \verb|instantiate|
and \verb|lower| constructs. The former is for function calls, while
the latter is for all other types of values. These operations are
performed with the \verb|instantiate| and \verb|lower| language
constructs, respectively. For example, \verb|instantiate [A] B f x|
calls $f$ with the argument $x$, where the layout $A$ is used for the
argument and the layout $B$ is used for the result of the function
application. This is also written $\instantiateS{A,B}{f}(x)$. The
latter is primarily used in \autoref{sec:formal}.

The code generator is instructed to generate a \SuSLik{} specification
for a certain function at a certain instantiation by using a
\generate{} directive. For example, if there is a function definition
with the type signature \verb|mapAdd1 : List -> List|, a line reading
\verb|%generate mapAdd1 [Sll] Sll| would instruct the
\Pika{} compiler to generate the \SuSLik{} inductive predicate
corresponding to \verb|mapAdd1| instantiated to the \verb|Sll| layout
for both its argument and its result.

An example of an ADT definition and a corresponding layout definition
is given in \autoref{fig:List-def}. There is one unusual part of
the syntax in particular which requires further explanation: layout
type signatures. A layout definition consists a \textit{layout type
  signature} and a pattern match (much like a function definition),
with lists of SSL heaplets on the right-hand sides. A layout type
signature has a special form $A : \alpha \monic \layout[x]$. This says
that the layout $A$ is for the algebraic data type $\alpha$ and the
SSL output variable is named $x$. The full \Pika{} grammar is given in
the Appendices. %x~\ref{sec:grammar}.

\begin{figure}[t]
\begin{lstlisting}
data List := Nil | Cons Int List;

Sll : List >-> layout[x];
Sll (Nil) := emp;
Sll (Cons head tail) := x :-> head, (x+1) :-> tail, Sll tail;
\end{lstlisting}
  \caption{\lstinline{List} algebraic data type together with its singly-linked list layout \lstinline{Sll}}
  \label{fig:List-def}
\end{figure}

\subsection{\Pika{} by Example}

We demonstrate the characteristic usages of \Pika{} on a series of
examples.
%
In these examples, we will often make use of the \verb|List| algebraic
data type and its \verb|Sll| layout from~\autoref{fig:List-def}. A
simple example of \Pika{} code that illustrates algebraic data types and
layouts is a function which creates a singleton list out of the given
integer argument:

\begin{lstlisting}
%generate singleton [Int] Sll

singleton : Int -> List;
singleton x := Cons x (Nil);
\end{lstlisting}

\noindent
This gets compiled to the following \SuSLik{} specification, with generated names simplified for greater clarity:

\begin{lstlisting}
predicate singleton(int p, loc r) {
| true => { r :-> p ** (r+1) :-> 0 ** [r,2] }
}
\end{lstlisting}

\noindent
A slightly more complicated example comes from trying to write a
functional-style \verb|map| function directly in \SuSLik. Consider a
function which adds 1 to each integer in a list of integers.
Considering the list implementation to be a singly-linked list with a
fixed layout, one way to express this in \SuSLik{} is shown in \autoref{fig:sllx}

\begin{figure}[t]
\begin{lstlisting}
predicate sll(loc x) {
| x == 0 => { emp }
| not (x == 0) => { [x, 2] ** x :-> v ** (x+1) :-> nxt ** sll(nxt) }
}

predicate mapAdd1(loc x, loc r) {
| x == 0 => { emp }
| not (x == 0) => {
    [x, 2] ** x :-> v ** (x+1) :-> xNxt **
    [r, 2] ** r :-> (v+1) ** (r+1) :-> rNxt **
    mapAdd1(xNxt, rNxt)
  }
}

void mapAdd1_fn(loc x, loc y)
  { sll(x) ** y :-> 0 }
  { y :-> r ** mapAdd1(x, r) }
{ ?? }
\end{lstlisting}
\caption{Specifying a function that adds one to each element of a
  singly-linked list in \SuSLik.}
\label{fig:sllx}
\end{figure}

\noindent
Note that inductive predicates are used for \textit{two} different purposes: the \verb|sll| inductive
predicate describes a singly-linked list data structure, while the \verb|mapAdd1| inductive predicate 
describes how the input list relates to the output list. Both are used in the specification of
\verb|mapAdd1_fn|: \verb|sll| in the precondition and \verb|mapAdd1| in the postcondition.

Using the \verb|mapAdd1| inductive predicate gives us two advantages over attempting to
put the SSL propositions directly into the postcondition of \verb|mapAdd1_fn|:
%
\begin{enumerate}
  \item We are able to express a conditional on the shape of the list. This is much like pattern
    matching in a language with algebraic data types, but we are examining the pointer involved directly.
  \item We are able to express \textit{recursion} part of the postcondition: the \verb|mapAdd1| inductive
    predicate refers to itself in the \verb|not (x == 0)| branch.
\end{enumerate}
%
These two features are both reminiscent of features common in
functional programming languages: pattern matching and recursion.
However, there are still some significant differences:

\begin{itemize}
  \item In traditional pattern matching, the underlying memory representation of the data structure is not exposed.
  \item Compared to a functional programming language, the meaning of the specification is more obscured. It is
    necessary to think about the structure of the linked data structure to determine what the specification is saying. This
    is related to the first point: The memory representation is front-and-center.
  \item In many functional languages, mutation is either restricted or generally discouraged. In \SuSLik, mutation is commonplace.
\end{itemize}

Say we want to write the functional program that corresponds to this specification. One way to do this in a Haskell-like language
is the following, using the \verb|List| type from \autoref{fig:List-def}.

\begin{lstlisting}
mapAdd1_fn : List -> List;
mapAdd1_fn (Nil) := 0;
mapAdd1_fn (Cons head tail) := Cons (head + 1) (mapAdd1_fn tail);
\end{lstlisting}

\noindent
The only missing information is the memory representation of the \verb|List| data structure. We do not want the
\verb|mapAdd1_fn| implementation to deal with this directly, however. We want to separate the more abstract notions
of pattern matching and constructors from the concrete memory layout that the data structure has.

To accomplish this, we now extend the code with the definition of
\verb|Sll| from \autoref{fig:List-def}. \verb|Sll| is a
\textit{layout} for the algebraic data type \verb|List|.
%
Now we have all of the information of the original specification but
rearranged so that the low-level memory layout is separated from the
rest of the code. This separation brings us to an important
observation about the language, manifested throughout these examples:
none of the function definitions need to \emph{directly} perform any
pointer manipulations. This is relegated entirely to the reusable
layout definitions for the ADTs. The examples are written entirely as
recursive functions that pattern match on, and construct, ADTs.

All that is left is to connect these two parts: the layouts and the
function definitions. We instruct a \SuSLik{} specification generator to
generate a \SuSLik{} specification from the \verb|mapAdd1_fn| function
using the \verb|Sll| layout:
%
\begin{lstlisting}
%generate mapAdd1_fn [Sll] Sll
\end{lstlisting}
%
The \verb|[Sll]| part of the directive tells the generator which
layouts are used for the arguments. In this case, the function only
has one argument and the \verb|Sll| layout is used. The \verb|Sll| at
the end specifies the layout for the result.

\subsubsection{map}

We can generalise our \verb|mapAdd1| to map arbitrary \verb|Int|
functions over a list and then redefine \verb|mapAdd1| using the new
\verb|map|.

\begin{lstlisting}
%generate mapAdd1 [Sll] Sll

data List := Nil | Cons Int List;

Sll : List >-> layout[x];
Sll (Nil) := emp;
Sll (Cons head tail) := x :-> head, (x+1) :-> tail, Sll tail;

map : (Int -> Int) -> List -> List;
map f (Nil) := Nil;
map f (Cons x xs) := Cons (instantiate [Int] Int f x) (map f xs);

add1 : Int -> Int;
add1 x := x + 1;

mapAdd1 : List -> List;
mapAdd1 xs :=
  instantiate [Int -> Int, Sll] Sll map add1 xs;
\end{lstlisting}

This example makes use of \lstinline{instantiate} in two places. In the first case where we have the call
\lstinline{instantiate [Int] Int f x}, the builtin \verb|Int| layout is used for both the input
and output. In this special case, the \verb|Int| layout shares a name with the \verb|Int| type
that it represents. This is necessary since \lstinline{instantiate} is used for all non-recursive calls that are not constructor applications.

In the second use, \lstinline{instantiate [Int -> Int, Sll] Sll map add1 xs}, we specify that
the second argument uses the \verb|Sll| layout for the \verb|List| type from \autoref{fig:List-def}. We also give \verb|Sll| as
the layout for the result of the call. Note that it is not necessary to use \verb|instantiate| for the recursive call to \verb|map|. This is because the appropriate layout
is inferred for recursive calls.

It might be surprising that \verb|instantiate| is required in the body of \verb|mapAdd1| since the type signature
of \verb|mapAdd1| suggests that it is layout polymorphic and yet we must pick a specific \verb|List| layout when
we use \verb|instantiate| to call \verb|map|. This is because, in general, a call inside the body of some function \verb|fn| might use
any layout, even layouts that have no relation to the layouts that \verb|fn| is instantiated to.

It is possible to do inference of some layouts, for example in
\verb|mapAdd1| we would usually want to use the same layout as the
argument layout, but we leave this for future work. Another approach
is to introduce type variables that correspond to layouts, as done in
the series of works on the \tname{Dargent}{}
tool~\cite{chen:2023:dargent}.
%
We leave this approach for future work as well.

\subsubsection{Guards}

While we have a pattern-matching construct at the top level of a function definition, we
have not seen a way to branch on a Boolean value so far. This is a feature that is
readily available at the level of \SuSLik, since the same conditional construct we
use to implement pattern matching can also use other Boolean expressions.

We can expose this in the functional language using a \textit{guard},
much like Haskell's guards. Say, we want to write a specialised
filter-like function. Specifically, we want a function that filters
out all elements of a list that are less than 9. This is a specific
example where the \SuSLik{} specification is noticeably more difficult
to read. For a \SuSLik{} specification of this example, see
\autoref{fig:SuSLik-filter}

\begin{figure}
  \begin{lstlisting}
predicate filterLt9(loc x, loc r) {
| (x == 0) => { r == 0 ; emp }
| not (x == 0) && head < 9 =>
    { x :-> head ** (x+1) :-> tail ** [x,2] ** filterLt9(tail, r) }
| not (x == 0) && not (head < 9) =>
    { x :-> head ** (x+1) :-> tail ** [x,2] ** filterLt9(tail, y)
      ** r :-> head ** (r+1) :-> y ** [r,2] }
}

void filterLt9(loc x1, loc r)
  { Sll(x1) ** r :-> 0 }
  { filterLt9(x1, r0) ** r :-> r0 }
{ ?? }
  \end{lstlisting}
  \caption{\SuSLik{} specification of \lstinline{filterLt9}, excluding \lstinline{Sll} which is given in \autoref{fig:List-def}}
  \label{fig:suslik-filter}
\end{figure}

On the other hand, an implementation of this in \Pika{} is:

\begin{lstlisting}
%generate filterLt9 [Sll] Sll

filterLt9 : List -> List;
filterLt9 (Nil) := Nil;
filterLt9 (Cons head tail)
  | head < 9       := filterLt9 tail;
  | not (head < 9) := Cons head (filterLt9 tail);
\end{lstlisting}

\noindent
When translating a guarded function body, the translator takes the conjunction of the Boolean guard condition with
the condition for the pattern match.

\subsubsection{\texttt{let} bindings and \texttt{if-then-else}}

Two more features that are common in functional languages are \verb|let| bindings and
\verb|if-then-else| expressions. Both of these have straightforward translations into
\SuSLik. A \verb|let| binding corresponds to introducing a new variable with an equality
constraint corresponding to the equality in the \verb|let| binding in the pure part of a \SuSLik{} assertion.

The \verb|if-then-else| construct corresponds to \SuSLik's C-like ternary operator. We can use these features together to implement a \verb|maximum| function:

\begin{lstlisting}
%generate maximum [Sll] Int

maximum : List -> Int;
maximum (Nil) := 0;
maximum (Cons x xs) :=
  let i := maximum xs
  in
  if i < x
    then x
    else i;
\end{lstlisting}

\noindent
The \verb|let| binding is required here. This is because the recursive predicate
application must use an additional variable, beyond the argument and output
parameters, to store the intermediate result \verb|i| and this is introduced in
\Pika{} using a \verb|let|. It is possible to extend the translator to
automatically introduce these \verb|let| bindings, but we relegate this to future work.

Furthermore, since the \verb|let| is required and the variable introduced by the \verb|let| is used
in the \verb|if| condition, we cannot replace the \verb|if| with a guard. Guards can only occur
at the top-level of a function definition, so it could not occur after a \verb|let|.

\subsubsection{Using multiple layouts}

To show the interaction between multiple algebraic data types, we write a
function that follows the left branches of a binary tree and collects the values
stored in those nodes into a list. This example demonstrates a binary tree
algebraic data type and a layout that corresponds to it.

\begin{lstlisting}
%generate leftList [TreeLayout] Sll

data Tree := Leaf | Node Int Tree Tree;

TreeLayout : Tree >-> layout[x];
TreeLayout (Leaf) := emp;
TreeLayout (Node payload left right) :=
  x :-> payload,
  (x+1) :-> left,
  (x+2) :-> right,
  TreeLayout left,
  TreeLayout right;

leftList : Tree -> List;
leftList (Leaf) := Nil;
leftList (Node a b c) := Cons a (leftList b);
\end{lstlisting}

\subsubsection{fold}
\label{sec:examples-fold}

A fold is a common kind of operation on a data structure in functional programming, where a binary function
is applied the elements of a data structure to create a summary value. For example, if the binary function
is the addition function, it will give the sum of all the elements of the data structure. The
classic example of such a fold is a fold on a list. In this example, we will write a
right fold over a \verb|List|.

We also demonstrate \verb|Ptr| types and the \verb|addr| builtin operation. This is somewhat similar to an \verb|@| pattern in Haskell,
but it is a low-level construct, indictating how pointers should be used, rather than a high-level construct.
Given a base type $\alpha$ (such as \verb|Int|), \verb|Ptr Int| is a type. This type is
passed around differently in the generated code. In particular, it is passed by reference.
\verb|addr| is the corresponding value-level introduction operation.
\begin{lstlisting}
%generate fold_List [Int, Sll] (Ptr Int)

fold_List : Int -> List -> Ptr Int;
fold_List z (Nil) := z;
fold_List z (Cons x xs) :=
  instantiate
    [Ptr Int, Ptr Int]
    (Ptr Int)
    f
    (addr x)
    (fold_List z xs);
\end{lstlisting}

\noindent
The compiler produces the following \SuSLik{} specification for \verb|fold_List|:

\begin{lstlisting}
predicate fold_List(int i, loc p_x ,loc __r) {
| (p_x == 0) => { __r :-> i }
| (not (p_x == 0)) => {
  p_x :-> x ** (p_x+1) :-> xs ** [p_x,2] **
  func f(p_x, __p_2, __r) **
  fold_List(i, xs, __p_2) **
  temploc __p_2 }
}
\end{lstlisting}
%
In the generated code, \verb|f| is applied to \verb|p_x| rather than
\verb|x| as a result of using \verb|Ptr| and \verb|addr|.

### 2023-08-22 new-paper/semantics.tex
Bring more things over

@@
 \label{sec:semantics}
+
+\subsection{Formal Semantics of \tool}
+\label{sec:formal}
+
+In this section, we have the following plan:
+
+\begin{itemize}
+  \item We define abstract machine semantics for executing a subset of \tool programs. This semantics
+    is given by the big-step relation $\step$ which we define later.
+  \item We define the translation from that subset of \tool into SSL. This translation is given by the
+    function $\Tsem{e}{V,r}$ from \tool expressions into SSL propositions. The $r$ is a variable name to
+    be used in the resulting proposition and $V$ is a collection of fresh names.
+  \item We prove a soundness theorem. Given any well-typed expression $e$ and an abstract machine reduction producing
+    the store-heap pair $\EndPair$, the SSL translation of $e$ should be satisfied by SSL model $\EndPair$. This is stated formally, and proven,
+    in \autoref{thm:gen-soundness}.
+  % \item We prove a soundness theorem. In particular, we first look at the heap-store pair that results from executing the abstract
+  % machine on an arbitrary function application expression \verb|f e| (omitting the \verb|instantiate| for brevity).
+  % We then show that the heap-store pair we got from the abstract machine is a model for the 
+  % SSL proposition given by translating \verb|f e| into SSL.
+\end{itemize}
+
+This subset of \tool does not have guards or \verb|if then else| expressions, but it does have pattern
+matching. It also has the requirement that functions can only have one argument. Unlike the implementation, there
+is no elaboration. As a result, every algebraic data type value must be lowered to a specific layout at every usage
+and every function application must be explicitly instantiated with a layout for the argument and a layout for the result. We
+also limit the available integer and Boolean operations for brevity.
+
+The grammar for this subset is given in \autoref{fig:subset-grammar}. The grammar for types, layout definitions and algebraic data type definitions remain the same
+as before and are therefore omitted from the following grammar.
+
+\begin{figure}
+\begin{grammar}
+  <i> ::= $\cdots$ $\mid$ -2 $\mid$ -1 $\mid$ 0 $\mid$ 1 $\mid$ 2 $\mid$ $\cdots$
+
+  <b> ::= \texttt{true} $\mid$ \texttt{false}
+
+  <e> ::= <var> \alt <i> \alt <b> \alt <e> + <e> \alt C $\overline{\langle e \rangle}$ \alt $\instantiateS{A,B}{f}(\langle \textit{e} \rangle)$ \alt $\lowerS{A}{\langle \textit{e} \rangle}$
+
+  <fn-def> ::= $\overline{\langle \textit{fn-case} \rangle}$
+
+  <fn-case> ::= f <pattern> := <e>
+\end{grammar}
+  \caption{Grammar for restricted \tool subset}
+  \label{fig:subset-grammar}
+\end{figure}
+
+\noindent
+The semantics for SSL are largely derived~\cite{polikarpova:2019:suslik} from standard separation logic semantics.~\cite{rowe:2017:auto-cyclic-term}
+
+\subsubsection{Overview of the Two Interpretations}
+
+The soundness theorem will link the abstract machine semantics to the translation. In fact, the abstract machine
+semantics and the translation are similar to each other. For the abstract machine we manipulate \textit{concrete} heaps, while
+for the translation we generate \textit{symbolic} heaps.
+
+Comparing the two further, there are two main points (beyond what we've already mentioned) where these two interpretations of \tool differ:
+
+% \tbox{
+    % \begin{center} \textbf{Two main differences between interpretations} \end{center}
+
+      \begin{enumerate}
+        \item When we need to unfold a layout, how do we know which layout branch to choose?
+        \item How do we translate function applications (including, but not limited to, recursive applications)?
+      \end{enumerate}
+  % }
+% \\
+
+% \subsubsection{Abstract Machine}
+First, consider the abstract machine semantics. In this case, we are able to choose which branch of a layout to use by evaluating the expression we are applying it
+          to until the expression is reduced to a constructor value (where a ``constructor value'' is either a value or a constructor applied to constructor values).
+If the expression is well-typed, this will always be a constructor
+          of the algebraic data type corresponding to the layout. The two rules that this applies to are {\sc AM-Lower} and
+          {\sc AM-Instantiate}.
+        To interpret a function application, we interpret its arguments and substitute the results into the body
+          of the function. We then proceed to interpret the substituted function body. This process is performed by the {\sc AM-Instantiate} rule.
+  % }
+% \\
+
+% \subsubsection{SSL Translation}
+Next, consider the SSL translation. Here we can determine which layout branch to use by generating a Boolean condition that will be
+          true if and only if the SSL proposition on the right-hand side of the branch holds for the heap. Note
+          that we assume that the programmer-supplied layout definitions are \textit{injective} functions from
+          algebraic data types to SSL assertions (up to bi-implication).
+We can directly translate function applications into SSL inductive predicate applications. Since inductive predicates
+          already allow for recursive applications, there is no special handling necessary for recursion.
+      % \end{enumerate}
+  % }
+% \\
+
+% We will return to these points when we talk about each of the two interpretations.
+%
+% The differences on the first point, again, results primarily from the difference between concrete heaps and symbolic heaps.
+%
+% For the second
+% point, on the other hand, there is an additional reason for the differences. Both \tool and SSL have a notion of application. In the former,
+% this is function application and in the latter it is predicate application. In both cases, these applications can be recursive. Since
+% \tool functions are translated into inductive predicates, function applications are translated into predicate applications. As SSL has
+% recursion, we do not need to handle this specially.
+%
+% For the abstract machine, in contrast, we must actually \textit{run} the body of the function.
+
+After defining these interpretations, we show how the abstract machine relation $\step$ and the SSL interpretation function $\Tsem{e}{V,r}$ relate to each other by the Soundness \autoref{thm:gen-soundness}.
+
+% We sketch the
+% statement of the theorem here, leaving the details for later. In particular, the part marked with question marks is filled in later. We
+% show the {\color{cyan} start state} in {\color{cyan} cyan} and the {\color{orange} end state} in {\color{orange} orange}.
+%
+%
+%   \[
+%     \begin{aligned}
+%     &\begin{aligned}
+%       (\fbox{      ???      }
+%       &\land (\exists \VarSet, r.\; ({\color{cyan} \sigma}, {\color{cyan} h}) \models \Tsem{e}{\VarSet, r})\\
+%       &\land (f_{A,B}(e), {\color{cyan} \sigma}, {\color{cyan} h}) \step (e_2, {\color{orange} \sigma'}, {\color{orange} h'}, v'))
+%      \end{aligned}
+%     \\
+%         & \implies
+%       (\exists \VarSet, r.\; ({\color{orange} \sigma'}, {\color{orange} h'}) \models \Tsem{f_{A,B}(e)}{\VarSet, r})
+%     \end{aligned}
+%   \]
+%
+% At a high-level, the theorem says: if we start in a machine state which is a model for the SSL interpretation of a given expression $e$ (this SSL proposition is given by $\Tsem{e}{\VarSet,r}$), then
+% any state that the abstract machine steps to from the original state, given an application expression $f_{A,B}(e)$, is a model for the SSL interpretation
+% $\Tsem{f_{A,B}(e)}{\VarSet,r}$. This is subject to the condition that there exists \textit{some} model for $\Tsem{f_{A,B}(e)}{\VarSet,r}$. The reason for this requirement
+% is that the SSL interpretation of an arbitrary \tool expression is not guaranteed to be satisfiable. There are further restrictions
+% that will be given when we fill in the question marks, but this is deferred to a later subsection.
+
+\subsubsection{Abstract Machine Semantics}
+In this section, we will define an abstract machine semantics for \tool and relate this to
+the standard semantics for SSL.
+
+% \subsubsection{Notation and Setup}
+
+% The only difference between the two sorts of values is that the set of \tool values also includes
+% constructor application values, while \BaseVal{} does not.
+
+The set of values is $\Val = \mathbb{Z} \cup \mathbb{B} \cup \Loc$. Each of these three sets
+is disjoint from the other two. In particular, note that $\Loc$ and $\mathbb{Z}$ are disjoint.
+
+\begin{figure}
+\[
+\begin{array}{c}
+  \labinfer[\FsVal-Base]{x \in \FsVal}
+    {x \in \Val}
+  ~~~
+  \labinfer[\FsVal-Constr]{(C\; x_1 \cdots x_n) \in \FsVal}
+    {x_1 \in \FsVal & \cdots & x_n \in \FsVal}
+\end{array}
+\]
+  \caption{\FsVal{} judgment rules}
+  \label{fig:FsVal-rules}
+\end{figure}
+
+There is also a set of \tool values \FsVal. This includes all the elements of \Val{}, but also includes
+``constructor values'' given by the rules in \autoref{fig:FsVal-rules}.
+In addition to the store and heap of standard SSL semantics, the abstract machine semantics uses an \FsStore. This is a partial function from locations
+to \tool values: $\FsStore = \Loc \rightharpoonup \FsVal$. The primary purpose of this is to recover constructor values when given a location.
+
+%\noindent
+The general format of the transition relation is $(e, \sigma, h, \mathcal{F}) \step (v, \sigma', h', \mathcal{F}', r)$, where the expression $e$ results
+in the store being updated from $\sigma$ to $\sigma'$, the heap being updated from $h$ to $h'$, $v$ is the \Val{} obtained by
+evaluating $e$, the initial and final $\FsStore$ are $\mathcal{F}$ and $\mathcal{F}'$ and the result is stored in variable $r$.
+We assume that there is a global environment $\Sigma$ which contains all layout definition equations and function definition equations.
+% We will have the following properties for the \FsStore:
+%
+% % \begin{lemma}
+% %   $(e, \sigma, h, \mathcal{F}) \step (v, \sigma', h', \mathcal{F}', r) \implies (dom(h) = dom(\mathcal{F}) \land dom(h') = dom(\mathcal{F}'))$
+% % \end{lemma}
+%
+% \begin{lemma}
+%   $(e, \sigma, h, \mathcal{F}) \step (v, \sigma', h', \mathcal{F}', r) \implies (v \in dom(h') \land v \in \Loc \land h'(v) \in \Loc \implies h'(v) \in dom(\mathcal{F}'))$
+% \end{lemma}
+
+% Describing this relation more explicitly, we have $\step\; \subset (\Expr \times \Store \times \Heap \times \FsStore) \times (\Val \times \Store \times \Heap \times \times \FsStore \times \Var)$.
+
+% \noindent
+% The syntax $dom(\sigma)$ allows us to refer to the set of names in a store. It is defined as
+% \[
+%   dom(\sigma) = \{ v \mid (v, a) \in \sigma \}
+% \]
+
+Given a heap $h$ and a layout body $H$, we will make use of the
+notation $h \cdot [H]$. This extends the heap with the location
+assignments given in $H$. We say that the layout body $H$ is
+\textit{acting} on the heap $h$. This is defined in
+\autoref{fig:layout-act}. The intuition for this is that $h$ gets
+updated using the symbolic heap description in $H$. For example, if
+$h$ is the empty heap and $H$ is $a \pointsto 7$, then the updated
+heap $h \cdot [H]$ will contain only the value $7$ at the location
+$a$. It is assumed that $H$ does not have any variables on the
+right-hand side of $\pointsto$.
+
+\begin{figure}
+\[
+\begin{array}{c}
+  \labinfer[L-Emp]{h \cdot [\emp] = h}
+    {}
+  ~~~
+  \labinfer[L-PointsTo]{h \cdot [\ell \pointsto a, \kappa] = h'[\ell \mapsto a]}
+    {h' = h \cdot [\kappa] & a \in \Val}
+  \\\\
+  \labinfer[L-Apply]{h \cdot [A[x](e), \kappa] = h \cdot [\kappa]}
+    {e \in \Val}
+\end{array}
+\]
+  \caption{Rules for layout bodies acting on heaps}
+  \label{fig:layout-act}
+\end{figure}
+
+% \subsubsection{Abstract Machine Rules}
+% The rules for the abstract machine transition relation are as follows.
+
+The abstract machine semantics provides big-step operational semantics
+for evaluating \tool expressions on a heap machine. Its rules, given by
+\autoref{fig:abs-machine}, make use of standard SSL models:
+
+\begin{align*}
+  &\text{Model} &\mathcal{M} ::= (\sigma, h)\\
+  &\text{Store} &\sigma : \text{Var} \rightharpoonup \text{Val}\\
+  &\text{Heap} &h : \text{Loc} \rightharpoonup \text{Val}\\
+\end{align*}
+
+\noindent
+Note that a compound expression, consisting of multiple
+subexpressions, uses \emph{disjoint} parts of the heap for each
+subexpression. This can be seen in the {\sc AM-Add}, {\sc AM-Lower}
+and {\sc AM-Instantiate} rules, which is essential for the proof of
+Soundness \autoref{thm:gen-soundness}.
+
+\begin{figure}
+{
+\[
+\begin{array}{c}
+  \labinfer[AM-Int]{(i, \sigma, \emptyheap, \mathcal{F}) \step (i, \sigma', \emptyheap, \mathcal{F}, r) }
+    {\freshVar{r} & i \in \mathbb{Z} & \sigma' = \sigma \cup \{ (r, i) \}}
+  ~~~
+  \labinfer[AM-Bool]{(b, \sigma, \emptyheap, \mathcal{F}) \step (b, \sigma', \emptyheap, \mathcal{F}, r) }
+    {\freshVar{r} & b \in \mathbb{B} & \sigma' = \sigma \cup \{ (r, b) \}}
+  \\\\
+  \labinfer[AM-Var-Base]{(v, \sigma, \emptyheap, \mathcal{F}) \step (\sigma(v), \sigma, \emptyheap, \mathcal{F}, v)}
+    {v \in dom(\sigma) & \sigma(v) \not\in \Loc}
+  \\\\
+  \labinfer[AM-Var-Loc]{(v, \sigma, \emptyheap, \mathcal{F}) \step (\mathcal{F}(\sigma(v)), \sigma, \emptyheap, \mathcal{F}, v)}
+    {v \in dom(\sigma) & \sigma(v) \in \Loc}
+  \\\\
+  % \labinfer[AM-Constr]{(C\; e_1 \cdots e_n, \sigma, h, \mathcal{F}) \step (C\; e_1 \cdots e_n, \sigma, h, \mathcal{F})}
+  %   {{\color{red} TODO} \textnormal{fix this}}
+  % \\\\
+  \labinfer[AM-Add]{(x + y, \sigma, h, \mathcal{F}) \step (z, \sigma', h, \mathcal{F}, r)}
+    {\begin{gathered}
+      (x, \sigma, \mathcal{F}, h_1) \step (x', \sigma_x, h_1', \mathcal{F}, v_x)
+      ~~~ (y, \sigma_x, \mathcal{F}, h_2) \step (y', \sigma_y, h_2', \mathcal{F}, v_y)
+    \\  \freshVar{r}
+    ~~~ h = h_1 \circ h_2
+    ~~~ h' = h_1' \circ h_2'
+    ~~~ z = x' + y'
+    ~~~ \sigma' = \sigma_y \cup \{ (r, z) \}
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[AM-Lower]{(\lowerS{A}{e}, \sigma_0, h, \mathcal{F}) \step ((C\; e_1' \cdots e_n'), \sigma', h' \cdot [H'], \mathcal{F}', r)}
+     {\begin{gathered}
+          (A[x]\; (C\; a_1 \cdots a_n) := H) \in \Sigma
+       ~~~ (e, \sigma_0, h_0) \step (C\; e_1 \cdots e_n, \sigma_1, h_1, y_1)
+       \\  (e_i, \sigma_i, h_i) \step (e_i', \sigma_{i+1}, h_{i}', v_i)\; \textnormal{for each $1 \le i \le n$}
+       \\ h' = h_1' \circ h_2' \circ \cdots \circ h_n'
+       ~~~ h = h_0 \circ h_1 \circ \cdots \circ h_n
+       ~~~ \sigma' = \sigma_{n+1} \cup \{ (r, \ell) \}
+       \\  \freshLoc{\ell}
+       ~~~ \freshVar{r}
+       \\  H' = H[x := \ell][a_1 := \sigma_2(v_1)][a_2 := \sigma_3(v_2)]\cdots[a_n := \sigma_{n+1}(v_n)]
+       \\  \mathcal{F}' = \mathcal{F} \cup \{ (\ell, (C\; e_1' \cdots e_n')) \}
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[AM-Instantiate]{(\instantiateS{A,B}{f}(e), \sigma, \mathcal{F}, h) \step (e_f', \sigma', h'', \mathcal{F}', r)}
+    {\begin{gathered}
+        (A[x]\; (C\; a) := H) \in \Sigma
+    ~~~ (f\; (C\; b) := e_f) \in \Sigma
+    \\  (e, \sigma, h) \step (C\; e_1, \sigma_1, h_1, y)
+    \\  (e_1, \sigma_1, h_1) \step (e_1', \sigma_2, h_2, r)
+    \\ \freshLoc{\ell}
+    ~~~ \freshVar{r}
+      ~~~ \freshVar{y}
+    \\  H' = H[x := \ell][a := e_1']
+    ~~~  h' = h_1 \cdot [H']
+    ~~~ \sigma' = \sigma_f \cup \{ (r, \ell) \}
+    \\  \mathcal{F}' = \mathcal{F} \cup \{ (\ell, (C\; e_1')) \}
+    \\  (\lowerS{B}{e_f[b := y]}, \sigma_2, h') \step (e_f', \sigma_f, h'', r)
+     \end{gathered}
+    }
+\end{array}
+\]
+}
+
+%   % \labinfer[AM-Constr]{(C\; e_1 \cdots e_n, \sigma, h, \mathcal{F}) \step (C\; e_1 \cdots e_n, \sigma, h, \mathcal{F})}
+%   %   {{\color{red} TODO} \textnormal{fix this}}
+% \[
+% \begin{array}{c}
+%   \labinfer[$\mathfrak{C}$-Constr]{(C\; e_1 \cdots e_n, \sigma, h, \mathcal{F}) \Cstep (C\; e_1 \cdots e_n, \sigma, h, \mathcal{F})}
+%     {}
+%   \\\\
+%   \labinfer[$\mathfrak{C}$-NonConstr]{(e, \sigma, h, \mathcal{F}) \Cstep (e', \sigma', h', \mathcal{F}')}
+%     {(e, \sigma, h, \mathcal{F}) \step (e', \sigma', h', \mathcal{F}', r)}
+% \end{array}
+% \]
+
+  \caption{Abstract machine semantics rules}
+  \label{fig:abs-machine}
+\end{figure}
+
+% Reviewing the two points of difference between the interpretations mentioned at the start of the section:
+
+\subsection{Translating \tool Specifications into SSL}
+
+We will define two translations: One from \tool \textit{expressions} into SSL propositions and the other from
+\tool \textit{definitions} into SSL inductive predicate definitions. We start with the former.
+
+\subsubsection{Translating Expressions}
+
+In the rules given in \autoref{fig:expr-rules}, the notation $\mathcal{I}_{A,B}(f)$ gives the name of the inductive predicate that the \tool
+function $f$ translates to when it is instantiated to the layouts $A$ and $B$.
+
+We start by defining the translation rules for expressions. We use these translation rules in \autoref{thm:ssem-fn} to
+define a translation function $\Tsem{\cdot}{\VarSet,r}$. Then, we will define translation rules for function definitions.
+
+The translation relation for expressions has the form $(e, \VarSet) \tstep (p, s, \VarSet', v)$,
+where $p$ and $s$ are the pure part and spatial part (respectively) of an SSL assertion and $\VarSet,\VarSet' \in \mathcal{P}(\Var)$.
+
+\begin{figure}
+  % \resizebox{.9 \textwidth}{!}
+  {
+\[
+\begin{array}{c}
+  \labinfer[S-Int]{(i, \VarSet) \tstep (v == i, \emp, \VarSet \cup \{v\}, v)}
+    {i \in \mathbb{Z} &
+     \freshVar{v}}
+  \\\\
+  \labinfer[S-Bool]{(b, \VarSet) \tstep (v == b, \emp, \VarSet \cup \{v\}, v)}
+    {b \in \mathbb{B} &
+     \freshVar{v}
+    }
+  \\\\
+  \labinfer[S-Var]{(v, \VarSet) \tstep (\texttt{true}, \emp, \VarSet, v)}
+    {v \in \Var}
+  \\\\
+  \ruleSAdd
+  \\\\
+  \labinfer[S-Lower-Var]{(\lowerS{A}{v}, \VarSet) \tstep (\texttt{true}, A(v), \VarSet \cup \{v\}, v)}
+    {\begin{gathered}
+      v \in \Var
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[S-Lower-Constr]{
+        (\lowerS{A}{C\; e_1 \cdots e_n}, \VarSet_1)
+          \tstep (p_1 \land \cdots \land p_n, H' \sep s_1 \sep \cdots \sep s_n, \VarSet', x)}
+    {\begin{gathered}
+          (A[x]\; (C\; a_1 \cdots a_n) := H) \in \Sigma
+      \\  (e_i, \VarSet_i) \tstep (p_i, s_i, \VarSet_{i+1}, v_i)\; \textnormal{for each $1 \le i \le n$}
+      \\  \freshVar{v}
+      \\  \VarSet' = \VarSet_{n+1} \cup \{x\}
+      ~~~ H' = H[a_1 := v_1]\cdots[a_n := v_n]
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[S-Inst-Var]{(\instantiateS{A,B}{f}(v), \VarSet) \tstep (\texttt{true}, \mathcal{I}_{A,B}(f)(v, r), \VarSet \cup \{r\}, r)}
+    {v \in \VarSet & \freshVar{r}}
+  \\\\
+  \labinfer[S-Inst-Constr]{
+        (\instantiateS{A,B}{f}(C\; e_1 \cdots e_n), \VarSet_1)
+                                     \tstep (p \land p_1 \land \cdots \land p_n, s \sep s_1 \sep \cdots \sep s_n, \VarSet', r)
+      }
+    {\begin{gathered}
+          (A[x]\; (C\; a_1 \cdots a_n) := H) \in \Sigma
+      \\  (e_i, \VarSet_i) \tstep (p_i, s_i, \VarSet_{i+1}, v_i)\; \textnormal{for each $1 \le i \le n$}
+      \\  \freshVar{x}
+      \\  \VarSet' = \VarSet_{n+1} \cup \{x\}
+      ~~~ H' = H[a_1 := v_1]\cdots[a_n := v_n]
+      \\
+          (f\; (C\; b_1 \cdots b_n) := e_f) \in \Sigma
+      ~~~ e_f' = e_f[b_1 := v_1]\cdots[b_n := v_n]
+      \\  (\lowerS{B}{e_f'}, \VarSet_{n+1}) \tstep (p, s, \VarSet', r)
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[S-Inst-Inst]{
+      (\instantiateS{B,C}{f}(\instantiateS{A,B}{g}(e)), \VarSet) \tstep (p_1 \land p_2, s_1 \sep s_2, \VarSet_2, r_2)
+    }
+    {\begin{gathered}
+      (\instantiateS{A,B}{g}(e), \VarSet) \tstep (p_1, s_1, \VarSet_1, r_1)
+      \\ (\instantiateS{B,C}{f}(r_1), \VarSet_1) \tstep (p_2, s_2, \VarSet_2, r_2)
+     \end{gathered}
+    }
+\end{array}
+\]
+}
+\caption{Expression Translation Rules}
+\label{fig:expr-rules}
+\end{figure}
+
+\begin{lemma}[$\Tsem{\cdot}{}$ function] \label{thm:ssem-fn} $(\cdot, \VarSet) \tstep (\cdot, \cdot, \cdot, r)$ is a computable function $\Expr \rightarrow (\Pure \times \Spatial \times \mathcal{P}(\Var))$, given fixed
+  $\VarSet$ and $r$ where $r \not\in \VarSet$.
+
+By throwing away the third element of the tuple in the codomain, we obtain a function $\Expr \rightarrow (\Pure \times \Spatial)$ from expressions to
+SSL propositions.
+
+  Call this function $\Tsem{\cdot}{\VarSet, r}$. That is, we define the function as follows where $r \not\in \VarSet$:
+  \[
+    \Tsem{e}{\VarSet, r} = (p ; s) \iff (e, \VarSet) \tstep (p, s, \VarSet', r)\; \textnormal{for some $\VarSet'$}
+  \]
+\end{lemma}
+
+We highlight the computability of this function to emphasise the fact
+that it can be used directly in an implementation of this subset of
+\tool.
+
+\subsubsection{Translating Function Definitions}
+
+The next step is to define the translation for \tool function definitions. In order to do this, we must first figure out how to determine the
+appropriate layout branch to use when unfolding a layout, a problem we highlighted earlier. Once this is accomplished, the rest of
+the translation can be defined. When this problem was solved for the abstract machine semantics,
+it was possible to simply evaluate the \tool expression until it a constructor application expression was reached.
+From there, it is possible to just look at the constructor name and match it against the appropriate layout branch.
+
+For the translation, however, we do not have the luxury of being able to evaluate expressions. Instead, we must instead rely on the fact
+that, in SSL, a ``pure'' (Boolean) condition can determine which inductive predicate branch to use. The question becomes: \textit{Given an algebraic
+data type and a layout for that ADT, how do we
+generate an appropriate Boolean condition for a given constructor for the ADT}?
+
+The solution is to find a Boolean condition which, given that the inductive predicate holds, is true \textit{if and only if} the layout branch corresponding to that
+ADT constructor is satisfiable.
+In more detail, to define the branches of an inductive predicate $\mathcal{I}_{A}(x)$, given an ADT $\alpha$, a constructor $C : \beta_1 \rightarrow \beta_2 \rightarrow \cdots \rightarrow \beta_n \rightarrow \alpha$, a layout $A : \alpha \monic \layout[x]$ with a branch $A[x]\; (C\; a_1 \cdots a_n) := H$ and given that $\mathcal{I}_{A}(x)$ holds, find a Boolean expression $b$ with one free variable $x$ such that
+\[
+  b \iff \exists \sigma, h.\; (\sigma, h) \models H
+\]
+
+\begin{lemma}[$\cond$ function]
+  \label{thm:cond}
+  There is a computable function $\cond(\cdot,\cdot)$ that takes in any layout $A : \alpha \monic \layout[x]$ with a branch $A[x]\; (C\; a_1 \cdots a_n) := H$ for a given constructor
+  $C : \beta_1 \rightarrow \beta_2 \rightarrow \cdots \rightarrow \beta_n \rightarrow \alpha$ and it produces a Boolean expression with one free variable $x$ such that the following holds under the assumption that $\mathcal{I}_{A}(x)$ holds.
+  Note that $\mathcal{I}_{A}(x)$ is the name of the generated inductive predicate corresponding to the layout $A$.
+  \[
+    \cond(A, C) \iff \exists \sigma, h.\; (\sigma, h) \models H
+  \]
+\end{lemma}
+
+With this function in hand, we are now ready to define the translation for \tool function definitions. This
+definition is in \autoref{fig:FnDef-rule}.
+
+\begin{figure}
+\[
+\begin{array}{c}
+  \labinfer[FnDef]{(f\; (C\; a_1 \cdots a_n) := e) \defstep{A,B} (\mathcal{I}_{A,B}(f)(x, r) : c \Ra \{p ; s\})}
+    {\begin{gathered}
+          \VarSet = \{ v_1 , \cdots , v_n \}\; \textnormal{where $v_1, \cdots, v_n$ are distinct variables}
+      \\  r \in \Var
+      ~~~ r \not\in \VarSet
+      ~~~ c = \cond(A, C, x)
+      \\  (p, s) = \Tsem{\instantiateS{A,B}{f}(C\; p_1 \cdots p_n)}{\VarSet, r}
+     \end{gathered}
+    }
+\end{array}
+\]
+\caption{Translation rule for function definitions}
+\label{fig:FnDef-rule}
+\end{figure}
+
+% \noindent
+% As mentioned before, we are looking at a subset of \tool where
+% functions only have a single argument. However, the solution we have just
+% described is the general solution that works for multiple arguments as well.
+
+\subsection{Typing Rules}
+
+Typing rules for \tool expressions are given in
+\autoref{fig:typing-rules}. These rules differ from standard typing
+rules for a functional language due to the existence of layouts and
+their associated constructors, like \verb|instantiate| and
+\verb|lower|. If an expression is well-typed, then each use of
+\verb|instantiate| and \verb|lower| only uses layouts together with
+the ADT that they are defined for.
+
+The rules also make use of a \textit{concreteness judgment}. The rules
+for this judgment are given in \autoref{fig:concreteness-rules}. The
+intuition of this judgment is that a type is ``concrete'' iff values
+of that type can be directly represented in the heap machine
+semantics. For example, an ADT type is \textit{not} concrete because a
+layout has not been specified. However, once a particular layout is
+specified for the ADT type, it becomes concrete. Base types, like
+\verb|Int|, are also concrete.
+
+Rules for the ensuring that global definitions are well-typed are
+given in \autoref{fig:globals-typing-rules}. In this figure, $\Delta$
+is the set of all (global) constructor type definitions.
+
+\begin{figure}[t]
+\[
+\begin{array}{c}
+  \labinfer[T-Int]{\Gamma \vdash i : \Int}
+    {i \in \mathbb{Z}}
+  ~~~
+  \labinfer[T-Bool]{\Gamma \vdash b : \Bool}
+    {b \in \mathbb{B}}
+  ~~~
+  \labinfer[T-Var]{\Gamma \vdash v : \alpha}
+    {(v : \alpha) \in \Gamma}
+  \\\\
+  \labinfer[T-Fn-Global]{\Gamma \vdash f : \alpha \rightarrow \beta}
+    {(f : \alpha \rightarrow \beta) \in \Sigma}
+  \\\\
+  \labinfer[T-Add]{\Gamma \vdash x + y : \Int}
+    {\Gamma \vdash x : \Int & \Gamma \vdash y : \Int}
+  \\\\
+  \labinfer[T-Lower-Var]{\Gamma \vdash \lowerS{A}{v} : A}
+    {(v : \alpha) \in \Gamma & (A : \alpha \monic \layout[x]) \in \Sigma}
+  \\\\
+  \labinfer[T-Lower-Constr]{\Gamma \vdash \lowerS{A}{C\; e_1 \cdots e_n} : B}
+    {\begin{gathered}
+      (C : \alpha_1 \rightarrow \cdots \rightarrow \alpha_n \rightarrow \beta) \in \Sigma
+      \\ (B : \beta \monic \layout[x]) \in \Sigma
+      \\  \Gamma \vdash e_i \concrete{\alpha_i}\; \textnormal{for each $i$ with $1 \le i \le n$}
+     \end{gathered}
+    }
+  \\\\
+  % \labinfer[T-Lower]{\Gamma \vdash \lowerS{A}{e} : A}
+  %   {\Gamma \vdash e : \alpha & (A : \alpha \monic \layout[x]) \in \Sigma}
+  % \\\\
+  \labinfer[T-Instantiate]{\Gamma \vdash \instantiateS{A,B}{f}(e) : B}
+    {\begin{gathered}
+          (A : \alpha \monic \layout[x]) \in \Sigma
+      ~~~ (B : \beta \monic \layout[y]) \in \Sigma
+      \\  \Gamma \vdash f : \alpha \rightarrow \beta
+      ~~~ \Gamma \vdash e : A
+     \end{gathered}
+    }
+  \\\\
+  \labinfer[T-Constr]{\Gamma \vdash C\; e_1 \cdots e_n : \beta}
+    {(C : \alpha_1 \rightarrow \cdots \rightarrow \alpha_n \rightarrow \beta) \in \Sigma &
+     \Gamma \vdash e_i : \alpha_i\; \textnormal{for each $i$ with $1 \le i \le n$}
+    }
+\end{array}
+\]
+  \caption{Typing rules}
+  \label{fig:typing-rules}
+\end{figure}
+
+\begin{figure}[b]
+%\vspace{-5pt}
+% \setlength{\belowcaptionskip}{-12pt}
+% \setlength{\abovecaptionskip}{0pt}
+\centering
+% \begin{subfigure}{0.49\textwidth}
+\begin{minipage}{1.0\linewidth}
+{\footnotesize{
+\[
+  \begin{array}{c}
+    \labinfer[C-Int]{e \concrete{\Int}}
+      {\Gamma \vdash e : \Int}
+    ~~~
+    \labinfer[C-Bool]{e \concrete{\Bool}}
+      {\Gamma \vdash e : \Bool}
+    \\\\
+    \labinfer[C-Layout]{e \concrete{\alpha}}
+      {(A : \alpha \monic \layout[x]) \in \Sigma &
+       \Gamma \vdash e : A
+      }
+  \end{array}
+\]
+}}
+\end{minipage}
+  \caption{Concreteness judgment rules}
+  \label{fig:concreteness-rules}
+% \end{subfigure}
+%%
+% \begin{subfigure}{0.49\textwidth}
+\begin{minipage}{1.0\linewidth}
+{\footnotesize{
+\[
+\begin{array}{c}
+  \labinfer[G-Fn]{(f\; (C\; b_1 \cdots b_n) := e) \Rightarrow f : \beta \rightarrow \gamma}
+    {(C : \alpha_1 \rightarrow \alpha_2 \rightarrow \cdots \rightarrow \alpha_n \rightarrow \beta) \in \Delta\\
+    b_1 : \alpha_1, b_2 : \alpha_2, \cdots, b_n : \alpha_n \vdash e : \gamma
+    }
+\end{array}
+\]
+}}
+\end{minipage}
+  \caption{Global definition typing}
+  \label{fig:globals-typing-rules}
+% \end{subfigure}
+%\setlength{\belowcaptionskip}{-15pt}
+\caption{Rules for concreteness judgement and typing global definitions}
+\label{fig:meh}
+\end{figure}

### 2023-08-22 new-paper/synth.tex
Bring more things over

@@
 \begin{aligned}
-  \textit{\synth-signatures}\;\; \sigma &::= p \synthsep \tau'\\
-  \textit{Pure propositions}\;\; p &::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
-    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p \mid p - p\\
-  \textit{Extended types}\;\; \tau' &::= ... \mid A\; \overline{G}\\
-  \textit{Ghost variables}\;\; G &::= @x
+  \textnormal{\synth-signatures}\;\; &\sigma ::= p \synthsep \tau'\\
+  \textnormal{Pure propositions}\;\; &p ::= x \mid i \mid b \mid p \mathbin{\&\&} p \mid p \mathbin{||} p
+    \mid \Not(p) \mid p == p \mid \{\} \mid \{p\} \mid p \app p \mid p + p\\
+  \textnormal{Extended types}\;\; &\tau' ::= ... \mid A\; \overline{G}\\
+  \textnormal{Ghost variables}\;\; &G ::= @x
 \end{aligned}
@@
 \noindent
-We also need to extend layout definitions so that they can include "ghost parameters." These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
+We also need to extend layout definitions so that they can include ``ghost parameters.'' These parameters will directly correspond to ghost variable parameters in the resulting \SuSLik{} specification.
 

### 2023-08-22 new-paper/introduction.tex
bibliography; a few sentences on layout polymorphism

@@
 
-Synthetic separation logic (SSL) is the specification language for the \SuSLik{} program synthesizer. \SuSLik{} will take a precondition and postcondition pair together
-with a function prototype and use these to synthesize a function that satisfies that specification in a C-like language called \SuSLang.
+Synthetic separation logic (SSL) is the specification language for the \SuSLik{} program synthesiser. \SuSLik{} will take a precondition and postcondition pair together
+with a function prototype and use these to synthesise a function that satisfies that specification in a C-like language called \SuSLang.
 
@@
 In addition to this translational semantics, we also provide a standard abstract machine semantics for \Pika{} and use this to show correctness of the translation into SSL.
+
+Two versions of \Pika{} have been built.
+\Pika{} 1 has a focus on using \SuSLik{} to generate all of the low-level code, while \Pika{} 2 adds an additional synthesis feature inside the \Pika{} language itself and has a builtin testing framework.
+In the present work, we will start by talking about \Pika{} 1, describe the fundamental limitations of using \SuSLik{} to generate everything and then use this to motivate the creation of \Pika{} 2.
+The two versions are then compared.
 

### 2023-08-22 new-paper/limitations.tex
Bring more things over
bibliography; a few sentences on layout polymorphism

@@
 
-There is a significant limitation to our approach so far: \SuSLik will often fail to synthesize code from valid specifications.
+There is a significant limitation to our approach so far: \SuSLik{} will often fail to synthesize code from valid specifications.
 One instance of this problem is the following: In a certain class of functions, \SuSLik{} requires a different number of pointer indirections than it will with other functions.
@@
 This allows us to easily bring all of these things together in our new implementation with the new \Pika{} synthesis feature.
+
+Furthermore, \Pika{} 1 relies on a kind of \textit{implicit} layout polymorphism: The layouts used are all inferred.
+There are no type variables that represent layouts to constrain a type signature.
+Towards solving this, \Pika{} 2 also uses explicit layout passing.
+We have not yet implemented full layout polymorphism, but this can be built on top of explicit layout passing.

### 2023-08-22 new-paper/semantics.tex
bibliography; a few sentences on layout polymorphism

@@
 
-\subsection{Formal Semantics of \tool}
-\label{sec:formal}
+% \subsection{Formal Semantics of \tool}
+% \label{sec:formal}
 
@@
 
-\subsubsection{Overview of the Two Interpretations}
+\subsection{Overview of the Two Interpretations}
 
@@
 % \subsubsection{Abstract Machine}
+\paragraph{Abstract machine}
 First, consider the abstract machine semantics. In this case, we are able to choose which branch of a layout to use by evaluating the expression we are applying it
@@
 % \subsubsection{SSL Translation}
+\paragraph{SSL translation}
 Next, consider the SSL translation. Here we can determine which layout branch to use by generating a Boolean condition that will be
@@
 
-\subsubsection{Abstract Machine Semantics}
+\subsection{Abstract Machine Semantics}
 In this section, we will define an abstract machine semantics for \tool and relate this to
@@
   \label{fig:globals-typing-rules}
-% \end{subfigure}
-%\setlength{\belowcaptionskip}{-15pt}
-\caption{Rules for concreteness judgement and typing global definitions}
-\label{fig:meh}
+% % \end{subfigure}
+% %\setlength{\belowcaptionskip}{-15pt}
+% \caption{Rules for concreteness judgement and typing global definitions}
+% \label{fig:meh}
 \end{figure}

### 2023-08-22 new-paper/suslik-extensions.tex
SuSLik extension section

\section{Extensions of \suslik}
\label{sec:suslik-extensions}

We have shown the translation from the functional specifications into
SSL specifications. However, some of the SSL specifications are not
supported in the original \suslik and existing variants.
%
In this section, we present how to extend the \suslik to support more
features to make the whole thing work. We will show the extensions on
the following three aspects:

\begin{itemize}
  \item How to describe and call an existing function within SSL predicates.
  \item How to make the result of one function call as the input of another function call.
  \item How to synthesise programs with inductive predicates without the help of pure theory.
\end{itemize}

\subsection{Function Predicates}
\label{sec:funcPred}

Without any modification upon the implementation, we find the SSL
predicate within some restrictions can be used to describe function
relations other than data structures (named function predicates). The
definition of \textbf{function predicates} is as follows:

\begin{definition}[Function Predicates]
    \label{def:funcPred}
    \normalfont
    Given any non-higher-order n-ary function $f$\lstinline{(x1, ..., xn)} in the functional language, the function predicate to synthesise $f$ has the following format:
    \begin{lstlisting}[language=SynLang]
    predicate predf(T x1, ... ,T xn, T output){...}
    \end{lstlisting}
%
    where \lstinline{T} $\in$ \{\lstinline{loc}, \lstinline{int}\}.
    The type of \lstinline{xi} (and \lstinline{output}) is decided by
    the type of $f$. If it is an integer in $f$, then its type is
    \lstinline{int}; otherwise, it is \lstinline{loc} (for any data
    structure in \tool).
\end{definition}

Since the input of the whole workflow is functional programs, the
"output" in the definition is to provide another location for the
output of the function. And the specification to synthesise function
$f$ should have the following format:

\begin{lstlisting}[language=SynLang]
void f(loc x1, ... ,loc xn)
{x1 :-> v1 ** x2 :-> l2 ** sll(l2) ** ... ** xn :-> vn ** output :-> 0}
{x1 :-> v1 ** x2 :-> l2 ** ... ** xn :-> vn ** output :-> output0 ** 
 predf(v1, l2, ..., vn, output0)}
\end{lstlisting}

\subsection{SSL Rules for \func Structure}

As we show in previous examples, the reason we can have \func structure is that the points-to structure in the post-condition is always eliminated after some write operations. For example, the in-placed \lstinline{inc1} functions specification is satisfied via the \writer operation (\autoref{fig:write}) on the location.

\begin{lstlisting}[language=SynLang]
void inc_y(loc y, loc x)
{x :-> vx ** y :-> vy}
{x :-> vx + xy ** y :-> vy}
\end{lstlisting}

\begin{figure}[t]
    \centering
    \begin{mathpar}
      \inferrule[\writer]
      {
      \mcode{\vars{e} \subseteq \env}
      \\
      \mcode{e \neq e'}
      \\
      \trans{\mcode{{\asn{\phi; \ispointsto{x}{e}{} \osep P}}}}
              {\mcode{\asn{\psi; \ispointsto{x}{e}{} \osep Q}}} {\mcode{\prog}}
      }
      {
      \mcode{
      \trans{\asn{\phi; {x} \pts e' \osep P}}
      {\asn{\psi; {x} \pts e \osep Q}}
      {{\deref{x} = e\ ;\ \prog}}
      }
      }
    \end{mathpar}

    \caption{The \writer rule in SSL}
    \label{fig:write}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{mathpar}
    \inferrule[\funcwrite]
      {
      \mcode{\forall i \in [1, n], \vars{e_i} \subseteq \env}
      \\
      \trans{\mcode{{\asn{\phi; P}}}}
            {\mcode{\asn{\psi; Q}}} {\mcode{\prog}}
      }
      {
      \mcode{\trans{\asn{\phi; {x} \pts e \osep P}}
      {\asn{\psi; \func\ f(e_1,\ldots,e_n,{x})\osep Q}}
      {{f(e_1,\ldots,e_n,{x})\ ;\ \prog}}
      }
      }
  \end{mathpar}

  \caption{The \funcwrite rule in SSL}
  \label{fig:funcwrite}
\end{figure}

The core insight of \func structure is: since the function synthesised
by function predicate behaves like the pure function, it is the same
as the \writer rule in the sense that only the output location is
modified. Thus, we add the new \funcwrite rule into the zoo of SSL
rules (see \autoref{fig:funcwrite}).
%
To make the \func structure correctly equal to some ``write''
operation, the following restrictions should hold, which are achieved
by the translation:

\begin{itemize}
    \item If \lstinline[language=SynLang]{func f(x1, ..., xn, output)} appears in a post-condition, then no write rule can be applied to any \lstinline[language = SynLang]{xi}. This is to avoid the ambiguity of the \func.
    \item The type of function f is consistent.
\end{itemize}

Note that based on the setting of the function predicate, the parameters of the function call are pointers, while the parameters of the function predicate are content to which pointers point. Furthermore, we have the \func generated from function predicates and with the format defined in \autoref{sec:funcPred}. As a result, the equivalent original SSL that duplicates points-to of one location is not a problem, since they can be merged as one.

\subsection{Temporary Location for the Sequential Application}

Though with rich expressiveness, SSL has difficulty in expressing the sequential application of functions. For example, given the \func structure available, the following function is not expressible within one function predicate:

\begin{lstlisting}[language=SynLang]
f x y = g (h x) y
\end{lstlisting}

If we attempt to express it, we will have the following part in the predicate:

\begin{lstlisting}[language=SynLang]
predicate f(loc x, loc y, loc output)
{... ** func h(x, houtput) ** func g(houtput, y, output)}
\end{lstlisting}

However, \lstinline{houtput} is not a location in the pre-condition, which is not allowed in SSL. Thus, we introduce a new keyword \lstinline{temp} to denote the temporary location for the sequential application. The new definition of \lstinline{func} is as follows:

\begin{lstlisting}[language=SynLang]
predicate f(loc x, loc y, loc output)
{... ** temp houtput ** func h(x, houtput) ** func g(houtput, y, output)}
\end{lstlisting}

Roughly speaking, the \lstinline{temp} structure will help to allocate a new location for the output of the first function, and then use it as the input of the second function. After all appearances of \lstinline{houtput} is eliminated, we will deallocate the location.

Note that the temporary variable is possible to appear in two different structures: recursive function predicates or \func call. The reason we don't need to consider the basic arithmetic operations is that the integer will be directly used as the predicate parameter, instead of the location as the parameter. For example, the sum of a list can be expressed as:

\begin{lstlisting}[language=SynLang]
predicate sum(loc l, int output){
| l == 0 => {output == 0; emp}
| l != 0 => {output == output1 + v; [l, 2] ** l :-> v ** l + 1 :-> lnxt ** sum(lnxt, output1)}
}
\end{lstlisting}

Such sequential application is common in functional programming, especially in the recursive function. For example, it is not elegant to flatten a list of lists without the sequential application.

\begin{lstlisting}[language=SynLang]
    flatten :: [[a]] -> [a]
    flatten [] = []
    flatten (x:xs) = x ++ flatten xs
\end{lstlisting}

We can express this function, but with some strange structure to store all temporary lists.

\begin{lstlisting}[language=SynLang]
predicate flatten(loc x, loc output){
| x == 0 => {output :-> 0}
| x != 0 => {[x, 2] ** x :-> x0 ** sll(x0) ** x + 1 :-> xnxt **
[output, 2] ** func append(x, outputnxt, output) ** output + 1 :-> outputnxt **
flatten(xnxt, outputnxt)}
}
\end{lstlisting}

With such a function predicate, though we can synthesise the function
whose result stored in \lstinline{output} is the flattened list, the
list \lstinline{output} is containing a lot of intermediate values,
which is neither consistent with the definition in the source language
nor space efficient.

\begin{figure}[t]
  \centering
  \begin{mathpar}
    \inferrule[\tempfuncalloc]
      {
      \trans{\mcode{{\asn{\phi; x \pts a \osep P}}}}
      {\asn{\psi; \func\ f(e_1,\ldots,e_n,{x})\osep temp (x, 1) \osep Q}} {\mcode{\prog}}
      }
      {
      \mcode{
      \trans{\asn{\phi; P}}
      {\asn{\psi; \func\ f(e_1,\ldots,e_n,{x})\osep temp (x, 0) \osep Q}}
      {{let\ x\ =\ malloc(1)\ ;\ \prog}}
      }
      }
  \end{mathpar}
  % \begin{mathpar}
  %   \inferrule[\temppredalloc]
  %     {
  %     \trans{\mcode{{\asn{\phi; x \pts x0 \osep x0 \pts a \osep P}}}}
  %     {\asn{\psi; p(e_1,\ldots,e_n,{x_0})\osep temp (x, 2, x_0) \osep x \pts x_0 \osep Q}} {\mcode{\prog}}
  %     }
  %     {
  %     \mcode{
  %     \trans{\asn{\phi; P}}
  %     {\asn{\psi; p(e_1,\ldots,e_n,{x})\osep temp (x, 0, \_) \osep Q}}
  %     {}
  %     }
  %     \\
  %     \mcode{
  %       let\ x\ =\ malloc(1);
  %       let\ x0\ =\ malloc(1);
  %       *x\ =\ x0\ ; \prog
  %     }
  %     }
  % \end{mathpar}

  \caption{New allocation rule for \lstinline{temp} in SSL }
  \label{fig:newalloc}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{mathpar}
    \inferrule[\tempfuncfree]
      {
      \trans{\mcode{{\asn{\phi; P}}}}
      {\asn{\psi; Q}} {\mcode{\prog}}
      }
      {
      \mcode{\not\exists x\in Q\ \wedge
      \trans{\asn{\phi; P}}
      {\asn{\psi;  temp (x, 1) \osep Q}}
      {}
      }\\
      \mcode{let\ x0\ =\ *x\ ;\ type\_free(x0);\ free(x);\ \prog}
      }
  \end{mathpar}
  % \begin{mathpar}
  %   \inferrule[\temppredfree]
  %     {
  %     \trans{\mcode{{\asn{\phi; P}}}}
  %     {\asn{\psi; Q}} {\mcode{\prog}}
  %     }
  %     {
  %     \mcode{\not\exists x_0\in Q\ \wedge
  %     \trans{\asn{\phi; P}}
  %     {\asn{\psi;  temp (x, 2, x_0) \osep  Q}}
  %     {}
  %     }\\
  %     \mcode{let\ x0\ =\ *x\ ;\ let\ x00\ =\ *x0\ ;\ ds\_free(x00);\ free(x0);\ free(x);\ \prog}
  %     }
  % \end{mathpar}

  \caption{New deallocating rule for \lstinline{temp} in SSL}
  \label{fig:newfree}
\end{figure}

The new rules consist of allocating (\autoref{fig:newalloc}) and deallocating rules (\autoref{fig:newfree}). Based on the definition of the \func structure and the function predicate, the allocated locations are different, where the \lstinline{temp} location for \func is directly used; while the \lstinline{temp} location for function predicate should allocate a new location for function predicates. As for the deallocation, not only the \lstinline{temp} location(s) but also the content they point to should be deallocated. That is the reason we have the \lstinline{type_free} function, which is syntax sugar to deallocate the content of a location based on the type information. For example, if the type of the location is \lstinline{tree}, then the \lstinline{type_free} will deallocate the content of the location via \lstinline{tree_free} function, which is synthesised based on the SSL predicate \lstinline{tree} as follows.
\begin{lstlisting}[language=SynLang]
void tree_free(loc x)
  {tree(x)}
  {emp}
\end{lstlisting}
Specifically, if the location contains the value with type \lstinline{int}, then the \lstinline{type_free} will do nothing.
Thus, the function predicate with \lstinline{temp} is much better, in the sense that no extra space is used, and the synthesised function is consistent with the source language.

\begin{lstlisting}[language=SynLang]
predicate flatten(loc x, loc output){
| x == 0 => {output :-> 0}
| x != 0 => {[x, 2] ** x :-> x0 ** sll(x0) ** x + 1 :-> xnxt ** 
temp outputnxt ** flatten(xnxt, outputnxt) ** func append(x, outputnxt, output)}
}
\end{lstlisting}

\subsection{Avoiding Excessive Heap Manipulation with Read-Only Locations}

The existing \suslik depends on the set theory to express the pure
relation. However, it is not trivial to automatically generate the
pure part of SSL specifications from the functional specifications. To
see why the set theory is needed, the following simple example shows
the functionality of the set theory, with \lstinline{sll_n} being the
\textbf{s}ingle-\textbf{l}inked \textbf{l}ist with \textbf{n}o set.

\begin{lstlisting}[language=SynLang]
predicate cp(loc x, loc y) {
|  x == 0        => {y == 0; emp }
|  not (x == 0)  => {
    [y, 2] ** y :-> v ** (y + 1) :-> ynxt **
    [x, 2] ** x :-> v ** (x + 1) :-> xnxt ** cp(xnxt, ynxt) }
}
\end{lstlisting}

While the intent of the function predicate \lstinline{cp} is to copy
the list \lstinline{x} to \lstinline{y}, without the set theory, the
output program will be somewhat surprising to see:

\begin{lstlisting}[language=SynLang]
void copy (loc x, loc y) {
  if (x == 0) {
  } else {
    let n = *(x + 1);
    copy(n, y);
    let y01 = *y;
    let y0 = malloc(2);
    *y = y0;
    *(y0 + 1) = y01;
    let vy = *y0
    *x = vy;
  }
}
\end{lstlisting}

The problem here is that, when we have the pure relation in the
predicate to indicate that the values are the same, the synthesiser
finds another possible way: instead of copying the value of
\lstinline{x} to \lstinline{y}, we can just change the value of x to
initial value \lstinline{vy} after \lstinline[language = c]{malloc}.
This is not the user intent, and the output program is not correct.
%
Turns out, the solution is not that difficult: we simply need add a
new kind of heaplet in the specification language, call
\textit{constant points-to}, which has a similar idea as read-only
borrows~\cite{costea2020concise}.
%
% Unlike the complex permissions in the robosuslik designed for
% different purposes, we only need to get rid of the set theory in the
% function predicate. Since the source language is a functional
% language, it will be natural to make the input remain unmodified. 
%
The only difference of the constant points-to from the original
\textit{points-to} heaplet is that the value of the location is
constant, which means that the \writer rule in SSL is not applicable.

% \subsection{In-place Function by Basic In-place Description}

% With all extensions above, we have a complete specification language for the translation result of a basic functional source. However, all result programs are performed as pure functions. That is a normal result, considering the definition of functional language. But in some cases, the user may want to have the in-place function, which means that the input will be modified. For example, the user may want to have a function to reverse a list, and the input list will be reversed after the function call. In this case, such intent is impossible to be expressed by the basic source language, which is the reason we have another extension in the functional specification. As shown previously, the function predicate in SSL is to describe the pure function. But for the in-place function, it is kind of necessary for the user to describe the memory behaviour of the function.

% As the example in Sec. \ref{sec:overview} shows, the good point is that the higher-order in-place function can be achieved by the basic in-place function. For the simplest case, no more effort is needed when the basic in-place function itself is  following the restriction of \func structure. However, it is possible that some basic in-place functions are modifying multiple locations. For example, the in-place version of function \lstinline{cons} has the following specification after translation into \suslik.

% \begin{lstlisting}[language=SynLang]
%   void cons(loc x, loc xs)
%   {x :-> v ** xs :-> vxs ** sll(vxs)}
%   {[x, 2] ** x :-> v ** x + 1 :-> vxs ** sll(vxs) ** xs :-> 0}
% \end{lstlisting}

% Then the previous \func will not work. But following the basic idea of \func, where the execution of function call performs similar as \writer , we can have the following extension of \inplace structure, which is similar to \func except that the \lstinline{output} becomes a list of locations which are modified in the specification.
% \begin{lstlisting}
%   in_place f(e, ..., [x1, .., xn])
% \end{lstlisting}
% And a similar extension of SSL rules is shown in \autoref{fig:inplacewrite}.

% \begin{figure}[t]
%   \centering
%   \begin{mathpar}
%     \inferrule[\inplacewrite]
%       {
%       \mcode{\forall i \in [1, n], \vars{e_i} \subseteq \env}
%       \\
%       \env; \trans{\mcode{{\asn{\phi; P}}}}
%             {\mcode{\asn{\psi; Q}}} {\mcode{\prog}}
%       }
%       {
%       \mcode{
%       \env; \translong{\asn{\phi; {x_1} \pts e_1' \sep \ldots \sep {x_n} \pts e_n' \sep P}}
%       {\asn{\psi; \inplace~f(e_1,\ldots,e_n,[{x_1},\ldots,{x_n}])\sep Q}}
%       {}
%       }
%       \\
%       \mcode{{f(e_1,\ldots,e_n,{x_1},\ldots, {x_n})\ ;\ \prog}}
%       }
%   \end{mathpar}

%   \caption{The \inplacewrite rule in SSL \todo{Gamma bug}}
%   \label{fig:inplacewrite}
% \end{figure}

% Moreover, we need to deal with the combination of basic in-place functions and temporary location, where the automatically free should be handled subtly.

### 2023-08-22 new-paper/conclusion.tex
Pika 2 examples

@@
 In order to do this, we built the notion of a \textit{layout} of an algebraic datatype on top of separation logic.
-Then we observed that \SuSLik, the tool we use to synthesize programs from SSL, has notable limitations in its ability to synthesize the SSL specifications we generate.
+Then we observed that \SuSLik, the tool we use to synthesise programs from SSL, has notable limitations in its ability to synthesise the SSL specifications we generate.
 This motivated the creation of the \synth{} keyword.

### 2023-08-22 new-paper/examples.tex
bibliography; a few sentences on layout polymorphism
Pika 2 examples
Some stats for Pika 2 examples
Description of the example tables

@@
 \subsection{\Pika{} 1}
+
+Working \Pika{} 1 examples are presented in Figure~\ref{size-comparison}.
+For each example, we compare the AST size of the \Pika{} code with that of the generated \SuSLik{} code.
+This is a rough measure of the relative expressiveness of \Pika{} and \SuSLik{} on these examples.
+Each example in the list consists of a single function which was tested by hand.
+
+\begin{figure}[b]
+\setlength{\abovecaptionskip}{5pt}
+\setlength{\belowcaptionskip}{-15pt}
+\begin{center}
+  \begin{table}[H]
+  \begin{tabular}{ c c c c }
+    \hline
+    Name & \suslik size & \tool 1 size & \tool 1 size / \suslik size\\
+    \hline
+    \verb|append| & 164 & 84 & 0.512 \\
+    \verb|cons| & 82 & 52 & 0.634 \\
+    \verb|filterLt9| & 136 & 66 & 0.485 \\
+    \verb|singleton| & 70 & 47 & 0.671 \\
+    \verb|map| & 147 & 93 & 0.633 \\
+    \verb|mapSum| & 138 & 97 & 0.703 \\
+    \verb|maximum| & 106 & 61 & 0.575 \\
+    \verb|leftList| & 144 & 104 & 0.722 \\
+    \verb|take| & 180 & 112 & 0.622
+  \end{tabular}
+  \end{table}
+\end{center}
+  \caption{\tool 1 spec size vs generated SSL spec size measured in number of AST nodes}
+  \label{fig:size-comparison}
+\end{figure}
 
@@
 The second aspect, in particular, is a substantial change from version 1 and motivated the rewrite.
+
+Figure~\ref{fig:pika-2-examples} is a table of \Pika{} 2 examples together with the number of functions they have (not including \synth) and whether or not they use \synth.
+These functions were tested using the unit testing framework built into \Pika{} 2.
+
+\begin{figure}
+  \begin{tabular}{ c c c }
+    \hline
+    Name & \# of functions & Uses \synth\\
+    \hline
+    \verb|add1Head| & 1 & No\\
+    \verb|anagram| & 4 & No\\
+    \verb|fact| & 1 & No\\
+    \verb|filterLt| & 1 & No\\
+    \verb|heap| & 6 & Yes\\
+    \verb|leftList| & 1 & No\\
+    \verb|mapAdd| & 1 & No\\
+    \verb|maximum| & 1 & No\\
+    \verb|set| & 3 & Yes\\
+    \verb|sum| & 1 & No\\
+    \verb|take| & 1 & No\\
+    \verb|treeSize| & 1 & No\\
+  \end{tabular}
+  \caption{\Pika{} 2 examples}
+  \label{fig:pika-2-examples}
+\end{figure}

### 2023-08-22 new-paper/overview.tex
bibliography; a few sentences on layout polymorphism
code formatting

@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=SynLang]
 void add1Proc(int x, loc r)
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 add1Proc : Int -> Int;
@@
 application. This is also written $\instantiateS{A,B}{f}(x)$. The
-latter is primarily used in \autoref{sec:formal}.
+latter is primarily used in \autoref{sec:semantics}.
 
@@
 that the layout $A$ is for the algebraic data type $\alpha$ and the
-SSL output variable is named $x$. The full \Pika{} grammar is given in
-the Appendices. %x~\ref{sec:grammar}.
+SSL output variable is named $x$.
+% The full \Pika{} grammar is given in
+% the Appendices. %x~\ref{sec:grammar}.
 
 \begin{figure}[t]
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 data List := Nil | Cons Int List;
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate singleton [Int] Sll
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=SynLang]
 predicate singleton(int p, loc r) {
@@
 \begin{figure}[t]
-\begin{lstlisting}
+  \begin{lstlisting}[language=SynLang]
 predicate sll(loc x) {
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 mapAdd1_fn : List -> List;
@@
 %
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate mapAdd1_fn [Sll] Sll
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate mapAdd1 [Sll] Sll
@@
 \begin{figure}
-  \begin{lstlisting}
+  \begin{lstlisting}[language=SynLang]
 predicate filterLt9(loc x, loc r) {
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate filterLt9 [Sll] Sll
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate maximum [Sll] Int
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate leftList [TreeLayout] Sll
@@
 \verb|addr| is the corresponding value-level introduction operation.
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 %generate fold_List [Int, Sll] (Ptr Int)
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=SynLang]
 predicate fold_List(int i, loc p_x ,loc __r) {

### 2023-08-22 new-paper/synth.tex
bibliography; a few sentences on layout polymorphism
Clarify section title
code formatting

@@
-\section{\synth{} Keyword: Synthesis Inside \Pika}
+\section{\synth{} Keyword: Synthesis Inside \Pika{} 2}
 \label{sec:synth}
 
-Next, we introduce a feature that allows us to synthesize a program from a special kind of \Pika{} type signature.
+Next, we introduce a feature that allows us to synthesise a program from a special kind of \Pika{} type signature.
 This special kind of type signature allows us to directly access \SuSLik's pure propositions.
@@
 
-\begin{lstlisting}
-Sll : layout[x](List)
+\begin{lstlisting}[language=Pika]
+Sll : List >-> layout[x]
 Sll Nil := emp;
@@
 
-\begin{lstlisting}
-Sll : @(s : set) layout[x](List);
+\begin{lstlisting}[language=Pika]
+Sll : @(s : set) List >-> layout[x];
 Sll Nil := s == {} ;; emp;
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Pika]
 synth append :

### 2023-08-22 new-paper/overview.tex
related work; more consistency in notation

@@
 Sll (Nil) := emp;
-Sll (Cons head tail) := x :-> head, (x+1) :-> tail, Sll tail;
+Sll (Cons head tail) := x :-> head ** (x+1) :-> tail ** Sll tail;
 \end{lstlisting}
@@
 Sll (Nil) := emp;
-Sll (Cons head tail) := x :-> head, (x+1) :-> tail, Sll tail;
+Sll (Cons head tail) := x :-> head ** (x+1) :-> tail ** Sll tail;
 
@@
 TreeLayout (Node payload left right) :=
-  x :-> payload,
-  (x+1) :-> left,
-  (x+2) :-> right,
-  TreeLayout left,
+  x :-> payload **
+  (x+1) :-> left **
+  (x+2) :-> right **
+  TreeLayout left **
   TreeLayout right;

### 2023-08-22 new-paper/related-work.tex
related work; more consistency in notation

\section{Related Work}
\label{sec:related-work}

\tool is built upon the \suslik synthesis framework. \suslik provides
a synthesis mechanism for heap-manipulating programs using a variant
of separation logic.~\cite{polikarpova:2019:suslik} However, it does
not have any high-level abstractions. In particular, writing \suslik
specifications directly involves a significant amount of pointer
manipulation. Further, it does not provide abstraction over specific
memory layouts. As described in \autoref{sec:language}, \tool
addresses these limitations.

The \tname{Dargent} language~\cite{chen:2023:dargent} also includes a
notion of layouts and layout polymorphism for a class of algebraic
data types, which differs from our treatment of layouts in two primary
ways:

\begin{enumerate}

\item In \tool, abstract memory locations (with offsets) are used. In
  contrast, \tname{Dargent} uses offsets that are all
  relative to a single ``beginning'' memory location. The \tool
  approach is more amenable to heap allocation, though
  this requires a separate memory manager of some kind. This is
  exposed in the generated language with \verb|malloc| and
  \verb|free|.
  On the other hand, the technique taken by \tname{Dargent} allows
  for greater control over memory management. This makes
  dealing with memory more complex for the programmer, but it is no
  longer necessary to have a separate memory manager.

  \item Algebraic data types in the present language include
    \emph{recursive} types and,
    as a result, \tool has recursive layouts for these ADTs. This
    feature is not currently available in \tname{Dargent}.
  \end{enumerate}

Furthermore, layout polymorphism also works differently. While
\tname{Dargent} tracks layout instantiations at a type-level with type
variables, in the present work we simply only check to see if a layout
is valid for a given type when type-checking. In particular, we cannot
write type signatures that \textit{require} the same layout in
multiple parts of the type (for instance, in a function type
\verb|List -> List| we have no way at the type-level of requiring that
the argument \verb|List| layout and the result \verb|List| layout are
the same). \Pika{} 2 makes progress towards eliminating this restriction but, as of now, it is layout monomorphic. This more rudimentary approach that \tool currently takes
could be extended in future work.
%
Overall, the examples in the \tname{Dargent} paper tend to focus on
the manipulation of integer values. In contrast, we have focused
largely on data structure manipulation, which follow the primary
motivation of \suslik.

\tname{Synquid} is another synthesis framework with a functional
surface language. While \tname{Synquid} allows an even higher-level
program specification than \tool through its liquid types, it does not
provide any access to low-level data structure
representation.~\cite{polikarpova:2016:synquid} In contrast, \tool's
level of abstraction is similar to that of a traditional functional
language but, similar to \tname{Dargent}, it also allows control over
the data structure representation in memory. The \synth{} feature
has similarities to \tname{Synquid}'s approach to program synthesis, but
\Pika{} 2's \synth{} allows lower level constraints to be expressed.

### 2023-08-22 new-paper/layouts.tex
'layouts' section

\subsection{Layouts}
\label{sec:layouts}

A \textit{layout} is a mapping from algebraic datatype values to separation logic assertions. This is achieved by a special kind of function definition that pattern matches on the algebraic datatype and produces SSL assertions. This function can be recursive and it can be mutually recursive with other layouts.

This construct is at the heart of the connection between SSL and \Pika, providing a bridge between the algebraic datatypes of a functional programming language and the heap states of an imperative language as seen through the lens of separation logic.

### 2023-08-22 new-paper/introduction.tex
Describe layouts

@@
 
-  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic datatypes and their representation in memory.
-    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic datatype. (Section~\ref{sec:layouts})
+  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic data types and their representation in memory.
+    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic data type. (Section~\ref{sec:layouts})
 

### 2023-08-22 new-paper/layouts.tex
Describe layouts

@@
 
-A \textit{layout} is a mapping from algebraic datatype values to separation logic assertions.
-This is achieved by a special kind of function definition that pattern matches on the algebraic datatype and produces SSL assertions.
+A \textit{layout} is a mapping from algebraic data type values to separation logic assertions.
+This is achieved by a special kind of function definition that pattern matches on the algebraic data type and produces SSL assertions.
 This function can be recursive and it can be mutually recursive with other layouts.
+We have seen the \verb|Sll| layout so far (Fig.~\ref{fig:List-def}), which is a layout for linked lists.
+In particular, it maps values of the \verb|List| type to SSL assertions representing a singly-linked list.
 
-This construct is at the heart of the connection between SSL and \Pika, providing a bridge between the algebraic datatypes of a functional programming language and the heap states of an imperative language as seen through the lens of separation logic.
+This construct is at the heart of the connection between SSL and \Pika, providing a bridge between the algebraic data types of a functional programming language and the heap states of an imperative language as seen through the lens of separation logic.

### 2023-08-22 new-paper/paper.tex
Bring more things over
SuSLik extension section
related work; more consistency in notation
Describe layouts
a small fix

@@
 \usepackage{listings}
+\usepackage{xspace}
+\usepackage{mathtools}
+\usepackage{mathpartir}
 \lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}
@@
   specification in a C-like language. However, these specifications are very low-level and SSL does not allow for much abstraction. In this paper,
-  we explore a connection between SSL specifications and functional programming. We give a first-order functional
+  we explore a connection between SSL specifications and functional programming. We give a functional
   programming language, \Pika, which has both a standard semantics and a translation into SSL that respects the standard semantics.
@@
 
+\input{selected-rules}
+
 \input{introduction}
 \input{overview}
-\input{syntax}
-\input{layouts}
+% \input{syntax}
 \input{semantics}
+\input{suslik-extensions}
 \input{limitations}
@@
 \input{examples}
+\input{related-work}
 \input{future-work}
 \input{conclusion}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{paper}
 

### 2023-08-22 new-paper/introduction.tex
a small fix
formatting

--- 2023-08-22 new-paper/introduction.tex
@@
 
-We have developed a first-order functional programming language, \Pika, which will compile to SSL specifications.
+We have developed a  functional programming language, \Pika, which will compile to SSL specifications.
 \SuSLik{} can then take these SSL specifications and generate \SuSLang.
@@
 
-  \item In order to provide seamless interoperability between SSL and \Pika, we provide a separation between algebraic datatypes and their representation in memory.
-    At a type-level, we track this representation in memory. Furthermore, we describe a kind of polymorphism over memory layouts of an algebraic datatype. (Section~\ref{sec:layouts})
+  \item In order to provide seamless interoperability between SSL and \Pika, we provide a mechanism to related algebraic data types to their representation in memory.
+In particular, Pika has a language construct that allows the programmer to define a kind of function from values of an algebraic data type to SSL assertions.
+This is called a \textit{layout}.
+(Section~\ref{sec:layouts})
 

### 2023-08-22 new-paper/layouts.tex
formatting

@@
 
+These functions should be injective, as the translator should be able to distinguish between the different constructors of the algebraic data type value exclusively using the SSL assertions.
+This injectivity requirement is not enforced by the compiler.
+Also, the choice of arrow in the layout type declaration syntax \lstinline[language=Pika]{L : T >-> layout[x]} is motivated by this requirement, since this kind of double-headed arrow is traditionally used for injective functions.
+
 This construct is at the heart of the connection between SSL and \Pika, providing a bridge between the algebraic data types of a functional programming language and the heap states of an imperative language as seen through the lens of separation logic.

### 2023-08-22 new-paper/synth.tex
related work; more consistency in notation
a bit on differences between the two versions

@@
-\section{\synth{} Keyword: Synthesis Inside \Pika{} 2}
+\section{\texttt{synth} Keyword: Synthesis Inside \Pika{} 2}
 \label{sec:synth}
@@
 \noindent
+Note that in \Pika{} 2, we distinguish between the pattern variables (such as \verb|tail|) and the SSL parameters (such as \verb|nxt|).
+In the recursive application of \verb|Sll|, we apply it to \verb|tail| and we substitute \verb|nxt| for \verb|x|.
+
 This does not use any ghost variables in \Pika{} or in the resulting \SuSLik{} specification.

### 2023-08-30 new-paper/paper.tex
Improve abstract

@@
   programming language, \Pika, which has both a standard semantics and a translation into SSL that respects the standard semantics.
+This translation also makes interfacing between SSL and \Pika{} easy.
+We introduce ``layouts'' to bridge the gap between a functional programming language and SSL called.
+This allows us to describe the memory layout of algebraic data types as SSL predicates inside \Pika.
 
-  Along the way, we encounter limitations in the \SuSLik{} tool. In confronting these limitations, we develop a new feature
-  in \Pika{} that allows the programmer to express synthesis tasks at a higher-level.
+  We then develop a second version of \Pika{} extended with an additional synthesis feature. These two versions are compared.
+
+  % Along the way, we encounter limitations in the \SuSLik{} tool. In confronting these limitations, we develop a new feature
+  % in \Pika{} that allows the programmer to express synthesis tasks at a higher-level.
 

### 2023-08-30 new-paper/introduction.tex
work on contributions in intro

@@
 \Pika{} 1 has a focus on using \SuSLik{} to generate all of the low-level code, while \Pika{} 2 adds an additional synthesis feature inside the \Pika{} language itself and has a builtin testing framework.
-In the present work, we will start by talking about \Pika{} 1, describe the fundamental limitations of using \SuSLik{} to generate everything and then use this to motivate the creation of \Pika{} 2.
+
+%In the present work, we will start by talking about \Pika{} 1, describe the fundamental limitations of using \SuSLik{} to generate everything and then use this to motivate the creation of \Pika{} 2.
 The two versions are then compared.
@@
 
-  \item We address the limitations of this approach.
-In particular, we develop an extension of \Pika{} that allows \SuSLik's synthesis capabilities to be exposed more naturally inside the \Pika{} language.
-(Section~\ref{sec:synth})
+  \item We develop an extension of \Pika{} that allows \SuSLik's synthesis capabilities to be exposed more naturally inside the \Pika{} language. (Section~\ref{sec:synth})
 

### 2023-09-02 new-paper/suslik-extensions.tex
formatting
Minor
revised for the first round

@@
   \item How to describe and call an existing function within SSL predicates.
-  \item How to make the result of one function call as the input of another function call.
+  \item How to take the result of one function call as the input of another function call.
   \item How to synthesise programs with inductive predicates without the help of pure theory.
@@
 
-Without any modification upon the implementation, we find the SSL
-predicate within some restrictions can be used to describe function
-relations other than data structures (named function predicates). The
-definition of \textbf{function predicates} is as follows:
+Predicates in separation logic is commonly used to represent some data structures.
+However, as the function is a special case of the relation, it is natural that SSL predicates can describe the functional relation with some restrictions without any extension.
+Here we name such predicates as \textbf{function predicates}.
+The definition of \textbf{function predicates} is as follows:
+
+% Without any modification upon the implementation, we find the SSL
+% predicate within some restrictions can be used to describe function
+% relations other than data structures (named function predicates). The
+% definition of \textbf{function predicates} is as follows:
 
@@
     where \lstinline{T} $\in$ \{\lstinline{loc}, \lstinline{int}\}.
-    The type of \lstinline{xi} (and \lstinline{output}) is decided by
+    The types of \lstinline{xi} and \lstinline{output} are decided by
     the type of $f$. If it is an integer in $f$, then its type is
@@
 
-Since the input of the whole workflow is functional programs, the
-"output" in the definition is to provide another location for the
-output of the function. And the specification to synthesise function
-$f$ should have the following format:
+Since the intent is to represent functional programs, the
+"output" in the predicate is to provide another location for the
+function output. And the specification to synthesise function
+$f$ should have the following format (the case is when the types of first two parameters are \lstinline{int} and \lstinline|sll|):
 
@@
 void f(loc x1, ... ,loc xn)
-{x1 :-> v1 ** x2 :-> l2 ** sll(l2) ** ... ** xn :-> vn ** output :-> 0}
-{x1 :-> v1 ** x2 :-> l2 ** ... ** xn :-> vn ** output :-> output0 ** 
- predf(v1, l2, ..., vn, output0)}
+  {x1 :-> v1 ** x2 :-> l2 ** sll(l2) ** ... ** output :-> 0}
+  {x1 :-> v1 ** x2 :-> l2 ** ... ** output :-> output0 ** predf(v1, l2, ..., vn, output0)}
 \end{lstlisting}
@@
 
-As we show in previous examples, the reason we can have \func structure is that the points-to structure in the post-condition is always eliminated after some write operations.
+As we show in \autoref{sec:examples-fold}, the reason we need \func structure is to represent some modifications which cannot be simply written as \textit{points-to} structure; and the reason we can have \func structure is that the \textit{points-to} structure in the post-condition is always eliminated after some write operations.
 For example, the in-placed \lstinline{inc1} functions specification is satisfied via the \writer operation (\autoref{fig:write}) on the location.
@@
 modified. Thus, we add the new \funcwrite rule into the zoo of SSL
-rules (see \autoref{fig:funcwrite}).
+rules as in \autoref{fig:funcwrite}.
 %
@@
 Furthermore, we have the \func generated from function predicates and with the format defined in \autoref{sec:funcPred}.
-As a result, the equivalent original SSL that duplicates points-to of one location is not a problem, since they can be merged as one.
+As a result, the equivalent original SSL that duplicates \textit{points-to} of one location is not a problem, since they can be merged as one.
 
@@
 
-Though with rich expressiveness, SSL has difficulty in expressing the sequential application of functions.
-For example, given the \func structure available, the following function is not expressible within one function predicate:
+Though with rich expressiveness, SSL still has the difficulty in expressing the sequential application of functions.
+For example, given the \func structure available, the following function is not expressible in the existing \suslik within one function predicate:
 
@@
 However, \lstinline{houtput} is not a location in the pre-condition, which is not allowed in SSL.
-Thus, we introduce a new keyword \lstinline{temp} to denote the temporary location for the sequential application.
+Thus, we introduce another keyword \lstinline{temp} to denote the temporary location for the sequential application.
 The new definition of \lstinline{func} is as follows:
@@
 
-\begin{lstlisting}[language=SynLang]
+\begin{lstlisting}[language=Pika]
     flatten :: [[a]] -> [a]
@@
 Specifically, if the location contains the value with type \lstinline{int}, then the \lstinline{type_free} will do nothing.
-Thus, the function predicate with \lstinline{temp} is much better, in the sense that no extra space is used, and the synthesised function is consistent with the source language.
+Thus, with the function predicate with \lstinline{temp} no extra space is used, and the synthesised function is consistent with the source language.
 
@@
 
-The existing \suslik depends on the set theory to express the pure
+The existing \suslik depends on the pure theories (set, interval or list) to express the pure
 relation. However, it is not trivial to automatically generate the
 pure part of SSL specifications from the functional specifications. To
-see why the set theory is needed, the following simple example shows
+see why the pure theory is needed, the following simple example shows
 the functionality of the set theory, with \lstinline{sll_n} being the
@@
 \begin{lstlisting}[language=SynLang]
+predicate sll_n(loc x) {
+|  x == 0        => { emp }
+|  not (x == 0)  => { [x, 2] ** x :-> v ** (x + 1) :-> nxt ** sll_n(nxt) }
+}
+
 predicate cp(loc x, loc y) {
@@
 }
+
+void copy (loc x, loc y)
+  {x :-> x0 ** sll_n(x0) ** y :-> 0}
+  {x :-> x0 ** y :-> y0 ** cp(x0, y0)}
 \end{lstlisting}
@@
 void copy (loc x, loc y) {
+  let x = *x;
   if (x == 0) {
@@
     let n = *(x + 1);
-    copy(n, y);
+    *x = n;
+    copy(x, y);
     let y01 = *y;
@@
     *(y0 + 1) = y01;
-    let vy = *y0
-    *x = vy;
+    *x = 666;
   }
@@
 \lstinline{x} to \lstinline{y}, we can just change the value of x to
-initial value \lstinline{vy} after \lstinline[language = c]{malloc}.
-This is not the user intent, and the output program is not correct.
+initial value of \lstinline|y| (which is set  to be 666 by default in \suslik) after \lstinline[language = c]{malloc}.
+So the output program is not correct in the sense of the user intent, but correct based on the specification.
 %
-Turns out, the solution is not that difficult: we simply need add a
+Turns out, the solution is not that difficult: we simply need to add a
 new kind of heaplet in the specification language, call
@@
 %
-The only difference of the constant points-to from the original
+The only difference of the \textit{constant points-to} from the original
 \textit{points-to} heaplet is that the value of the location is
 constant, which means that the \writer rule in SSL is not applicable.
+Such read-only heaplet is also consistent with the functional language.
+
+Now with only modifying the \lstinline|sll_n| into the following \lstinline|ssl_c| (c for constant) where the \textit{constant points-to} is denoted by \lstinline|:=>|, the synthesiser will generate the correct program.
+\begin{lstlisting}[language=SynLang]
+predicate sll_c(loc x) {
+|  x == 0        => { emp }
+|  not (x == 0)  => { [x, 2] ** x :=> v ** (x + 1) :=> nxt ** sll_n(nxt) }
+}
+\end{lstlisting}
 

### 2023-09-06 new-paper/soundness.tex
soundness

\subsection{From \tool to SLL Specifications: Soundness of the Translation}
\label{sec:soundness}

We want to show that our abstract machine semantics and our SSL
translation fit together. In particular, our abstract machine
semantics should generate models that satisfy the separation logic
propositions given by our SSL translation.
Figure~\ref{fig:soundness-diagram} gives a high-level overview of how
these pieces fit together. We will give a more specific description of
this in Theorem~\ref{thm:gen-soundness}.

\input{soundness-diagram.tex}

\begin{restatable}[Soundness]{theorem}{soundnessThm} \label{thm:gen-soundness}
  For any well-typed expression $e$, the following holds where $\VarSet = \dom(\EndSigma)$:
  \begin{itemize}
    \item If $\Tsem{e}{\VarSet, r}$ is satisfiable and $(e, \StartSigma, \Start{h}, \mathcal{F}) \step (e', \EndSigma, \End{h'}, \mathcal{F}', r)$\\
          then $\EndPair \models \Tsem{e}{\VarSet, r}$
  \end{itemize}
  That is, given an expression $e$ with a satisfiable SSL translation, any heap machine state that $e$ transitions to (by the abstract machine semantics) will
  be a model for the SSL translation of $e$. See Figure~\ref{fig:soundness-diagram}.
\end{restatable}

\begin{proof}
  See the Appendices
  %~\ref{sec:soundness-proof}
  in the extended version of the paper.
\end{proof}

The fact that, at the top level, we only translate function definitions suggests an additional theorem. We want to specifically
show that any possible function application is sound, in the sense just described. This immediately follows from Theorem~\ref{thm:gen-soundness}.

Abbreviating $\instantiateS{A,B}{f}$ as $f_{A,B}$:

\begin{theorem}[Application soundness] \label{thm:fn-soundness}
  For any well-typed function application $f_{A,B}(e)$, the following should hold where $V = \dom(\EndSigma)$:
  \begin{itemize}
    \item If $\Tsem{f_{A,B}(e)}{\VarSet, r}$ is satisfiable and $(f_{A,B}(e), \StartSigma, \Start{h}, \mathcal{F}) \step (e', \EndSigma, \End{h'}, \mathcal{F}', r)$\\
          then $\EndPair \models \Tsem{f_{A,B}(e)}{\VarSet, r}$
  \end{itemize}
\end{theorem}

\begin{proof}
  This follows immediately from Theorem~\ref{thm:gen-soundness}.
\end{proof}

### 2023-09-06 new-paper/extended-soundness.tex
soundness appendix

\section{Soundness Proof}
\label{sec:soundness-proof}

First, we introduce an operator that will make it more convenient to talk about the conjunction of two SSL propositions:
\[
  (p_1 ; s_1) \otimes (p_2 ; s_2) = (p_1 \land p_2 ; s_1 \sep s_2)
\]

\noindent
Next, we present a lemma regarding this operator that will be used in the soundness proof:
\begin{lemma}[$\otimes$ pairing]\label{thm:otimes-entail}
  If $(\sigma_1, h_1) \models p$ and $(\sigma_2, h_2) \models q$ and $\sigma_1 \subseteq \sigma_2$ and $h_1 \mathbin{\bot} h_2$,\\
  then $(\sigma_2, h_1 \circ h_2) \models p \otimes q$
\end{lemma}

\noindent
Now we proceed to soundness.

\soundnessThm*

\begin{proof}
  Proceeding by induction on $e$:
  \begin{itemize}
    \item $e = i$ for some $i \in \mathbb{Z}$.\\
      We have 
      \begin{itemize}
        \item $\Tsem{i}{V,r} = (v == i ; \emp)$
        \item $v \fresh V$
        \item $v = r$
        \item $\EndSigma = \StartSigma \cup \{ (r,i) \}$
        \item $\End{h'} = \Start{h} = \emptyheap$
      \end{itemize}
      Plugging these in, we find that we want to prove
      \[
        (\StartSigma \cup \{ (r,i) \}, \emptyheap) \models (r == i ; \emp)
      \]
      This immediately follows from the rules of separation logic.

    \item $e = b$ for some $b \in \mathbb{B}$. This case proceeds exactly as the previous case.

    \item $e = e_1 + e_2$.
      From the $\Tsem{\cdot}{}$ hypothesis, we have
      \begin{itemize}
        \item $(e_1, V_0) \tstep (p_1, s_2, V_1, v_1)$
        \item $(e_2, V_1) \tstep (p_2, s_2, V_2, v_2)$
        \item $v \fresh V_2$
        \item $v = r$
        \item $\Tsem{e_1 + e_2}{V, r} = (v == v_1 + v_2 \land p_1 \land p_2 ; s_1 \sep s_2)$
        \item $\Tsem{e_1}{V_0,v_1} = (p_1 ; s_1)$
        \item $\Tsem{e_2}{V_1,v_2} = (p_2 ; s_2)$
        % \item $ $
      \end{itemize}

      Plugging these in, we want to show
      \[
        \EndPair \models (v == v_1 + v_2 \land p_1 \land p_2 ; s_1 \sep s_2)
      \]

      Furthermore, we see that
      \[
        \Tsem{e}{V,r} = (v == v_1 + v_2 ; \emp) \otimes \Tsem{e_1}{V_0,v_1} \otimes \Tsem{e_2}{V_1,v_2}
      \]

      From the {\sc AM} relation hypothesis we have
      \begin{enumerate}
        \item $\EndSigma = \sigma_y \cup \{(r, z)\}$
        \item $z = x' + y'$
        \item $(x, \sigma, \mathcal{F}, h_1) \step (x', \sigma_x, h_1', \mathcal{F}, v_x)$
        \item $(y, \sigma_x, \mathcal{F}, h_2) \step (y', \sigma_y, h_2', \mathcal{F}, v_y)$
        \item $v_1 = v_x$
        \item $v_2 = v_y$
        \item $h = h_1 \circ h_2$
        \item $h' = h_1' \circ h_2'$
      \end{enumerate}

      From the first six items, we can derive
      \[
        (\EndSigma, \emptyheap) \models (v == v_1 + v_2 ; \emp)
      \]

      From the inductive hypotheses, we get
      \begin{itemize}
        \item $(\sigma_x, h_1') \models \Tsem{e_1}{\dom(\sigma_x), v_x}$
        \item $(\sigma_y, h_2') \models \Tsem{e_2}{\dom(\sigma_y), v_y}$
      \end{itemize}

      By Lemma~\ref{thm:otimes-entail}
      \[
        (\sigma_y, h_1' \circ h_2') \models \Tsem{e_1}{\dom(\sigma_x), v_x} \otimes \Tsem{e_2}{\dom(\sigma_y), v_y}
      \]

      and we know that $\sigma_x \subseteq \sigma_y$.

      By Lemma~\ref{thm:otimes-entail}, we conclude
      \[
        (\sigma_y \cup \{(r, z)\}, h') \models (v == v_1 + v_2 ; \emp) \otimes \Tsem{e_1}{\dom(\sigma_x),v_1} \otimes \Tsem{e_2}{\dom(\sigma_y),v_2}
      \]

    \item $e = v$ for some $v \in \Var$.
      Therefore the only two {\sc AM} rules that apply are {\sc AM-Base-Var} and {\sc AM-Loc-Var}. Both
      cases proceed in the same way.

      We know
      \begin{itemize}
        \item $h = h' = \emptyheap$
      \end{itemize}

      So we want to show
      \[
        (\EndSigma, \emptyheap) \models (\verb|true| ; \emp)
      \]
      This trivially holds.

    \item $e = \lowerS{A}{v}$ for some layout $A$ and $v \in \Var$.
      The applicable {\sc AM} rule is {\sc AM-Lower}.
      From the premises of {\sc AM-Lower}, we have
      \begin{itemize}
        \item $h' = \hat{h} \cdot H'$
      \end{itemize}
      where $H'$ comes from applying the layout $A$ to the constructor application expression
      obtained by reducing $e$. As a result, $H'$ exactly fits the specification of
      one of the branches of the $A$ layout.

      $v$ must be associated with some high-level value in $\mathcal{F}$ and $\hat{h}$
      is the part of the heap that is updated when this is reduced. Since all expressions
      are required to be well-typed, $\hat{h}$ must satisfy some collection of heaplets
      in the $A$ layout branch that is associated to $H'$.

      Note that $\Tsem{e}{V,r} = A(v)$

      From these facts, we can conclude
      \[
        \EndPair \models A(v)
      \]

    \item $e = \lowerS{A}{C\; e_1 \cdots e_n}$. This case is similar to the previous case,
      except that we do not need to lookup a variable in the store before dealing with the constructor
      application.

    \item $e = \instantiateS{A,B}{f}(v)$ for some $v \in \Var$.
        The rule {\sc AM-Instantiate} applies here.

        The {\sc S-Inst-Var} rule requires that we satisfy the inductive predicate
        associated to $f$, that is $\mathcal{I}_{A,B}(f)(v)$. Note that the SSL propositions in the definition of
        such an inductive predicate, given by {\sc FnDef}, will always not only explicitly
        describe the memory used in its result (given by the layout $B$) but also the memory used
        in its argument (given by the layout $A$).

        Now, consider that the {\sc AM-Instantiate} rule uses the layout $A$ to the function argument to
        update the heap and uses the $B$ layout to produce the function's result on the heap. This will match the
        inductive predicate associated to $f$ instantiated at the layouts $A$ and $B$, as required.

      %   Among our hypotheses, we have
      % \begin{itemize}
      %   \item $\End{h'} = \hat{h}_1 \circ \hat{h}_2 \circ \cdots \circ \hat{h}_{n+1}$
      %   \item $\EndSigma = \sigma_f$
      %   \item $(A[x]\; (C\; a_1 \cdots a_n) := H) \in \Sigma$
      %   \item $(f\; (C\; b_1 \cdots b_n) := e_f) \in \Sigma$
      %   \item $(\lowerS{B}{e_f[b_1 := r_1]\cdots[b_n := r_n]}, \sigma_{n+1}, \hat{h}_1) \step (e_f', \sigma_f, \hat{h}_2, r)$
      % \end{itemize}
      %
      % Let $\hat{h} = \hat{h}_1 \circ \hat{h}_2 \circ \cdots \circ \hat{h}_n$. Note that we stop at $n$, not $n+1$. This is the
      % difference between $\hat{h}$ and $\End{h'}$.
      %
      % We can recursively apply the present lemma to the last item, since it is a derivation of an abstract machine reduction which is
      % strictly smaller than the abstract machine reduction derivation we started with. From this, we obtain:
      %
      % \[
      %   (\EndSigma, \hat{h}) \models \lowerS{B}{e_f[b_1 := r_1]\cdots[b_n := r_n]}
      % \]

      % and from the {\sc AM} relation hypothesis we have
      % \begin{itemize}
      %   \item $(x, \sigma, \mathcal{F}, h) \step (x', \sigma_x, h, \mathcal{F}, v_x)$
      % \end{itemize}

      \item $e = \instantiateS{A,B}{f}(C\; e_1 \cdots e_n)$.
        This is much like the previous case. The only difference is that the {\sc S-Inst-Constr} rule
        unfolds the inductive predicate $\mathcal{I}_{A,B}(f)(C\; e_1 \cdots e_n)$. The resulting SSL proposition
        will still be satisfied by the model $\EndPair$ given by {\sc AM-Instantiate}.

      \item $e = \instantiateS{B,C}{f}(\instantiateS{A,B}{g}(e_0))$.
        This case combines to instantiates together. The main condition we need to check here is that the
        two instantiate applications use disjoint parts of the heap in $\EndPair$ in {\sc AM-Instantiate}.

        This can be seen to be true by the fact that the resulting heap is built up out of the subexpressions
        using $\circ$, ensuring that the parts of the heap are disjoint.
  \end{itemize}
  % {\color{red} ???}
\end{proof}

### 2023-09-06 new-paper/overview.tex
Describe layouts
soundness appendix

@@
 \generate{} directive. For example, if there is a function definition
-with the type signature \verb|mapAdd1 : List -> List|, a line reading
-\verb|%generate mapAdd1 [Sll] Sll| would instruct the
+with the type signature \lstinline[language=Pika]{mapAdd1 : List -> List}, a line reading
+\lstinline[language=Pika]{$\%$generate mapAdd1 [Sll] Sll} would instruct the
 \Pika{} compiler to generate the \SuSLik{} inductive predicate
@@
 \end{figure}
+
+\input{layouts}
 

### 2023-09-07 new-paper/paper.tex
table comparing to Dargent
soundness
soundness appendix
formatting
abstract

@@
 \usepackage{mathpartir}
+\usepackage{pifont}
+\usepackage{thmtools}
 \lstset{basicstyle=\ttfamily, mathescape=true, literate={~} {$\sim$}{1}}
@@
 This translation also makes interfacing between SSL and \Pika{} easy.
-We introduce ``layouts'' to bridge the gap between a functional programming language and SSL called.
+We introduce ``layouts'' to bridge the gap between functional programming and SSL.
 This allows us to describe the memory layout of algebraic data types as SSL predicates inside \Pika.
 
-  We then develop a second version of \Pika{} extended with an additional synthesis feature. These two versions are compared.
+  We then develop a second version of \Pika{} extended with an additional synthesis feature.
+This synthesis feature allows SSL propositions to be included in type signatures.
+Then, an implementation is synthesized from the type signature using \SuSLik.
+These two versions of \Pika{} are compared.
 
@@
 
+\appendix
+
 \bibliographystyle{ACM-Reference-Format}
 \bibliography{paper}
+
+\input{extended-soundness}
 

### 2023-09-07 new-paper/semantics.tex
soundness
formatting

@@
 
-This subset of \tool does not have guards or \verb|if then else| expressions, but it does have pattern
+This subset of \tool does not have guards or \lstinline[language=Pika]{if then else} expressions, but it does have pattern
 matching. It also has the requirement that functions can only have one argument. Unlike the implementation, there
@@
 rules for a functional language due to the existence of layouts and
-their associated constructors, like \verb|instantiate| and
-\verb|lower|. If an expression is well-typed, then each use of
-\verb|instantiate| and \verb|lower| only uses layouts together with
+their associated constructors, like \lstinline[language=Pika]{instantiate} and
+\lstinline[language=Pika]{lower}. If an expression is well-typed, then each use of
+\lstinline[language=Pika]{instantiate} and \lstinline[language=Pika]{lower} only uses layouts together with
 the ADT that they are defined for.
@@
 specified for the ADT type, it becomes concrete. Base types, like
-\verb|Int|, are also concrete.
+\lstinline[language=Pika]{Int}, are also concrete.
 
@@
 \end{figure}
+
+\input{soundness.tex}

### 2023-09-07 new-paper/examples.tex
bold in table
slight cleanup
a small fix
formatting
examples

@@
 
-Working \Pika{} 1 examples are presented in Figure~\ref{size-comparison}.
+Working \Pika{} 1 examples are presented in Figure~\ref{fig:size-comparison}.
 For each example, we compare the AST size of the \Pika{} code with that of the generated \SuSLik{} code.
@@
 Each example in the list consists of a single function which was tested by hand.
+
+The \verb|filterLt9| example makes use of the Boolean guard feature. Guards must always occur at the level of a function
+definition, rather than inside of a subexpression. On the other hand, the \verb|maximum| example uses the \lstinline[language=Pika]{if-then-else}
+construct. This is necessary since it occurs inside of a \lstinline[language=Pika]{let}. The \verb|leftList| example uses two different
+ADTs: lists and binary trees. A singly-linked list layout is used for the list and a linked representation is used as a layout for the binary tree.
 
@@
   \begin{table}[H]
-  \begin{tabular}{ c c c c }
+  \begin{tabular}{|c|c|c|c|}
     \hline
@@
     \verb|leftList| & 144 & 104 & 0.722 \\
-    \verb|take| & 180 & 112 & 0.622
+    \verb|take| & 180 & 112 & 0.622\\
+    \hline
   \end{tabular}
@@
 
+The \lstinline[language=Pika]{synth} feature is used in the \verb|set| example for two functions: The \verb|append| function mentioned in \autoref{sec:synth} and the
+\verb|setToSll| function that converts set values to linked list values:
+
+\begin{lstlisting}[language=Pika]
+synth setToSll : SetLayout @s -> Sll @s;
+\end{lstlisting}
+
 \begin{figure}
-  \begin{tabular}{ c c c }
+  \begin{tabular}{|c|c|}
     \hline
-    Name & \# of functions & Uses \synth\\
+    Name & \# of functions\\
     \hline
-    \verb|add1Head| & 1 & No\\
-    \verb|anagram| & 4 & No\\
-    \verb|fact| & 1 & No\\
-    \verb|filterLt| & 1 & No\\
-    \verb|heap| & 6 & Yes\\
-    \verb|leftList| & 1 & No\\
-    \verb|mapAdd| & 1 & No\\
-    \verb|maximum| & 1 & No\\
-    \verb|set| & 3 & Yes\\
-    \verb|sum| & 1 & No\\
-    \verb|take| & 1 & No\\
-    \verb|treeSize| & 1 & No\\
+    \verb|add1Head| & 1\\
+    \verb|anagram| & 4\\
+    \verb|fact| & 1\\
+    \verb|filterLt| & 1\\
+    \verb|heap| & 6\\
+    \verb|leftList| & 1\\
+    \verb|mapAdd| & 1\\
+    \verb|maximum| & 1\\
+    \verb|set| & 3\\
+    \verb|sum| & 1\\
+    \verb|take| & 1\\
+    \verb|treeSize| & 1\\
+    \hline
   \end{tabular}

### 2023-09-07 new-paper/limitations.tex
rename section 5; merge sections 5 and 6
British-ise

@@
-\section{Synthesis Limitations And Solutions}
+\section{Adding Synthesis Capabilities to \Pika}
 \label{sec:limitations}
 
-There is a significant limitation to our approach so far: \SuSLik{} will often fail to synthesize code from valid specifications.
+There is a significant limitation to our approach so far: \SuSLik{} will often fail to synthesise code from valid specifications.
 One instance of this problem is the following: In a certain class of functions, \SuSLik{} requires a different number of pointer indirections than it will with other functions.
-We have not found a way to perfectly characterize such functions and, since this will change the way the functions will need to be called, this represents a major impediment to our approach.
-It is worth noting that this is also not the \textit{only} situation in which \SuSLik{} fails to synthesize a program.
+We have not found a way to perfectly characterise such functions and, since this will change the way the functions will need to be called, this represents a major impediment to our approach.
+It is worth noting that this is also not the \textit{only} situation in which \SuSLik{} fails to synthesise a program.
 

### 2023-09-07 new-paper/examples.tex
update table description

@@
 
-Figure~\ref{fig:pika-2-examples} is a table of \Pika{} 2 examples together with the number of functions they have (not including \synth) and whether or not they use \synth.
+Figure~\ref{fig:pika-2-examples} is a table of \Pika{} 2 examples together with the number of functions they have (not including \synth).
 These functions were tested using the unit testing framework built into \Pika{} 2.

### 2023-10-03 writing/new-paper/future-work.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

\section{Future Work}
\label{sec:future-work}

One direction for future work is the extension of \Pika{} 2 to include support for higher-order functions and finishing the implementation of layout polymorphic type checking.

The reverse transformation deserves further investigation: if we go from
an SSL specification to \Pika{} program and then compile to, \eg, C, can
we synthesise additional programs that a traditional SSL synthesisers
would struggle with? What are the limitations of this approach?

Finally, is it possible to derive translations for languages such as \Pika{}
from abstract machine semantics? In this paper, we have given a
language with abstract machine semantics. We then give a translation
of that language into SSL. We then show that the final states given by
the abstract machine semantics are models for the SSL propositions
produced by our translation. But is it possible to begin by specifying the
abstract machine semantics and then mathematically (or automatically)
\textit{derive} an appropriate translation into SSL, with the
requirement that the translation satisfies the soundness theorem?

### 2023-10-03 writing/new-paper/overview.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

--- 2023-09-06 new-paper/overview.tex
@@
 function definition is instantiated at. In this case, we must
-explicitly specify a layout. This is done using the \verb|instantiate|
-and \verb|lower| constructs. The former is for function calls, while
+explicitly specify a layout. This is done using the \lstinline[language=Pika]{instantiate}
+and \lstinline[language=Pika]{lower} constructs. The former is for function calls, while
 the latter is for all other types of values. These operations are
-performed with the \verb|instantiate| and \verb|lower| language
-constructs, respectively. For example, \verb|instantiate [A] B f x|
+performed with the \lstinline[language=Pika]{instantiate} and \lstinline[language=Pika]{lower} language
+constructs, respectively. For example, \lstinline[language=Pika]|instantiate [A] B f x|
 calls $f$ with the argument $x$, where the layout $A$ is used for the
@@
 \Pika{} compiler to generate the \SuSLik{} inductive predicate
-corresponding to \verb|mapAdd1| instantiated to the \verb|Sll| layout
+corresponding to \lstinline[language=Pika]{mapAdd1} instantiated to the \lstinline[language=Pika]{Sll} layout
 for both its argument and its result.
@@
 %
-In these examples, we will often make use of the \verb|List| algebraic
-data type and its \verb|Sll| layout from~\autoref{fig:List-def}. A
+In these examples, we will often make use of the \lstinline[language=Pika]{List} algebraic
+data type and its \lstinline[language=Pika]{Sll} layout from~\autoref{fig:List-def}. A
 simple example of \Pika{} code that illustrates algebraic data types and
@@
 A slightly more complicated example comes from trying to write a
-functional-style \verb|map| function directly in \SuSLik. Consider a
+functional-style \lstinline[language=Pika]{map} function directly in \SuSLik. Consider a
 function which adds 1 to each integer in a list of integers.
@@
 \noindent
-Note that inductive predicates are used for \textit{two} different purposes: the \verb|sll| inductive
-predicate describes a singly-linked list data structure, while the \verb|mapAdd1| inductive predicate 
+Note that inductive predicates are used for \textit{two} different purposes: the \lstinline[language=Pika]{sll} inductive
+predicate describes a singly-linked list data structure, while the \lstinline[language=Pika]{mapAdd1} inductive predicate 
 describes how the input list relates to the output list. Both are used in the specification of
-\verb|mapAdd1_fn|: \verb|sll| in the precondition and \verb|mapAdd1| in the postcondition.
+\lstinline[language=Pika]{mapAdd1_fn}: \lstinline[language=Pika]{sll} in the precondition and \lstinline[language=Pika]{mapAdd1} in the postcondition.
 
-Using the \verb|mapAdd1| inductive predicate gives us two advantages over attempting to
-put the SSL propositions directly into the postcondition of \verb|mapAdd1_fn|:
+Using the \lstinline[language=Pika]{mapAdd1} inductive predicate gives us two advantages over attempting to
+put the SSL propositions directly into the postcondition of \lstinline[language=Pika]{mapAdd1_fn}:
 %
@@
     matching in a language with algebraic data types, but we are examining the pointer involved directly.
-  \item We are able to express \textit{recursion} part of the postcondition: the \verb|mapAdd1| inductive
-    predicate refers to itself in the \verb|not (x == 0)| branch.
+  \item We are able to express \textit{recursion} part of the postcondition: the \lstinline[language=Pika]{mapAdd1} inductive
+    predicate refers to itself in the \lstinline[language=Pika]{not (x == 0)} branch.
 \end{enumerate}
@@
 Say we want to write the functional program that corresponds to this specification. One way to do this in a Haskell-like language
-is the following, using the \verb|List| type from \autoref{fig:List-def}.
+is the following, using the \lstinline[language=Pika]{List} type from \autoref{fig:List-def}.
 
@@
 \noindent
-The only missing information is the memory representation of the \verb|List| data structure. We do not want the
-\verb|mapAdd1_fn| implementation to deal with this directly, however. We want to separate the more abstract notions
+The only missing information is the memory representation of the \lstinline[language=Pika]{List} data structure. We do not want the
+\lstinline[language=Pika]{mapAdd1_fn} implementation to deal with this directly, however. We want to separate the more abstract notions
 of pattern matching and constructors from the concrete memory layout that the data structure has.
@@
 To accomplish this, we now extend the code with the definition of
-\verb|Sll| from \autoref{fig:List-def}. \verb|Sll| is a
-\textit{layout} for the algebraic data type \verb|List|.
+\lstinline[language=Pika]{Sll} from \autoref{fig:List-def}. \lstinline[language=Pika]{Sll} is a
+\textit{layout} for the algebraic data type \lstinline[language=Pika]{List}.
 %
@@
 function definitions. We instruct a \SuSLik{} specification generator to
-generate a \SuSLik{} specification from the \verb|mapAdd1_fn| function
-using the \verb|Sll| layout:
+generate a \SuSLik{} specification from the \lstinline[language=Pika]{mapAdd1_fn} function
+using the \lstinline[language=Pika]{Sll} layout:
 %
@@
 %
-The \verb|[Sll]| part of the directive tells the generator which
+The \lstinline[language=Pika]{[Sll]} part of the directive tells the generator which
 layouts are used for the arguments. In this case, the function only
-has one argument and the \verb|Sll| layout is used. The \verb|Sll| at
+has one argument and the \lstinline[language=Pika]{Sll} layout is used. The \lstinline[language=Pika]{Sll} at
 the end specifies the layout for the result.
@@
 
-We can generalise our \verb|mapAdd1| to map arbitrary \verb|Int|
-functions over a list and then redefine \verb|mapAdd1| using the new
-\verb|map|.
+We can generalise our \lstinline[language=Pika]{mapAdd1} to map arbitrary \lstinline[language=Pika]{Int}
+functions over a list and then redefine \lstinline[language=Pika]{mapAdd1} using the new
+\lstinline[language=Pika]{map}.
 
@@
 This example makes use of \lstinline{instantiate} in two places. In the first case where we have the call
-\lstinline{instantiate [Int] Int f x}, the builtin \verb|Int| layout is used for both the input
-and output. In this special case, the \verb|Int| layout shares a name with the \verb|Int| type
+\lstinline[language=Pika]{instantiate [Int] Int f x}, the builtin \lstinline[language=Pika]{Int} layout is used for both the input
+and output. In this special case, the \lstinline[language=Pika]{Int} layout shares a name with the \lstinline[language=Pika]{Int} type
 that it represents. This is necessary since \lstinline{instantiate} is used for all non-recursive calls that are not constructor applications.
 
-In the second use, \lstinline{instantiate [Int -> Int, Sll] Sll map add1 xs}, we specify that
-the second argument uses the \verb|Sll| layout for the \verb|List| type from \autoref{fig:List-def}. We also give \verb|Sll| as
-the layout for the result of the call. Note that it is not necessary to use \verb|instantiate| for the recursive call to \verb|map|. This is because the appropriate layout
+In the second use, \lstinline[language=Pika]{instantiate [Int -> Int, Sll] Sll map add1 xs}, we specify that
+the second argument uses the \lstinline[language=Pika]{Sll} layout for the \lstinline[language=Pika]{List} type from \autoref{fig:List-def}. We also give \lstinline[language=Pika]{Sll} as
+the layout for the result of the call.
+Note that it is not necessary to use \lstinline[language=Pika]{instantiate} for the recursive call to \lstinline[language=Pika]{map}.
+This is because the appropriate layout
 is inferred for recursive calls.
 
-It might be surprising that \verb|instantiate| is required in the body of \verb|mapAdd1| since the type signature
-of \verb|mapAdd1| suggests that it is layout polymorphic and yet we must pick a specific \verb|List| layout when
-we use \verb|instantiate| to call \verb|map|. This is because, in general, a call inside the body of some function \verb|fn| might use
-any layout, even layouts that have no relation to the layouts that \verb|fn| is instantiated to.
+It might be surprising that \lstinline[language=Pika]{instantiate} is required in the body of \lstinline[language=Pika]{mapAdd1} since the type signature
+of \lstinline[language=Pika]{mapAdd1} suggests that it is layout polymorphic and yet we must pick a specific \lstinline[language=Pika]{List} layout when
+we use \lstinline[language=Pika]{instantiate} to call \lstinline[language=Pika]{map}. This is because, in general, a call inside the body of some function \lstinline[language=Pika]{fn} might use
+any layout, even layouts that have no relation to the layouts that \lstinline[language=Pika]{fn} is instantiated to.
 
 It is possible to do inference of some layouts, for example in
-\verb|mapAdd1| we would usually want to use the same layout as the
+\lstinline[language=Pika]{mapAdd1} we would usually want to use the same layout as the
 argument layout, but we leave this for future work. Another approach
@@
 
-Two more features that are common in functional languages are \verb|let| bindings and
-\verb|if-then-else| expressions. Both of these have straightforward translations into
-\SuSLik. A \verb|let| binding corresponds to introducing a new variable with an equality
-constraint corresponding to the equality in the \verb|let| binding in the pure part of a \SuSLik{} assertion.
+Two more features that are common in functional languages are \lstinline[language=Pika]{let} bindings and
+\lstinline[language=Pika]{if-then-else} expressions. Both of these have straightforward translations into
+\SuSLik. A \lstinline[language=Pika]{let} binding corresponds to introducing a new variable with an equality
+constraint corresponding to the equality in the \lstinline[language=Pika]{let} binding in the pure part of a \SuSLik{} assertion.
 
-The \verb|if-then-else| construct corresponds to \SuSLik's C-like ternary operator. We can use these features together to implement a \verb|maximum| function:
+The \lstinline[language=Pika]{if-then-else} construct corresponds to \SuSLik's C-like ternary operator. We can use these features together to implement a \lstinline[language=Pika]{maximum} function:
 
@@
 \noindent
-The \verb|let| binding is required here. This is because the recursive predicate
+The \lstinline[language=Pika]{let} binding is required here. This is because the recursive predicate
 application must use an additional variable, beyond the argument and output
-parameters, to store the intermediate result \verb|i| and this is introduced in
-\Pika{} using a \verb|let|. It is possible to extend the translator to
-automatically introduce these \verb|let| bindings, but we relegate this to future work.
+parameters, to store the intermediate result \lstinline[language=Pika]{i} and this is introduced in
+\Pika{} using a \lstinline[language=Pika]{let}. It is possible to extend the translator to
+automatically introduce these \lstinline[language=Pika]{let} bindings, but we relegate this to future work.
 
-Furthermore, since the \verb|let| is required and the variable introduced by the \verb|let| is used
-in the \verb|if| condition, we cannot replace the \verb|if| with a guard. Guards can only occur
-at the top-level of a function definition, so it could not occur after a \verb|let|.
+Furthermore, since the \lstinline[language=Pika]{let} is required and the variable introduced by the \lstinline[language=Pika]{let} is used
+in the \lstinline[language=Pika]{if} condition, we cannot replace the \lstinline[language=Pika]{if} with a guard. Guards can only occur
+at the top-level of a function definition, so it could not occur after a \lstinline[language=Pika]{let}.
 
@@
 classic example of such a fold is a fold on a list. In this example, we will write a
-right fold over a \verb|List|.
+right fold over a \lstinline[language=Pika]{List}.
 
-We also demonstrate \verb|Ptr| types and the \verb|addr| builtin operation. This is somewhat similar to an \verb|@| pattern in Haskell,
+We also demonstrate \lstinline[language=Pika]{Ptr} types and the \lstinline[language=Pika]{addr} builtin operation. This is somewhat similar to an \verb|@| pattern in Haskell,
 but it is a low-level construct, indictating how pointers should be used, rather than a high-level construct.
-Given a base type $\alpha$ (such as \verb|Int|), \verb|Ptr Int| is a type. This type is
+Given a base type $\alpha$ (such as \lstinline[language=Pika]{Int}), \lstinline[language=Pika]{Ptr Int} is a type. This type is
 passed around differently in the generated code. In particular, it is passed by reference.
-\verb|addr| is the corresponding value-level introduction operation.
+\lstinline[language=Pika]{addr} is the corresponding value-level introduction operation.
 \begin{lstlisting}[language=Pika]
@@
 \noindent
-The compiler produces the following \SuSLik{} specification for \verb|fold_List|:
+The compiler produces the following \SuSLik{} specification for \lstinline[language=Pika]{fold_List}:
 
@@
 %
-In the generated code, \verb|f| is applied to \verb|p_x| rather than
-\verb|x| as a result of using \verb|Ptr| and \verb|addr|.
+In the generated code, \lstinline[language=Pika]{f} is applied to \lstinline[language=Pika]{p_x} rather than
+\lstinline[language=Pika]{x} as a result of using \lstinline[language=Pika]{Ptr} and \lstinline[language=Pika]{addr}.

### 2023-10-03 writing/new-paper/related-work.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

--- 2023-08-22 new-paper/related-work.tex
@@
 multiple parts of the type (for instance, in a function type
-\verb|List -> List| we have no way at the type-level of requiring that
+\lstinline[language=Pika]{List -> List} we have no way at the type-level of requiring that
 the argument \verb|List| layout and the result \verb|List| layout are
@@
 largely on data structure manipulation, which follow the primary
-motivation of \suslik.
+motivation of \suslik. \autoref{fig:dargent-comparison} has a comparison of some of the major features of \Pika{} and \tname{Dargent}.
+
+\begin{figure}
+  \begin{tabular}{|c|c|c|}
+    \hline
+    Feature & \Pika & \tname{Dargent}\\
+    \hline
+    Bit-level layouts & \xmark & \cmark\\
+    Recursive layouts & \cmark & \xmark\\
+    Synthesis from type signatures & \cmark & \xmark\\
+    Custom record accessors & \xmark & \cmark\\
+    \hline
+  \end{tabular}
+  \caption{Comparison to \tname{Dargent}}
+  \label{fig:dargent-comparison}
+\end{figure}
 

### 2023-10-03 writing/notes-on-pikac/introduction.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

\section{Introduction}

\Pika{} is a functional programming language with layouts and layout
polymorphism. \PikaCore{} is an intermediate representation of \Pika{} where all
layout polymorphism has been removed. That is, in \PikaCore, all layout
polymorphism is specialized to specific layouts and SSL assertions are
first-class values. Furthermore, \PikaCore{} is in ANF form.

From \PikaCore, the program can be translated to several different targets. This
pipeline is given in Fig.~\ref{fig:translation-pipeline}.

\begin{figure}
  \input{pipeline.tex}
  \caption{Translation pipeline}
  \label{fig:translation-pipeline}
\end{figure}

### 2023-10-03 pika.sh
Update pika.sh

@@ -1,3 +1,3 @@
 #!/bin/sh
 
-cabal exec PikaC -- "$@"
+cabal exec PikaC --no-c -- "$@"

### 2024-01-15 src/Bench.hs
benchmark suite

@@ -1,0 +1,14 @@
+module Main
+  where
+
+import PikaC.Benchmark.Benchmark
+import Criterion
+import Criterion.Main
+
+main :: IO ()
+main = do
+  benchmarks <- parseBenchmarks benchmarkConfigPath
+  results <- runBenchmarks benchmarks
+  putStrLn $ toLaTeX results
+  -- defaultMain [runBenchmarks benchmarks]
+

### 2024-01-15 src/PikaC/Backend/SuSLik/SuSLang/Syntax.hs

@@ -5,6 +5,8 @@ module PikaC.Backend.SuSLik.SuSLang.Syntax
 
 import Unbound.Generics.LocallyNameless
 
+import Control.DeepSeq
+
 import GHC.Generics
 import GHC.Stack
 
@@ -52,6 +54,11 @@ data Function =
   , functionBody :: Bind [ExprName] [Command]
   }
   deriving (Show, Generic)
+
+instance NFData SuSType
+instance NFData Command
+instance NFData Expr
+instance NFData Function
 
 instance Alpha Expr
 instance Alpha Command

### 2024-01-15 src/PikaC/Backend/SuSLik/Syntax.hs

@@ -18,6 +18,8 @@ import PikaC.Syntax.Pika.Layout (GhostType (..))
 import Unbound.Generics.LocallyNameless
 import Unbound.Generics.LocallyNameless.Unsafe
 
+import Control.DeepSeq
+
 import GHC.Generics
 import GHC.Stack
 
@@ -94,6 +96,9 @@ data InductivePredicate
     }
   deriving (Show, Generic)
 
+instance NFData InductivePredicate
+instance NFData PredicateBranch
+
 data HeapletS
   = PointsToS (PointsTo Expr)
   | ReadOnlyPointsToS (PointsTo Expr)
@@ -102,6 +107,9 @@ data HeapletS
   | BlockS (Name Expr) Int
   | TempLoc (Name Expr)
   deriving (Show, Generic)
+
+instance NFData HeapletS
+instance NFData Expr
 
 nubBlocks :: [HeapletS] -> [HeapletS]
 nubBlocks = nubBy go
@@ -137,6 +145,8 @@ data FnSig
     }
   deriving (Show, Generic)
 
+instance NFData FnSig
+
 data FnSpec
   = FnSpec
     { _fnSpecPrecond :: [HeapletS]
@@ -145,6 +155,8 @@ data FnSpec
         ,Assertion)
     }
   deriving (Show, Generic)
+
+instance NFData FnSpec
 
 toReadOnlyPointsTo :: HeapletS -> HeapletS
 toReadOnlyPointsTo (PointsToS p) = ReadOnlyPointsToS p

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs

@@ -1,0 +1,225 @@
+{-# LANGUAGE DeriveGeneric #-}
+{-# LANGUAGE RecordWildCards #-}
+{-# LANGUAGE LambdaCase #-}
+
+module PikaC.Benchmark.Benchmark
+  where
+
+import Dhall hiding (map)
+
+import qualified PikaC.Backend.C.Syntax as C
+import PikaC.Backend.SuSLik.Syntax (InductivePredicate, FnSig)
+import PikaC.Backend.SuSLik.SuSLang.Syntax as SuSLang
+import PikaC.Backend.SuSLik.CodeGen
+import PikaC.Backend.SuSLik.Invoke
+import PikaC.Syntax.Pika.Parser
+import PikaC.Syntax.Pika.FnDef
+import qualified PikaC.Syntax.PikaCore.FnDef as PikaCore
+import PikaC.Syntax.ParserUtils
+import PikaC.Stage.ToPikaCore.Monad
+import PikaC.Stage.ToPikaCore.SimplifyM
+import PikaC.Stage.ToPikaCore
+
+import Control.Monad.Identity
+
+import System.FilePath
+import Data.String
+import Text.Printf
+
+import Criterion
+import Criterion.Types
+import Criterion.Main.Options
+import Statistics.Types
+
+import Control.DeepSeq
+
+benchmarkPath :: FilePath
+benchmarkPath = "tests"
+
+benchmarkPathC :: FilePath
+benchmarkPathC = benchmarkPath </> "c"
+
+benchmarkConfigPath :: FilePath
+benchmarkConfigPath = benchmarkPathC </> "config.dhall"
+
+data PikaCode =
+  PikaCode
+  { pikaCodeFile :: FilePath
+  , pikaCodeFileData :: String
+  }
+  deriving (Generic, Show)
+
+data PikaBenchmark a =
+  PikaBenchmark
+  { benchName :: String
+  , benchContents :: a
+  }
+  deriving (Generic, Show, Functor)
+
+type PikaCompileBenchmark = PikaBenchmark PikaCode
+type PikaSynthBenchmark = PikaBenchmark Compiled
+
+data BenchmarkResult =
+  BenchmarkResult
+  { benchResultName :: String
+  , benchResultCompileReport :: Report
+  , benchResultSynthReport :: Report
+  -- , benchResultCTime :: Double
+  }
+  deriving (Show)
+
+instance NFData PikaCode
+instance NFData a => NFData (PikaBenchmark a)
+
+instance FromDhall PikaCode
[…]

### 2024-01-15 tests/c/common/generators.h

@@ -1,0 +1,114 @@
+#ifndef GENERATORS_H
+#define GENERATORS_H
+
+#include "common.h"
+
+void _generateInt(int* p) {
+  static int i = 0;
+  *p = i++;
+}
+
+void _printIntList(loc x)
+{
+  loc curr = READ_LOC(x, 0);
+
+  printf("{");
+
+  while (curr != NULL) {
+    long num = (long)(READ_INT(curr, 0));
+
+    printf("%d", num);
+
+    loc next = (loc)READ_LOC(curr, 1);
+
+    if (next != NULL) {
+      printf(", ");
+    }
+
+    curr = next;
+  }
+
+  printf(" }");
+
+  return;
+}
+
+void _generateIntListHelper(long start, long end, loc x) {
+  loc head = malloc(sizeof(loc));
+
+  loc n = NULL;
+  for (int i = start; i < end; ++i) {
+    if (i == start) {
+      n = malloc(2 * sizeof(loc));
+
+      WRITE_LOC(head, 0, n);
+    }
+
+    WRITE_LOC(n, 0, i);
+
+    if (i < end-1) {
+      loc next = malloc(2 * sizeof(loc));
+      WRITE_LOC(n, 1, next);
+      n = next;
+    } else {
+      WRITE_LOC(n, 1, NULL);
+    }
+  }
+
+  loc r = READ_LOC(head, 0);
+  WRITE_LOC(x, 0, r);
+}
+
+void _generateIntList(loc x) {
+  generateIntListHelper(0, 9, x);
+}
+
+void _generateBinaryTree(int** tree) {
+  // TODO
+}
+
+void _printIntList(loc list) {
+  while (list != NULL) {
+    _printInt(list);
+    list = READ_LOC(list, 1);
+  }
+}
+
+// Peano naturals //
+void printNat(loc x) {
+  loc curr = READ_LOC(x, 0);
+
+  int i = 0;
+  while (curr != NULL) {
+    ++i;
+    curr = (loc)READ_LOC(curr, 0);
+  }
+
+  printf("%d", i);
+}
+
+void generateNat(int num, loc x) {
+  loc head = malloc(sizeof(loc));
+
+  loc n = NULL;
+  for (int i = 0; i < num; ++i) {
+    if (i == 0) {
+      n = malloc(sizeof(loc));
+      WRITE_LOC(head, 0, n);
+    }
+
+    if (i < num-1) {
[…]

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs
fix LaTeX generation

@@ -147,7 +147,7 @@ toLaTeX results =
       "verb|" ++ benchResultName ++ "| & " ++ fromReport benchResultCompileReport ++ " & "  ++ fromReport benchResultSynthReport ++ "\\\\"
 
     cmd :: String -> String
-    cmd s = "\\\\" <> s
+    cmd s = "\\" <> s
 
     fromReport = printf "%.2f" . estPoint . anMean . reportAnalysis
 

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs
Some fixes for LaTeX generation, put benchmark results into paper

@@ -134,7 +134,7 @@ toLaTeX results =
     [cmd "begin{table}"
     ,cmd "begin{tabular}{|c|c|c|}"
     ,cmd "hline"
-    ,cmd "Name & Compilation time & Synthesis time\\\\"
+    ,"Name & Compilation time & Synthesis time\\\\"
     ,cmd "hline"
     ]
     ++ map toRow results ++
@@ -144,12 +144,12 @@ toLaTeX results =
     ]
   where
     toRow BenchmarkResult { .. } =
-      "verb|" ++ benchResultName ++ "| & " ++ fromReport benchResultCompileReport ++ " & "  ++ fromReport benchResultSynthReport ++ "\\\\"
+      cmd "verb|" ++ benchResultName ++ "| & " ++ fromReport benchResultCompileReport ++ " & "  ++ fromReport benchResultSynthReport ++ "\\\\"
 
     cmd :: String -> String
     cmd s = "\\" <> s
 
-    fromReport = printf "%.2f" . estPoint . anMean . reportAnalysis
+    fromReport = printf "%.3f" . estPoint . anMean . reportAnalysis
 
 -- runBenchmarks :: [PikaCompileBenchmark] -> Benchmark
 -- runBenchmarks benchmarks =

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs
More precision in benchmark results; some work on LaTeX generation

@@ -117,7 +117,7 @@ toBenchmark f theBenchmark@PikaBenchmark { .. } =
 runBenchmarks :: [PikaCompileBenchmark] -> IO [BenchmarkResult]
 runBenchmarks benchmarks = do
   let compiledBenchmarks = map compileBenchmark benchmarks
-      synthConfig = defaultConfig { resamples = 3 }
+      synthConfig = defaultConfig { resamples = 2 }
   compileReports <- traverse (benchmarkGo "compile" defaultConfig benchmarkToCriterionCompile) benchmarks
   synthReports <- compiledBenchmarks `deepseq` traverse (benchmarkGo "synthesize" synthConfig benchmarkToCriterionSynth) compiledBenchmarks --(benchmarkWith' synthConfig . benchmarkToCriterionSynth) compiledBenchmarks
 
@@ -131,16 +131,14 @@ runBenchmarks benchmarks = do
 toLaTeX :: [BenchmarkResult] -> String
 toLaTeX results =
   unlines $
-    [cmd "begin{table}"
-    ,cmd "begin{tabular}{|c|c|c|}"
+    [cmd "begin{tabular}{|c|c|c|}"
     ,cmd "hline"
-    ,"Name & Compilation time & Synthesis time\\\\"
+    ,"Name & Compilation (s) time & Synthesis time (s)\\\\"
     ,cmd "hline"
     ]
     ++ map toRow results ++
     [cmd "hline"
     ,cmd "end{tabular}"
-    ,cmd "end{table}"
     ]
   where
     toRow BenchmarkResult { .. } =

### 2024-01-15 src/Bench.hs
Include AST sizes

@@ -10,5 +10,6 @@ main = do
   benchmarks <- parseBenchmarks benchmarkConfigPath
   results <- runBenchmarks benchmarks
   putStrLn $ toLaTeX results
+  -- print results
   -- defaultMain [runBenchmarks benchmarks]
 

### 2024-01-15 src/PikaC/Backend/SuSLik/Syntax.hs

@@ -55,6 +55,26 @@ data Expr
   | SetUnion Expr Expr
   deriving (Show, Generic)
 
+instance Size Expr where
+  size (V x) = size x
+  size (IntLit x) = size x
+  size (BoolLit x) = size x
+  size (Div x y) = visibleNode $ size x + size y
+  size (Mod x y) = visibleNode $ size x + size y
+  size (Add x y) = visibleNode $ size x + size y
+  size (And x y) = visibleNode $ size x + size y
+  size (IfThenElse x y z) = visibleNode $ size x + size y + size z
+  size (Mul x y) = visibleNode $ size x + size y
+  size (Sub x y) = visibleNode $ size x + size y
+  size (Equal x y) = visibleNode $ size x + size y
+  size (Lt x y) = visibleNode $ size x + size y
+  size (Le x y) = visibleNode $ size x + size y
+  size (Not x) = visibleNode $ size x
+  size EmptySet = 1
+  size (SingletonSet x) = visibleNode $ size x
+  size (SetUnion x y) = visibleNode $ size x + size y
+
+
 getV :: HasCallStack => Expr -> ExprName
 getV (V x) = x
 
@@ -96,6 +116,10 @@ data InductivePredicate
     }
   deriving (Show, Generic)
 
+instance Size InductivePredicate where
+  size (InductivePredicate a b c d e) =
+    visibleNode $ size a + size b + size c + size d + size e
+
 instance NFData InductivePredicate
 instance NFData PredicateBranch
 
@@ -107,6 +131,14 @@ data HeapletS
   | BlockS (Name Expr) Int
   | TempLoc (Name Expr)
   deriving (Show, Generic)
+
+instance Size HeapletS where
+  size (PointsToS x) = size x
+  size (ReadOnlyPointsToS x) = size x
+  size (ApplyS x y) = visibleNode $ size x + size y
+  size (RecApply x y) = visibleNode $ size x + size y
+  size (BlockS x y) = visibleNode $ size x + size y
+  size (TempLoc x) = size x
 
 instance NFData HeapletS
 instance NFData Expr
@@ -131,6 +163,9 @@ data PredicateBranch
     }
   deriving (Show, Generic)
 
+instance Size PredicateBranch where
+  size (PredicateBranch x y z) = visibleNode $ size x + size y + size z
+
 data FnSig
   = FnSig
     { _fnSigName :: String
@@ -145,6 +180,9 @@ data FnSig
[…]

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs

@@ -19,12 +19,15 @@ import PikaC.Syntax.ParserUtils
 import PikaC.Stage.ToPikaCore.Monad
 import PikaC.Stage.ToPikaCore.SimplifyM
 import PikaC.Stage.ToPikaCore
+import PikaC.Utils
 
 import Control.Monad.Identity
 
 import System.FilePath
 import Data.String
 import Text.Printf
+
+import Data.List
 
 import Criterion
 import Criterion.Types
@@ -64,6 +67,8 @@ data BenchmarkResult =
   { benchResultName :: String
   , benchResultCompileReport :: Report
   , benchResultSynthReport :: Report
+  , benchResultCompileAstSize :: Int
+  , benchResultSynthAstSize :: Int
   -- , benchResultCTime :: Double
   }
   deriving (Show)
@@ -117,12 +122,18 @@ toBenchmark f theBenchmark@PikaBenchmark { .. } =
 runBenchmarks :: [PikaCompileBenchmark] -> IO [BenchmarkResult]
 runBenchmarks benchmarks = do
   let compiledBenchmarks = map compileBenchmark benchmarks
-      synthConfig = defaultConfig { resamples = 2 }
+      synthConfig = defaultConfig { resamples = 1 }
+
+      parsed :: [PikaModule]
+      parsed = map toParsed  benchmarks
+
+      compiled :: [Compiled]
+      compiled = map benchContents compiledBenchmarks
+
   compileReports <- traverse (benchmarkGo "compile" defaultConfig benchmarkToCriterionCompile) benchmarks
-  synthReports <- compiledBenchmarks `deepseq` traverse (benchmarkGo "synthesize" synthConfig benchmarkToCriterionSynth) compiledBenchmarks --(benchmarkWith' synthConfig . benchmarkToCriterionSynth) compiledBenchmarks
+  synthReports <- compiledBenchmarks `deepseq` traverse (benchmarkGo "synthesize" synthConfig benchmarkToCriterionSynth) compiledBenchmarks
 
-  pure $ zipWith3 BenchmarkResult (map benchName benchmarks) compileReports synthReports
-
+  pure $ zipWith5 BenchmarkResult (map benchName benchmarks) compileReports synthReports (map size parsed) (map size compiled)
   where
     benchmarkGo prefix config f x = do
[…]

### 2024-01-15 src/PikaC/Syntax/Heaplet.hs

@@ -51,6 +51,9 @@ instance Ppr a => Ppr (LayoutArg a) where
 data PointsTo a = Loc a :-> a
   deriving (Show, Generic, Eq, Ord)
 
+instance Size a => Size (PointsTo a) where
+  size (x :-> y) = visibleNode $ size x + size y
+
 instance Subst a b => Subst a (PointsTo b)
 instance Subst a b => Subst a (Loc b)
 
@@ -65,6 +68,9 @@ pointsToRhsLens =
 -- data Loc = LocName :+ Int
 data Loc a = a :+ Int
   deriving (Show, Functor, Eq, Ord, Generic)
+
+instance Size a => Size (Loc a) where
+  size (x :+ y) = visibleNode $ size x + size y
 
 mapPointsTo :: (a -> b) -> PointsTo a -> PointsTo b
 mapPointsTo f ((x :+ i) :-> y) = (f x :+ i) :-> f y

### 2024-01-15 src/PikaC/Syntax/Pika/Expr.hs

@@ -71,6 +71,29 @@ data Expr
   | SetUnion Expr Expr
   deriving (Show, Generic)
 
+instance Size Expr where
+  size (V _) = 1
+  size (IntLit _) = 1
+  size (BoolLit _) = 1
+  size (LayoutLambda x y) = visibleNode $ size x + size y
+  size (ApplyLayout x y) = visibleNode $ size x + size y
+  size (App f xs) = visibleNode $ size f + size xs
+  size (Div x y) = visibleNode $ size x + size y
+  size (Mod x y) = visibleNode $ size x + size y
+  size (Add x y) = visibleNode $ size x + size y
+  size (And x y) = visibleNode $ size x + size y
+  size (IfThenElse x y z) = visibleNode $ size x + size y + size z
+  size (Mul x y) = visibleNode $ size x + size y
+  size (Sub x y) = visibleNode $ size x + size y
+  size (Equal x y) = visibleNode $ size x + size y
+  size (Lt x y) = visibleNode $ size x + size y
+  size (Le x y) = visibleNode $ size x + size y
+  size (Not x) = visibleNode $ size x
+  size EmptySet = 1
+  size (SingletonSet x) = visibleNode $ size x
+  size (SetUnion x y) = visibleNode $ size x + size y
+  
+
 data Test a =
   Test
   { _testName :: String
@@ -78,6 +101,9 @@ data Test a =
   , _testResultType :: Type
   }
   deriving (Show, Generic)
+
+-- NOTE: We completely exclude test directives from the AST node count
+instance Size (Test a) where size _ = 0
 
 instance NFData Expr
 instance NFData a => NFData (Test a)

### 2024-01-15 src/PikaC/Syntax/Pika/FnDef.hs

@@ -39,6 +39,9 @@ data FnDef' f =
     }
   deriving (Generic)
 
+instance Size (f [FnDefBranch]) => Size (FnDef' f) where
+  size (FnDef x y) = visibleNode $ size x + size y
+
 overTypedBranches :: (f [FnDefBranch] -> g [FnDefBranch]) -> FnDef' f -> FnDef' g
 overTypedBranches f (FnDef x y) = FnDef x (f y)
 
@@ -52,6 +55,9 @@ data GuardedExpr =
     Expr -- Body
   deriving (Show, Generic)
 
+instance Size GuardedExpr where
+  size (GuardedExpr x y) = visibleNode $ size x + size y
+
 newtype FnDefBranch =
   FnDefBranch
   { fnBranchMatch :: PatternMatches Expr GuardedExpr
@@ -60,6 +66,9 @@ newtype FnDefBranch =
     -- , fnBranchBody :: Bind [ExprName] Expr
     -- }
   deriving (Show, Generic)
+
+instance Size FnDefBranch where
+  size (FnDefBranch x) = size x
 
 getFnTypeSig :: FnDef -> (String, Type)
 getFnTypeSig fnDef = (fnDefName fnDef, typePairType $ fnDefTypedBranches fnDef)
@@ -72,6 +81,9 @@ data Synth =
     [Type]   -- | Argument types
     Type     -- | Result type
   deriving (Show, Generic)
+
+instance Size Synth where
+  size (Synth a b c d) = visibleNode $ size a + size b + size c + size d
 
 getSynthTypeSig :: Synth -> (String, Type)
 getSynthTypeSig (Synth f _ argTypes resultType) =

### 2024-01-15 src/PikaC/Syntax/Pika/Layout.hs

@@ -86,6 +86,11 @@ data GhostType = SetGhost | IntGhost
 data Ghost a = Ghost GhostType (Name a)
   deriving (Show, Generic)
 
+instance Size (Ghost a) where
+  size (Ghost x y) = size x + size y
+
+instance Size GhostType where size _ = 1
+
 getGhostType :: Ghost a -> GhostType
 getGhostType (Ghost t _) = t
 
@@ -104,6 +109,9 @@ instance NFData a => NFData (Ghost a)
 data GhostCondition a b =
   GhostCondition { _ghostCond :: (Maybe a), _ghostCondBody :: b }
   deriving (Show, Generic)
+
+instance (Size a, Size b) => Size (GhostCondition a b) where
+  size (GhostCondition x y) = visibleNode $ size x + size y
 
 instance (Typeable a, Alpha a, Typeable b, Alpha b) => Alpha (GhostCondition a b)
 
@@ -142,12 +150,21 @@ instance NFData a => NFData (LayoutBody a)
 instance NFData a => NFData (LayoutHeaplet a)
 instance NFData Mode
 
+instance Size a => Size (Layout a) where
+  size (Layout _ y z) = visibleNode $ 1 + size y + size z
+
+instance Size a => Size (LayoutBranch a) where
+  size (LayoutBranch x) = visibleNode $ size x
+
 
 -- | A layout where the layout parameters have been substituted
 type OpenedLayout a = [LayoutBranch a]
 
 data Moded' (s :: Stage) a = Moded' (XModed s) Mode a
   deriving (Generic, Functor)
+
+instance (Size (XModed s), Size a) => Size (Moded' s a) where
+  size (Moded' x y z) = visibleNode $ size x + size y + size z
 
 deriving instance (Show (XModed s), Show a) => Show (Moded' s a)
 
@@ -164,6 +181,8 @@ type ModedName a = Moded (Name a)
 
 data Mode = In | Out
   deriving (Show, Generic, Eq)
+
+instance Size Mode where size _ = 1
 
 type ModeEnv a = [ModedName a]
 
@@ -200,6 +219,9 @@ instance Alpha Mode
 newtype Exists a = Exists { getExists :: ModedName a }
   deriving (Show, Generic)
 
+instance Size a => Size (Exists a) where
+  size (Exists x) = visibleNode $ size x
+
 instance (Typeable a) => Alpha (Exists a)
[…]

### 2024-01-15 src/PikaC/Syntax/Pika/Parser.hs

@@ -2,6 +2,7 @@
 {-# LANGUAGE TemplateHaskell #-}
 {-# LANGUAGE TypeApplications #-}
 {-# LANGUAGE UndecidableInstances #-}
+{-# LANGUAGE RecordWildCards #-}
 
 module PikaC.Syntax.Pika.Parser
   where
@@ -61,6 +62,10 @@ data PikaModule' f =
   deriving (Generic)
 
 deriving instance Show (f [FnDefBranch]) => Show (PikaModule' f)
+
+instance Size (FnDef' f) => Size (PikaModule' f) where
+  size PikaModule { .. } =
+    sizeList moduleAdts + size moduleLayouts + sizeList moduleFnDefs + size moduleSynths + size moduleGenerates + size moduleTests
 
 type PikaModule = PikaModule' TypeSig'
 type PikaModuleElaborated = PikaModule' Typed

### 2024-01-15 src/PikaC/Syntax/Pika/Pattern.hs

@@ -40,6 +40,10 @@ data Pattern a
   | Pattern String [Name a]
   deriving (Show, Generic)
 
+instance Size (Pattern a) where
+  size (PatternVar _) = 1
+  size (Pattern _ xs) = 1 + size xs
+
 instance NFData a => NFData (Pattern a)
 
 makeLenses ''Pattern
@@ -66,6 +70,12 @@ newtype PatternMatch a b =
 newtype PatternMatches a b =
   PatternMatches { getPatternMatches :: Bind [Pattern a] b }
   deriving (Show, Generic)
+
+instance Size b => Size (PatternMatch a b) where
+  size (PatternMatch m) = size m
+
+instance Size b => Size (PatternMatches a b) where
+  size (PatternMatches xs) = size xs
 
 instance (NFData a, NFData b) => NFData (PatternMatch a b)
 instance (NFData a, NFData b) => NFData (PatternMatches a b)

### 2024-01-15 src/PikaC/Syntax/Type.hs

@@ -49,6 +49,19 @@ data Type
   | PlaceholderVar TypeName -- These should be replaced by actual TyVars or LayoutIds
   deriving (Show, Generic)
 
+instance Size Type where
+  size IntType = 1
+  size BoolType = 1
+  size (FnType x y) = visibleNode $ size x + size y
+  size (TyVar _) = 1
+  size (LayoutId _) = 1
+  size (ForAll bnd) = visibleNode $ size bnd
+  size (GhostApp t xs) = visibleNode $ size t + length xs
+  size (PlaceholderVar _) = 1
+
+instance Size AdtName where
+  size _ = 1
+
 type TypeName = Name Type
 
 newtype AdtName = AdtName { unAdtName :: String }
@@ -116,6 +129,9 @@ data TypeSig' a =
   }
   deriving (Show, Generic)
 
+instance Size a => Size (TypeSig' a) where
+  size (TypeSig x) = visibleNode $ size x
+
 type TypeSig = TypeSig' ()
 
 data ConstrainedType =
@@ -124,6 +140,9 @@ data ConstrainedType =
   , _ctypeTy :: Type
   }
   deriving (Show, Generic)
+
+instance Size ConstrainedType where
+  size (ConstrainedType x y) = visibleNode $ size x + size y
 
 instance TypePair TypeSig' where
   type TypePairType TypeSig' = TypeSig
@@ -230,12 +249,18 @@ data Adt =
   }
   deriving (Show, Generic)
 
+instance Size Adt where
+  size (Adt _ xs) = visibleNode $ 1 + sizeList xs
+
 data Constructor =
   Constructor
   { _constructorName :: String
   , _constructorType :: Type
   }
   deriving (Show, Generic)
+
+instance Size Constructor where
+  size (Constructor _ t) = visibleNode $ 1 + size t
 
 -- | Build the appropriate layout polymorphic type, assuming
 -- that one layout is used for each type.
@@ -289,6 +314,9 @@ instance Alpha AdtName
 
 data LayoutConstraint = TypeName :~ AdtName
   deriving (Show, Generic)
+
+instance Size LayoutConstraint where
+  size (x :~ y) = visibleNode $ size x + size y
 
 -- instance Bound LayoutTypeArg where
 

### 2024-01-15 src/PikaC/Utils.hs

@@ -341,6 +341,44 @@ discardM = discard --pure undefined `suchThat` const False
 isConstructor :: String -> Bool
 isConstructor = isUpper . head
 
+class Size a where
+  size :: a -> Int
+
+instance (Size a, Size b) => Size (a, b) where
+  size (x, y) = size x + size y
+
+instance (Size a, Size b) => Size (Bind a b) where
+  size (B x y) = size x + size y
+
+instance Size (Name a) where
+  size _ = 1
+
+instance Size a => Size (Embed a) where
+  size _ = 1
+
+instance Size [Char] where
+  size _ = 1
+
+instance {-# OVERLAPPABLE #-} Size a => Size [a] where
+  size = sum . map size
+
+instance Size () where size _ = 1
+
+instance Size a => Size (Maybe a) where
+  size Nothing = 1
+  size (Just x) = size x
+
+instance Size Int where size _ = 1
+instance Size Bool where size _ = 1
+
+-- | Use in @size@ definition when the node is visible in the AST when it's
+-- printed out
+visibleNode :: Int -> Int
+visibleNode x = x + 1
+
+sizeList :: Size a => [a] -> Int
+sizeList = sum . map size
+
 -- deriving instance (Data a, Data b) => Data (Bind a b)
 -- deriving instance Data a => Data (Name a)
 -- instance (Data a, Data b) => Plated (Bind a b)

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs
fix format in generated LaTeX table

@@ -142,7 +142,7 @@ runBenchmarks benchmarks = do
 toLaTeX :: [BenchmarkResult] -> String
 toLaTeX results =
   unlines $
-    [cmd "begin{tabular}{|c|c|c|}"
+    [cmd "begin{tabular}{|c|c|c|c|c|c|}"
     ,cmd "hline"
     ,"Name & Pika AST size & SuSLik AST size & Pika AST size / SuSLik AST size & Compilation (s) time & Synthesis time (s)\\\\"
     ,cmd "hline"

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs
fix column name

@@ -144,7 +144,7 @@ toLaTeX results =
   unlines $
     [cmd "begin{tabular}{|c|c|c|c|c|c|}"
     ,cmd "hline"
-    ,"Name & Pika AST size & SuSLik AST size & Pika AST size / SuSLik AST size & Compilation (s) time & Synthesis time (s)\\\\"
+    ,"Name & Pika AST size & SuSLik AST size & Pika AST size / SuSLik AST size & Compilation time (s) & Synthesis time (s)\\\\"
     ,cmd "hline"
     ]
     ++ map toRow results ++

### 2024-01-15 src/Main.hs
remove some unneeded equalities from translator output

@@ -223,7 +223,7 @@ main = do
         -- mapM_ (putStrLn . ppr') (moduleLayouts pikaModule)
 
         let checkEnv = mkCheckEnv pikaModule'
-        -- print $ moduleAdts pikaModule'
+        print pikaModule'
 
         -- let (_:firstFnDef:_) = moduleFnDefs pikaModule'
         -- case elaborateFnDef checkEnv firstFnDef of

### 2024-01-15 src/PikaC/Backend/SuSLik/CodeGen.hs

@@ -169,7 +169,7 @@ codeGenLayout useGhosts layout = runFreshM $ do
           then params' ++ map (convertName . getGhostName) ghosts
           else params'
 
-  pure $ SuSLik.InductivePredicate
+  pure $ reduceEqualities $ SuSLik.InductivePredicate
     { SuSLik._indPredName = _layoutName layout
     , SuSLik._indPredArgTypes = map (const (LayoutId "Unused")) params
     , SuSLik._indPredResultType = LayoutId "Unused2" --resultType
@@ -305,9 +305,9 @@ codeGenIndPred fnDef0 = runFreshM $ do
   --   else mapM (genBranch fnName outSizes nonBaseParams unmodedOutParams) branches
 
 
-  pure 
+  pure $ reduceEqualities $
     -- $ trace ("fnName = " ++ fnName ++ ", argTypes = " ++ show argTypes)
-    $ SuSLik.InductivePredicate
+    SuSLik.InductivePredicate
     { SuSLik._indPredName = fnName
     , SuSLik._indPredArgTypes = argTypes ++ [resultType] --[LayoutId "Unused"] -- TODO: We need a way to deal with layouts that have multiple parameters
     , SuSLik._indPredResultType = resultType
@@ -663,6 +663,39 @@ getEquals v e = v : eqNames
     go (SuSLik.And x y) = go x *> go y
     go (SuSLik.Equal (SuSLik.V x) (SuSLik.V y)) = equate x y
     go _ = pure ()
+
+-- | Eliminated some unused equalities
+reduceEqualities :: SuSLik.InductivePredicate -> SuSLik.InductivePredicate
+reduceEqualities indPred =
+  let (_, branches) = SuSLik._indPredBody indPred
+  in
+  indPred
+    { SuSLik._indPredBody = (params, map go branches)
+    }
+  where
+    (params, _) = SuSLik._indPredBody indPred
+
+    go :: SuSLik.PredicateBranch -> SuSLik.PredicateBranch
+    go branch =
+      let vars = params ++ toListOf fv (SuSLik._predBranchCond branch)
+                        ++ toListOf fv (SuSLik._predBranchAssertion branch)
+      in
+      branch
+        { SuSLik._predBranchPure = SuSLik.mkSuSLikEqns $ overEqualities vars $ SuSLik.collectSuSLikEqns $ SuSLik._predBranchPure branch
+        }
+
[…]

### 2024-01-15 src/PikaC/Benchmark/Benchmark.hs

@@ -144,7 +144,7 @@ toLaTeX results =
   unlines $
     [cmd "begin{tabular}{|c|c|c|c|c|c|}"
     ,cmd "hline"
-    ,"Name & Pika AST size & SuSLik AST size & Pika AST size / SuSLik AST size & Compilation time (s) & Synthesis time (s)\\\\"
+    ,"Name & Pika AST & SuSLik AST & Pika AST / SuSLik AST & Compilation time (s) & Synthesis time (s)\\\\"
     ,cmd "hline"
     ]
     ++ map toRow results ++
@@ -162,19 +162,6 @@ toLaTeX results =
     cmd s = "\\" <> s
 
     fromReport = printf "%.3f" . estPoint . anMean . reportAnalysis
-
--- runBenchmarks :: [PikaCompileBenchmark] -> Benchmark
--- runBenchmarks benchmarks =
---   let compiledBenchmarks = map compileBenchmark benchmarks
---   in
---   bgroup "main"
---     [bgroup "Compilation"
---       (map (toBenchmark benchmarkToCriterionCompile) benchmarks)
---
---     ,compiledBenchmarks `deepseq`
---       bgroup "Synthesis"
---         (map (toBenchmark benchmarkToCriterionSynth) compiledBenchmarks)
---     ]
 
 -- compileC :: [String] -> String -> C.CFunction -> FilePath -> IO ()
 -- compileC inputGenerators outputPrinter fn execFileName = undefined

### 2024-01-16 src/Bench.hs
generated C benchmarks

@@ -5,10 +5,32 @@ import PikaC.Benchmark.Benchmark
 import Criterion
 import Criterion.Main
 
+import Control.DeepSeq
+
+import Data.Maybe (catMaybes)
+
 main :: IO ()
 main = do
   benchmarks <- parseBenchmarks benchmarkConfigPath
-  results <- runBenchmarks benchmarks
+
+  putStrLn "compiling benchmarks..."
+
+  let compiledBenchmarks0 = map compileBenchmark benchmarks
+
+  compiledBenchmarks0 `deepseq` putStrLn "...benchmarks compiled."
+
+  putStrLn "synthing benchmarks..."
+  synthedBenchmarks <- traverse synthBenchmark compiledBenchmarks0
+  putStrLn "...benchmarks synthed."
+
+  putStrLn "generating C benchmarks..."
+  cbenchResults <- runCBenchmarks (catMaybes synthedBenchmarks)
+  putStrLn "...C benchmarks generated."
+
+  putStrLn $ cBenchmarkToLaTeX $ cbenchResults
+
+
+  (compiledBenchmarks, results) <- runBenchmarks benchmarks
   putStrLn $ toLaTeX results
   -- print results
   -- defaultMain [runBenchmarks benchmarks]

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs

@@ -1,6 +1,7 @@
 {-# LANGUAGE DeriveGeneric #-}
 {-# LANGUAGE RecordWildCards #-}
 {-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE UndecidableInstances #-}
 
 module PikaC.Benchmark.Benchmark
   where
@@ -12,6 +13,7 @@ import PikaC.Backend.SuSLik.Syntax (InductivePredicate, FnSig)
 import PikaC.Backend.SuSLik.SuSLang.Syntax as SuSLang
 import PikaC.Backend.SuSLik.CodeGen
 import PikaC.Backend.SuSLik.Invoke
+import PikaC.Backend.SuSLik.SuSLang.ToC (functionToC)
 import PikaC.Syntax.Pika.Parser
 import PikaC.Syntax.Pika.FnDef
 import qualified PikaC.Syntax.PikaCore.FnDef as PikaCore
@@ -20,10 +22,22 @@ import PikaC.Stage.ToPikaCore.Monad
 import PikaC.Stage.ToPikaCore.SimplifyM
 import PikaC.Stage.ToPikaCore
 import PikaC.Utils
+import PikaC.Ppr
+
+import Data.Maybe (catMaybes)
 
 import Control.Monad.Identity
+import Control.Monad
+import Control.Arrow (first, second, (***), (&&&))
+import Data.Tuple.Extra (firstM)
+
+import System.Process
+import System.Exit
+import Control.Exception
+import System.Directory (removeFile)
 
 import System.FilePath
+import System.IO
 import Data.String
 import Text.Printf
 
@@ -45,10 +59,49 @@ benchmarkPathC = benchmarkPath </> "c"
 benchmarkConfigPath :: FilePath
 benchmarkConfigPath = benchmarkPathC </> "config.dhall"
 
+cCompiler :: FilePath
+cCompiler = "gcc"
+
+includePath :: FilePath
+includePath = "tests/c"
+
+haskellCompiler :: FilePath
+haskellCompiler = "ghc"
+
+data CType = CInt | CPtr String | CNoPtr String
+  deriving (Generic, Show)
+
+instance FromDhall CType
+
+data CTest =
+  CTest
+  { haskellFile :: FilePath
+  , inputGenerators :: [CType]
+  , outputPrinter :: CType
+  }
+  deriving (Generic, Show)
+
+data PikaTest' f =
+  PikaTest
+  { fileName :: FilePath
[…]

### 2024-01-16 test.c

@@ -1,0 +1,31 @@
+#include "common/common.h"
+#include "common/generators.h"
+
+void filter(int i1, loc x, loc x2) {
+ if (x == 0) {
+  WRITE_LOC(x2, 0, 0);
+ } else {
+  loc h = READ_LOC(x, 0);
+  loc n = READ_LOC(x, 1);
+  if (i1 <= h) {
+   filter(i1, n, x2);
+   loc x4 = READ_LOC(x2, 0);
+   loc x = (loc)malloc(2 * sizeof(loc));
+   WRITE_LOC(x2, 0, x);
+   WRITE_LOC(x, 1, x4);
+   WRITE_LOC(x, 0, h);
+  } else {
+   filter(i1, n, x2);
+  }
+ }
+}
+
+int main() {
+  loc _out = malloc(sizeof(loc));
+  int _x0 = 0;
+  loc _x1 = 0;
+  _x0 = 7;
+  _x1 = _generateIntList();
+  filter(_x0, _x1, _out);
+  return 0;
+}

### 2024-01-16 tests/c/common/generators.h

@@ -3,75 +3,113 @@
 
 #include "common.h"
 
-void _generateInt(int* p) {
+loc _generateInt() {
   static int i = 0;
-  *p = i++;
+  loc p = (loc)malloc(sizeof(loc));
+  WRITE_INT(p, 0, i);
+  return p;
 }
 
 void _printIntList(loc x)
 {
-  loc curr = READ_LOC(x, 0);
+    if (x == 0) return;
 
-  printf("{");
+    printf("[");
 
-  while (curr != NULL) {
-    long num = (long)(READ_INT(curr, 0));
+    loc curr = x;  // Start from the head of the list
 
-    printf("%d", num);
+    while (curr != NULL) {
+        long num = READ_INT(curr, 0);  // Read the integer value
 
-    loc next = (loc)READ_LOC(curr, 1);
+        printf("%ld", num);  // Print the integer value
 
-    if (next != NULL) {
-      printf(", ");
+        curr = (loc)READ_LOC(curr, 1);  // Move to the next node
+
+        if (curr != NULL) {
+            printf(",");
+        }
     }
 
-    curr = next;
-  }
+    printf("]");
 
-  printf(" }");
-
-  return;
+    return;
 }
 
-void _generateIntListHelper(long start, long end, loc x) {
-  loc head = malloc(sizeof(loc));
+// void _printIntList(loc x)
+// {
+//   if (x == 0) return;
+//
+//   loc curr = READ_LOC(x, 0);
+//
+//   printf("{");
+//
+//   while (curr != NULL) {
+//     long num = (long)(READ_INT(curr, 0));
+//
+//     printf("%d", num);
+//
+//     loc next = (loc)READ_LOC(curr, 1);
+//
+//     if (next != NULL) {
+//       printf(", ");
+//     }
+//
+//     curr = next;
+//   }
+//
+//   printf(" }");
+//
+//   return;
+// }
 
-  loc n = NULL;
-  for (int i = start; i < end; ++i) {
-    if (i == start) {
-      n = malloc(2 * sizeof(loc));
+#define LIST_MAX 10000
 
-      WRITE_LOC(head, 0, n);
-    }
+loc _generateIntListHelper(int i, int len) {
+  if (len == 0) return NULL;
 
-    WRITE_LOC(n, 0, i);
-
-    if (i < end-1) {
[…]

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -1,0 +1,12 @@
+theList :: [Int]
+theList = [1..100000]
+
+filterLt :: Int -> [Int] -> [Int]
+filterLt n [] = []
+filterLt n (x:xs)
+  | x < n = filterLt n xs
+  | otherwise = x : filterLt n xs
+
+main :: IO ()
+main = print $ filterLt 7 theList
+

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
More C benchmarks

@@ -301,19 +301,37 @@ benchC name inputGenerators outputPrinter fn haskellCodeFile =
           hPutStrLn cCodeHandle code
           hClose cCodeHandle
 
-          putStrLn code
+          bracket (openTempFile "temp" "c-out.txt")
+            (\(cOutName, cOutHandle) -> do
+              hClose cOutHandle
+              removeFile cOutName)
+            (\(cOutName, cOutHandle) ->
+              bracket (openTempFile "temp" "haskell-out.txt")
+                (\(haskellOutName, haskellOutHandle) -> do
+                  hClose haskellOutHandle
+                  removeFile haskellOutName)
+                (\(haskellOutName, haskellOutHandle) -> do
 
-          systemEchoed $ cCompiler ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
-          systemEchoed $ haskellCompiler ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
+                  putStrLn code
 
-          cReport <- benchmark' $ nfIO $ systemEchoed execTempName
-          haskellReport <- benchmark' $ nfIO $ systemEchoed execHaskellTempName
+                  systemQuiet $ cCompiler ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
+                  systemQuiet $ haskellCompiler ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
 
-          pure $ CBenchmarkResult
-            { cbenchResultName = name
-            , cbenchResultCTime = cReport
-            , cbenchResultHaskellTime = haskellReport
-            })
+                  cReport <- benchmark' $ nfIO $ systemVeryQuiet $ execTempName -- ++ " > " ++ cOutName
+                  haskellReport <- benchmark' $ nfIO $ systemVeryQuiet $ execHaskellTempName -- ++ " > " ++ haskellOutName
+
+                  cOut <- hGetContents cOutHandle
+                  haskellOut <- hGetContents haskellOutHandle
+
+                  when (cOut /= haskellOut) $ do
+                    putStrLn $ "ERROR: Benchmark results differ between C and Haskell."
+                    exitFailure
+
+                  pure $ CBenchmarkResult
+                    { cbenchResultName = name
+                    , cbenchResultCTime = cReport
[…]

### 2024-01-16 test.c

@@ -1,31 +1,27 @@
 #include "common/common.h"
 #include "common/generators.h"
 
-void filter(int i1, loc x, loc x2) {
+void treeSize(loc x, loc i1) {
  if (x == 0) {
-  WRITE_LOC(x2, 0, 0);
+  WRITE_LOC(i1, 0, 1);
  } else {
-  loc h = READ_LOC(x, 0);
-  loc n = READ_LOC(x, 1);
-  if (i1 <= h) {
-   filter(i1, n, x2);
-   loc x4 = READ_LOC(x2, 0);
-   loc x = (loc)malloc(2 * sizeof(loc));
-   WRITE_LOC(x2, 0, x);
-   WRITE_LOC(x, 1, x4);
-   WRITE_LOC(x, 0, h);
-  } else {
-   filter(i1, n, x2);
-  }
+  loc p = READ_LOC(x, 1);
+  loc q = READ_LOC(x, 2);
+  treeSize(p, i1);
+  loc i3 = READ_LOC(i1, 0);
+  treeSize(q, i1);
+  loc i = READ_LOC(i1, 0);
+  WRITE_LOC(i1, 0, (long)((long)1 + (long)i3) + (long)i);
  }
 }
 
 int main() {
   loc _out = malloc(sizeof(loc));
-  int _x0 = 0;
-  loc _x1 = 0;
-  _x0 = 7;
-  _x1 = _generateIntList();
-  filter(_x0, _x1, _out);
+  loc _x0 = 0;
+  _x0 = _generateBinaryTree();
+  treeSize(_x0, _out);
+  loc _derefOut = READ_LOC(_out, 0);
+  _printInt(_derefOut);
+
   return 0;
 }

### 2024-01-16 tests/c/common/generators.h

@@ -108,9 +108,6 @@ loc _generateIntList() {
 //   _generateIntListHelper(0, LIST_MAX, x);
 // }
 //
-void _generateBinaryTree(int** tree) {
-  // TODO
-}
 
 // Peano naturals //
 void printNat(loc x) {
@@ -146,6 +143,28 @@ void generateNat(int num, loc x) {
 
   loc r = READ_LOC(head, 0);
   WRITE_LOC(x, 0, r);
+}
+
+loc createBinaryTree(int depth) {
+    if (depth == 0) {
+        // Create a leaf node
+        loc leaf = (loc)malloc(3 * sizeof(union sslval)); // Allocate space for value and two children
+        WRITE_INT(leaf, 0, 0); // Set value to 0 or a sentinel value for leaves
+        WRITE_LOC(leaf, 1, NULL); // Left child is NULL
+        WRITE_LOC(leaf, 2, NULL); // Right child is NULL
+        return leaf;
+    } else {
+        // Create an internal node
+        loc node = (loc)malloc(3 * sizeof(union sslval)); // Allocate space for value and two children
+        WRITE_INT(node, 0, depth); // Set the value of the node
+        WRITE_LOC(node, 1, createBinaryTree(depth - 1)); // Create left subtree
+        WRITE_LOC(node, 2, createBinaryTree(depth - 1)); // Create right subtree
+        return node;
+    }
+}
+
+loc _generateBinaryTree() {
+  return createBinaryTree(10);
 }
 
 #endif

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -1,5 +1,5 @@
 theList :: [Int]
-theList = [1..100000]
+theList = [1..10000]
 
 filterLt :: Int -> [Int] -> [Int]
 filterLt n [] = []

### 2024-01-16 tests/haskell/LeftList.hs

@@ -1,0 +1,20 @@
+
+data Tree = Leaf | Node Int Tree Tree
+  deriving (Show)
+
+leftList :: Tree -> [Int]
+leftList Leaf = []
+leftList (Node x l r) = x : leftList l
+
+mkCompleteTree :: Int -> Tree
+mkCompleteTree 0 = Leaf
+mkCompleteTree depth =
+  let subtree = mkCompleteTree (depth - 1)
+  in
+  Node depth subtree subtree
+
+tree :: Tree
+tree = mkCompleteTree 10
+
+main :: IO ()
+main = print $ leftList tree

### 2024-01-16 tests/haskell/Take.hs

@@ -1,0 +1,15 @@
+
+data Nat = Z | S Nat
+  deriving (Show)
+
+myTake :: Nat -> [Int] -> [Int]
+myTake Z _ = []
+myTake (S n) (x:xs) = x : myTake n xs
+
+theList :: [Int]
+theList = [1..10000]
+
+
+main :: IO ()
+main = print $ myTake 7 theList
+

### 2024-01-16 tests/haskell/TreeSize.hs

@@ -1,0 +1,21 @@
+
+data Tree = Leaf | Node Int Tree Tree
+  deriving (Show)
+
+treeSize :: Tree -> Int
+treeSize Leaf = 1
+treeSize (Node _ l r) = 1 + treeSize l + treeSize r
+
+mkCompleteTree :: Int -> Tree
+mkCompleteTree 0 = Leaf
+mkCompleteTree depth =
+  let subtree = mkCompleteTree (depth - 1)
+  in
+  Node depth subtree subtree
+
+tree :: Tree
+tree = mkCompleteTree 10
+
+main :: IO ()
+main = print $ treeSize tree
+

### 2024-01-16 src/Bench.hs
fix some benchmarks; add benchmark sanity check (currently fails)

@@ -1,6 +1,5 @@
 module Main
   where
-
 import PikaC.Benchmark.Benchmark
 import Criterion
 import Criterion.Main
@@ -23,15 +22,28 @@ main = do
   synthedBenchmarks <- traverse synthBenchmark compiledBenchmarks0
   putStrLn "...benchmarks synthed."
 
+  putStrLn "comparing C benchmark outputs with Haskell benchmark outputs..."
+  runCBenchmarks SanityCheck CUnoptimized HaskellUnoptimized (catMaybes synthedBenchmarks)
+  putStrLn "...benchmark outputs match."
+
   putStrLn "generating C benchmarks..."
-  cbenchResults <- runCBenchmarks (catMaybes synthedBenchmarks)
+  cbenchResultsUnoptimized <- runCBenchmarks NoDiff CUnoptimized HaskellUnoptimized (catMaybes synthedBenchmarks)
+  cbenchResultsOptimized <- runCBenchmarks NoDiff CO3 HaskellO2 (catMaybes synthedBenchmarks)
   putStrLn "...C benchmarks generated."
 
-  putStrLn $ cBenchmarkToLaTeX $ cbenchResults
+  let printCBenchResults = do
+        putStrLn "-- Unoptimized --"
+        putStrLn $ cBenchmarkToLaTeX $ cbenchResultsUnoptimized
+        putStrLn ""
+        putStrLn "-- C with -O3 and Haskell with -O2 --"
+        putStrLn $ cBenchmarkToLaTeX $ cbenchResultsOptimized
+
+  printCBenchResults
 
 
   (compiledBenchmarks, results) <- runBenchmarks benchmarks
   putStrLn $ toLaTeX results
-  -- print results
-  -- defaultMain [runBenchmarks benchmarks]
+  putStrLn ""
+  putStrLn ""
+  printCBenchResults
 

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs

@@ -2,6 +2,7 @@
 {-# LANGUAGE RecordWildCards #-}
 {-# LANGUAGE LambdaCase #-}
 {-# LANGUAGE UndecidableInstances #-}
+{-# LANGUAGE TypeFamilies #-}
 
 module PikaC.Benchmark.Benchmark
   where
@@ -25,6 +26,11 @@ import PikaC.Utils
 import PikaC.Ppr
 
 import Data.Maybe (catMaybes)
+import qualified Data.Text as T
+import qualified Data.Text.Encoding as T
+import qualified Data.ByteString.Lazy as LBS
+
+import qualified System.Process.Typed as Typed
 
 import Control.Monad.Identity
 import Control.Monad
@@ -126,11 +132,11 @@ data BenchmarkResult =
   }
   deriving (Show)
 
-data CBenchmarkResult =
+data CBenchmarkResult a =
   CBenchmarkResult
   { cbenchResultName :: String
-  , cbenchResultCTime :: Report
-  , cbenchResultHaskellTime :: Report
+  , cbenchResultCTime :: a
+  , cbenchResultHaskellTime :: a
   }
 
 instance NFData CType
@@ -210,7 +216,7 @@ runBenchmarks benchmarks = do
       putStrLn $ "benchmark: " ++ prefix ++ "/" ++ benchName x
       benchmarkWith' config $ f x
 
-cBenchmarkToLaTeX :: [CBenchmarkResult] -> String
+cBenchmarkToLaTeX :: [CBenchmarkResult Report] -> String
 cBenchmarkToLaTeX results =
   unlines $
     [cmd "begin{tabular}{|c|c|c|}"
@@ -254,20 +260,37 @@ fromReport = printf "%.3f" . estPoint . anMean . reportAnalysis
 cmd :: String -> String
 cmd s = "\\" <> s
 
-runCBenchmarksCompiled :: [PikaSynthBenchmark] -> IO [CBenchmarkResult]
-runCBenchmarksCompiled = (runCBenchmarks . catMaybes) <=< traverse synthBenchmark
+runCBenchmarksCompiled :: Diff a -> COpts -> HaskellOpts -> [PikaSynthBenchmark] -> IO [CBenchmarkResult a]
+runCBenchmarksCompiled diff cOpts haskellOpts = (runCBenchmarks diff cOpts haskellOpts . catMaybes) <=< traverse synthBenchmark
 
[…]

### 2024-01-16 tests/c/common/generators.h

@@ -62,7 +62,7 @@ void _printIntList(loc x)
 //   return;
 // }
 
-#define LIST_MAX 10000
+#define LIST_MAX 100000
 
 loc _generateIntListHelper(int i, int len) {
   if (len == 0) return NULL;
@@ -122,27 +122,17 @@ void printNat(loc x) {
   printf("%d", i);
 }
 
-void generateNat(int num, loc x) {
-  loc head = malloc(sizeof(loc));
+loc _generateNatN(int n) {
+  if (n == 0) return NULL;
 
-  loc n = NULL;
-  for (int i = 0; i < num; ++i) {
-    if (i == 0) {
-      n = malloc(sizeof(loc));
-      WRITE_LOC(head, 0, n);
-    }
+  loc node = (loc)malloc(2 * sizeof(loc));
+  WRITE_LOC(node, 0, _generateNatN(n - 1));
+    
+  return node;
+}
 
-    if (i < num-1) {
-      loc next = malloc(sizeof(loc));
-      WRITE_LOC(n, 0, next);
-      n = next;
-    } else {
-      WRITE_LOC(n, 0, NULL);
-    }
-  }
-
-  loc r = READ_LOC(head, 0);
-  WRITE_LOC(x, 0, r);
+loc _generateNat() {
+  return _generateNatN(5000);
 }
 
 loc createBinaryTree(int depth) {
@@ -164,7 +154,7 @@ loc createBinaryTree(int depth) {
 }
 
 loc _generateBinaryTree() {
-  return createBinaryTree(10);
+  return createBinaryTree(15);
 }
 
 #endif

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -1,5 +1,5 @@
 theList :: [Int]
-theList = [1..10000]
+theList = [1..100000]
 
 filterLt :: Int -> [Int] -> [Int]
 filterLt n [] = []

### 2024-01-16 tests/haskell/Take.hs

@@ -7,9 +7,12 @@ myTake Z _ = []
 myTake (S n) (x:xs) = x : myTake n xs
 
 theList :: [Int]
-theList = [1..10000]
+theList = [1..100000]
 
+fromInt :: Int -> Nat
+fromInt 0 = Z
+fromInt n = S (fromInt (n - 1))
 
 main :: IO ()
-main = print $ myTake 7 theList
+main = print $ myTake (fromInt 5000) theList
 

### 2024-01-16 tests/haskell/TreeSize.hs

@@ -14,7 +14,7 @@ mkCompleteTree depth =
   Node depth subtree subtree
 
 tree :: Tree
-tree = mkCompleteTree 10
+tree = mkCompleteTree 15
 
 main :: IO ()
 main = print $ treeSize tree

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
Use one file for sizes of data structures for both C and Haskell benchmarks

@@ -69,7 +69,10 @@ cCompiler :: FilePath
 cCompiler = "gcc"
 
 includePath :: FilePath
-includePath = "tests/c"
+includePath = "tests" </> "c"
+
+haskellIncludePath :: FilePath
+haskellIncludePath = "tests" </> "haskell"
 
 haskellCompiler :: FilePath
 haskellCompiler = "ghc"
@@ -327,7 +330,7 @@ benchC name diff cOpts ghcOpts inputGenerators outputPrinter fn haskellCodeFile 
           -- putStrLn code
 
           systemQuiet $ cCompiler ++ " " ++ cOpts ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
-          systemQuiet $ haskellCompiler ++ " " ++ ghcOpts ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
+          systemQuiet $ haskellCompiler ++ " -XCPP " ++ ghcOpts ++ " -I" ++ haskellIncludePath ++ " " ++ (haskellIncludePath </> "Common.hs") ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
 
           (cReport, haskellReport) <- diffBenchmarkResults name diff (execTempName, []) (execHaskellTempName, [])
 
@@ -391,7 +394,7 @@ diffBenchmarkResults name diff cCmd haskellCmd = do
 
                 if (cOut /= haskellOut)
                   then do
-                    putStrLn $ "ERROR: Benchmark results differ between C and Haskell."
+                    putStrLn $ "ERROR: Benchmark results differ between C and Haskell for benchmark " ++ show name
                     exitFailure
                   else pure ((), ())))
 

### 2024-01-16 tests/c/common/generators.h

@@ -2,6 +2,7 @@
 #define GENERATORS_H
 
 #include "common.h"
+#include "../../haskell/Defaults.hs"
 
 loc _generateInt() {
   static int i = 0;
@@ -61,8 +62,6 @@ void _printIntList(loc x)
 //
 //   return;
 // }
-
-#define LIST_MAX 100000
 
 loc _generateIntListHelper(int i, int len) {
   if (len == 0) return NULL;
@@ -132,7 +131,7 @@ loc _generateNatN(int n) {
 }
 
 loc _generateNat() {
-  return _generateNatN(5000);
+  return _generateNatN(DEFAULT_NAT);
 }
 
 loc createBinaryTree(int depth) {
@@ -154,7 +153,7 @@ loc createBinaryTree(int depth) {
 }
 
 loc _generateBinaryTree() {
-  return createBinaryTree(15);
+  return createBinaryTree(BINARY_TREE_SIZE);
 }
 
 #endif

### 2024-01-16 tests/haskell/Common.hs

@@ -1,0 +1,29 @@
+module Common where
+
+#include "Defaults.hs"
+
+data Tree = Leaf | Node Int Tree Tree
+  deriving (Show)
+
+theList :: [Int]
+theList = [1 .. LIST_MAX]
+
+mkCompleteTree :: Int -> Tree
+mkCompleteTree 0 = Leaf
+mkCompleteTree depth =
+  let subtree = mkCompleteTree (depth - 1)
+  in
+  Node depth subtree subtree
+
+tree :: Tree
+tree = mkCompleteTree BINARY_TREE_SIZE
+
+data Nat = Z | S Nat
+  deriving (Show)
+
+fromInt :: Int -> Nat
+fromInt 0 = Z
+fromInt n = S (fromInt (n - 1))
+
+theNat :: Nat
+theNat = fromInt DEFAULT_NAT

### 2024-01-16 tests/haskell/Defaults.hs

@@ -1,0 +1,4 @@
+#define LIST_MAX 100000
+#define DEFAULT_NAT 5000
+#define BINARY_TREE_SIZE 15
+

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -1,5 +1,4 @@
-theList :: [Int]
-theList = [1..100000]
+import Common
 
 filterLt :: Int -> [Int] -> [Int]
 filterLt n [] = []

### 2024-01-16 tests/haskell/Take.hs

@@ -1,6 +1,4 @@
-
-data Nat = Z | S Nat
-  deriving (Show)
+import Common
 
 myTake :: Nat -> [Int] -> [Int]
 myTake Z _ = []
@@ -8,10 +6,6 @@ myTake (S n) (x:xs) = x : myTake n xs
 
 theList :: [Int]
 theList = [1..100000]
-
-fromInt :: Int -> Nat
-fromInt 0 = Z
-fromInt n = S (fromInt (n - 1))
 
 main :: IO ()
 main = print $ myTake (fromInt 5000) theList

### 2024-01-16 tests/haskell/TreeSize.hs

@@ -1,20 +1,8 @@
-
-data Tree = Leaf | Node Int Tree Tree
-  deriving (Show)
+import Common
 
 treeSize :: Tree -> Int
 treeSize Leaf = 1
 treeSize (Node _ l r) = 1 + treeSize l + treeSize r
-
-mkCompleteTree :: Int -> Tree
-mkCompleteTree 0 = Leaf
-mkCompleteTree depth =
-  let subtree = mkCompleteTree (depth - 1)
-  in
-  Node depth subtree subtree
-
-tree :: Tree
-tree = mkCompleteTree 15
 
 main :: IO ()
 main = print $ treeSize tree

### 2024-01-16 src/Bench.hs
Disable sanity check for now

@@ -8,24 +8,31 @@ import Control.DeepSeq
 
 import Data.Maybe (catMaybes)
 
+performSanityCheck :: Bool
+performSanityCheck = False
+
 main :: IO ()
 main = do
   benchmarks <- parseBenchmarks benchmarkConfigPath
 
-  putStrLn "compiling benchmarks..."
+  putStrLn "\ncompiling benchmarks..."
 
   let compiledBenchmarks0 = map compileBenchmark benchmarks
 
   compiledBenchmarks0 `deepseq` putStrLn "...benchmarks compiled."
+  putStrLn ""
 
   putStrLn "synthing benchmarks..."
   synthedBenchmarks <- traverse synthBenchmark compiledBenchmarks0
   putStrLn "...benchmarks synthed."
+  putStrLn ""
 
-  putStrLn "comparing C benchmark outputs with Haskell benchmark outputs..."
-  runCBenchmarks SanityCheck CUnoptimized HaskellUnoptimized (catMaybes synthedBenchmarks)
-  putStrLn "...benchmark outputs match."
+  when performSanityCheck $ do
+    putStrLn "comparing C benchmark outputs with Haskell benchmark outputs..."
+    runCBenchmarks SanityCheck CUnoptimized HaskellUnoptimized (catMaybes synthedBenchmarks)
+    putStrLn "...benchmark outputs match."
 
+  putStrLn ""
   putStrLn "generating C benchmarks..."
   cbenchResultsUnoptimized <- runCBenchmarks NoDiff CUnoptimized HaskellUnoptimized (catMaybes synthedBenchmarks)
   cbenchResultsOptimized <- runCBenchmarks NoDiff CO3 HaskellO2 (catMaybes synthedBenchmarks)
@@ -39,9 +46,11 @@ main = do
         putStrLn $ cBenchmarkToLaTeX $ cbenchResultsOptimized
 
   printCBenchResults
-
+  putStrLn ""
 
   (compiledBenchmarks, results) <- runBenchmarks benchmarks
+
+  putStrLn ""
   putStrLn $ toLaTeX results
   putStrLn ""
   putStrLn ""

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs

@@ -3,6 +3,7 @@
 {-# LANGUAGE LambdaCase #-}
 {-# LANGUAGE UndecidableInstances #-}
 {-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE ScopedTypeVariables #-}
 
 module PikaC.Benchmark.Benchmark
   where
@@ -30,6 +31,8 @@ import qualified Data.Text as T
 import qualified Data.Text.Encoding as T
 import qualified Data.ByteString.Lazy as LBS
 
+import Data.Bifunctor
+
 import qualified System.Process.Typed as Typed
 
 import Control.Monad.Identity
@@ -53,6 +56,9 @@ import Criterion
 import Criterion.Types
 import Criterion.Main.Options
 import Statistics.Types
+
+import Data.Algorithm.Diff
+import Data.Algorithm.DiffOutput
 
 import Control.DeepSeq
 
@@ -263,7 +269,7 @@ fromReport = printf "%.3f" . estPoint . anMean . reportAnalysis
 cmd :: String -> String
 cmd s = "\\" <> s
 
-runCBenchmarksCompiled :: Diff a -> COpts -> HaskellOpts -> [PikaSynthBenchmark] -> IO [CBenchmarkResult a]
+runCBenchmarksCompiled :: BenchCheck a -> COpts -> HaskellOpts -> [PikaSynthBenchmark] -> IO [CBenchmarkResult a]
 runCBenchmarksCompiled diff cOpts haskellOpts = (runCBenchmarks diff cOpts haskellOpts . catMaybes) <=< traverse synthBenchmark
 
 data HaskellOpts = HaskellUnoptimized | HaskellO2
@@ -280,7 +286,7 @@ instance Ppr COpts where
   ppr CUnoptimized = text "-O0"
   ppr CO3 = text "-O3"
 
-runCBenchmarks :: Diff a -> COpts -> HaskellOpts -> [PikaCBenchmark] -> IO [CBenchmarkResult a]
+runCBenchmarks :: BenchCheck a -> COpts -> HaskellOpts -> [PikaCBenchmark] -> IO [CBenchmarkResult a]
 runCBenchmarks diff cOpts haskellOpts =
   mapM $ \bench -> do
     benchC (benchName bench)
@@ -292,7 +298,7 @@ runCBenchmarks diff cOpts haskellOpts =
            (firstSynthed (fst (benchContents bench)))
[…]

### 2024-01-16 tests/haskell/Common.hs

@@ -6,7 +6,7 @@ data Tree = Leaf | Node Int Tree Tree
   deriving (Show)
 
 theList :: [Int]
-theList = [1 .. LIST_MAX]
+theList = [0 .. LIST_MAX-1]
 
 mkCompleteTree :: Int -> Tree
 mkCompleteTree 0 = Leaf

### 2024-01-16 tests/haskell/Defaults.hs

@@ -1,4 +1,4 @@
-#define LIST_MAX 100000
+#define LIST_MAX 100
 #define DEFAULT_NAT 5000
 #define BINARY_TREE_SIZE 15
 

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -3,7 +3,7 @@ import Common
 filterLt :: Int -> [Int] -> [Int]
 filterLt n [] = []
 filterLt n (x:xs)
-  | x < n = filterLt n xs
+  | x < n     = filterLt n xs
   | otherwise = x : filterLt n xs
 
 main :: IO ()

### 2024-01-16 src/Bench.hs
an import

@@ -5,6 +5,7 @@ import Criterion
 import Criterion.Main
 
 import Control.DeepSeq
+import Control.Monad
 
 import Data.Maybe (catMaybes)
 

### 2024-01-16 tests/haskell/Take.hs
a small fix

@@ -4,9 +4,6 @@ myTake :: Nat -> [Int] -> [Int]
 myTake Z _ = []
 myTake (S n) (x:xs) = x : myTake n xs
 
-theList :: [Int]
-theList = [1..100000]
-
 main :: IO ()
 main = print $ myTake (fromInt 5000) theList
 

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
silence some unnecessary output

@@ -371,7 +371,7 @@ diffBenchmarkResults name diff cCmd haskellCmd = do
   let doSystem :: BenchCheck a -> (String, [String]) -> IO ()
       doSystem = \case
           SanityCheck -> \(cmd, args) -> systemVeryQuiet $ unwords (cmd : args)
-          NoDiff -> \(cmd, args) -> systemQuiet $ unwords (cmd : args)
+          NoDiff -> \(cmd, args) -> systemVeryQuiet $ unwords (cmd : args)
 
 
   bracket (openTempFile "temp" "c-out.txt")

### 2024-01-16 tests/haskell/Take.hs
Fix Haskell 'Take' benchmark

@@ -3,6 +3,7 @@ import Common
 myTake :: Nat -> [Int] -> [Int]
 myTake Z _ = []
 myTake (S n) (x:xs) = x : myTake n xs
+mkTake (S _) xs = xs
 
 main :: IO ()
 main = print $ myTake (fromInt 5000) theList

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
force recompilation in Haskell benchmarks

@@ -342,7 +342,7 @@ benchC name diff cOpts ghcOpts inputGenerators outputPrinter fn haskellCodeFile 
 
           flip finally cleanUp $ do
             systemQuiet $ cCompiler ++ " " ++ cOpts ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
-            systemQuiet $ haskellCompiler ++ " -XCPP " ++ ghcOpts ++ " -I" ++ haskellIncludePath ++ " " ++ (haskellIncludePath </> "Common.hs") ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
+            systemQuiet $ haskellCompiler ++ " -fforce-recomp -XCPP " ++ ghcOpts ++ " -I" ++ haskellIncludePath ++ " " ++ (haskellIncludePath </> "Common.hs") ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
 
             (cReport, haskellReport) <- diffBenchmarkResults name diff (execTempName, []) (execHaskellTempName, [])
 

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
more work on benchmarks

@@ -313,10 +313,12 @@ benchC name diff cOpts ghcOpts inputGenerators outputPrinter fn haskellCodeFile 
           ,"int main() {"
           ]
           ++ decls
-          ++ zipWith applyInputGenerator inputGenerators params
+          ++ ["  for (int i = 0; i < TEST_ITERATIONS; ++i) {"]
+          ++ map ("  " ++) (zipWith applyInputGenerator inputGenerators params)
           ++ ["  " ++ C.cfunctionName fn ++ "(" ++ (intercalate ", " params) ++ ", " ++ out ++ ");"]
-          ++ [applyOutputPrinter outputPrinter out]
-          ++ ["printf(\"\\n\");"]
+          ++ ["  " ++ applyOutputPrinter outputPrinter out]
+          ++ ["    printf(\"\\n\");"]
+          ++ ["  }"]
           ++
           ["  return 0;"
           ,"}"

### 2024-01-16 tests/haskell/Common.hs

@@ -1,6 +1,14 @@
 module Common where
 
+import Control.Monad
+
 #include "Defaults.hs"
+
+testIterations :: Int
+testIterations = TEST_ITERATIONS
+
+testLoop :: IO () -> IO ()
+testLoop = replicateM_ testIterations
 
 data Tree = Leaf | Node Int Tree Tree
   deriving (Show)

### 2024-01-16 tests/haskell/Defaults.hs

@@ -1,4 +1,6 @@
-#define LIST_MAX 100
+#define LIST_MAX 10000
 #define DEFAULT_NAT 5000
 #define BINARY_TREE_SIZE 15
+
+#define TEST_ITERATIONS 10
 

### 2024-01-16 tests/haskell/FilterLt.hs

@@ -1,4 +1,5 @@
 import Common
+import Control.Monad
 
 filterLt :: Int -> [Int] -> [Int]
 filterLt n [] = []
@@ -7,5 +8,5 @@ filterLt n (x:xs)
   | otherwise = x : filterLt n xs
 
 main :: IO ()
-main = print $ filterLt 7 theList
+main = testLoop $ print $ filterLt 7 theList
 

### 2024-01-16 tests/haskell/LeftList.hs

@@ -1,20 +1,9 @@
 
-data Tree = Leaf | Node Int Tree Tree
-  deriving (Show)
+import Common
 
 leftList :: Tree -> [Int]
 leftList Leaf = []
 leftList (Node x l r) = x : leftList l
 
-mkCompleteTree :: Int -> Tree
-mkCompleteTree 0 = Leaf
-mkCompleteTree depth =
-  let subtree = mkCompleteTree (depth - 1)
-  in
-  Node depth subtree subtree
-
-tree :: Tree
-tree = mkCompleteTree 10
-
 main :: IO ()
-main = print $ leftList tree
+main = testLoop $ print $ leftList tree

### 2024-01-16 tests/haskell/Take.hs

@@ -3,8 +3,8 @@ import Common
 myTake :: Nat -> [Int] -> [Int]
 myTake Z _ = []
 myTake (S n) (x:xs) = x : myTake n xs
-mkTake (S _) xs = xs
+myTake (S _) xs = xs
 
 main :: IO ()
-main = print $ myTake (fromInt 5000) theList
+main = testLoop $ print $ myTake (fromInt 5000) theList
 

### 2024-01-16 tests/haskell/TreeSize.hs

@@ -5,5 +5,5 @@ treeSize Leaf = 1
 treeSize (Node _ l r) = 1 + treeSize l + treeSize r
 
 main :: IO ()
-main = print $ treeSize tree
+main = testLoop $ print $ treeSize tree
 

### 2024-01-16 src/PikaC/Benchmark/Benchmark.hs
update C benchmarks

@@ -230,7 +230,7 @@ cBenchmarkToLaTeX results =
   unlines $
     [cmd "begin{tabular}{|c|c|c|}"
     ,cmd "hline"
-    ,"Name & C Time (s) & GHC Time (s)\\\\"
+    ,"Name & C Time (s) & Haskell Time (s)\\\\"
     ,cmd "hline"
     ]
     ++ map toRow results ++
@@ -313,8 +313,8 @@ benchC name diff cOpts ghcOpts inputGenerators outputPrinter fn haskellCodeFile 
           ,"int main() {"
           ]
           ++ decls
+          ++ zipWith applyInputGenerator inputGenerators params
           ++ ["  for (int i = 0; i < TEST_ITERATIONS; ++i) {"]
-          ++ map ("  " ++) (zipWith applyInputGenerator inputGenerators params)
           ++ ["  " ++ C.cfunctionName fn ++ "(" ++ (intercalate ", " params) ++ ", " ++ out ++ ");"]
           ++ ["  " ++ applyOutputPrinter outputPrinter out]
           ++ ["    printf(\"\\n\");"]
@@ -343,8 +343,8 @@ benchC name diff cOpts ghcOpts inputGenerators outputPrinter fn haskellCodeFile 
           -- putStrLn code
 
           flip finally cleanUp $ do
-            systemQuiet $ cCompiler ++ " " ++ cOpts ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
-            systemQuiet $ haskellCompiler ++ " -fforce-recomp -XCPP " ++ ghcOpts ++ " -I" ++ haskellIncludePath ++ " " ++ (haskellIncludePath </> "Common.hs") ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
+            system $ cCompiler ++ " " ++ cOpts ++ " -I" ++ includePath ++ " " ++ cCodeTempName ++ " -o " ++ execTempName
+            system $ haskellCompiler ++ " -no-keep-o-files -no-keep-hi-files" ++ " -package deepseq" ++ " -fforce-recomp -XCPP " ++ ghcOpts ++ " -I" ++ haskellIncludePath ++ " " ++ (haskellIncludePath </> "Common.hs") ++ " " ++ haskellCodeFile ++ " -o " ++ execHaskellTempName
 
             (cReport, haskellReport) <- diffBenchmarkResults name diff (execTempName, []) (execHaskellTempName, [])
 

### 2024-01-16 tests/haskell/LeftList.hs

@@ -5,5 +5,11 @@ leftList :: Tree -> [Int]
 leftList Leaf = []
 leftList (Node x l r) = x : leftList l
 
+go :: Int -> IO ()
+go 0 = pure ()
+go n = do
+  print $ leftList tree
+  go (n - 1)
+
 main :: IO ()
-main = testLoop $ print $ leftList tree
+main = go testIterations
